---
title: "diff136_iNd35_ctrl_asah1e1_axonalproteome"
output: html_document
date: "`r Sys.Date()`"
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# load packackes 
```{r}
# --- Load Required Libraries ---
# --- Core Tidyverse ---
library(tidyverse)     # includes dplyr, ggplot2, tidyr, purrr, tibble, readr, stringr, etc.
library(dplyr)
library(tidyr)

# --- Plotting Enhancements ---
library(ggplot2)
library(ggdist)        # stat_halfeye
library(gghalves)      # geom_half_point
library(ggrepel)       # text labels (if used)
library(pheatmap)      # for heatmaps
library(RColorBrewer)  # custom color scales
library(viridis)       # color scales (optional)
library(NatParksPalettes)
library(scales) 
library(eulerr)

# --- Plot Layouts ---
library(gridExtra)     # arranging plots
library(cowplot)       # aligning and combining plots
library(patchwork)     # alternate plot composition syntax


# --- Data Reshaping and Manipulation ---
library(reshape2)      # melt, dcast (legacy but used in some places)
library(forcats)       # factor reordering
library(data.table)    # fast table handling (if used)

library(lintr)
library(devtools)

```


# output dirs overview:
```{r}
out_dir_diff136_dataframes <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/dataframes"
dir.create(out_dir_diff136_dataframes, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_heatmap  <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/heatmaps"
dir.create(out_dir_diff136_heatmap, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_corrplots <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_CrossCorr"
dir.create(out_dir_diff136_corrplots, recursive = TRUE, showWarnings = FALSE)


out_dir_diff136_organelle_dist <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution"
dir.create(out_dir_diff136_organelle_dist, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_organelle_dist_bar <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution/barplots_general"
dir.create(out_dir_diff136_organelle_dist_bar, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_organelle_dist_barnorm <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution/barplots_WC_norm"
dir.create(out_dir_diff136_organelle_dist_barnorm, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_organelle_dist_norm_ranked <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution/barplots_WC_norm_ranked"
dir.create(out_dir_diff136_organelle_dist_norm_ranked, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_organelle_dist_norm_composite <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution/barplots_WC_norm_ranked_composite"
dir.create(out_dir_diff136_organelle_dist_norm_composite, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_organelle_dist_norm_composite_euler <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution/barplots_WC_norm_ranked_composite/euer_plots"
dir.create(out_dir_diff136_organelle_dist_norm_composite_euler, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_organelle_dist_ror_norm <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution/RoR_norm"
dir.create(out_dir_diff136_organelle_dist_ror_norm, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_rainbow <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution_Rainbow"
dir.create(out_dir_diff136_rainbow, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_rainbow_3anno <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution_Rainbow/3anno"
dir.create(out_dir_diff136_rainbow_3anno, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_rainbow_RoR <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/Organelle_Distribution_Rainbow/ror"
dir.create(out_dir_diff136_rainbow_RoR, recursive = TRUE, showWarnings = FALSE)



out_dir_diff136_GO <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/GO"
dir.create(out_dir_diff136_GO, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_GOcluster <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/GO/heatmapclusters"
dir.create(out_dir_diff136_GOcluster, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_GOror <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/GO/RoR"
dir.create(out_dir_diff136_GOror, recursive = TRUE, showWarnings = FALSE)

out_dir_diff136_GOranked <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/GO/ranking_sensitivity"
dir.create(out_dir_diff136_GOranked, recursive = TRUE, showWarnings = FALSE)

```



# I. Background diff 136 axonal proteome Ctrl / ASAH1 e1 day 35 iN differentiation 
Dataset contains in vitro differentiaed iNeurons at day 35 of differentiation
Genotypes: H9 NGN2 TMEM192-3xHA (het) Control and ASAH1-/- clone E1
Samples (all in triplicates): whole-cell (WC), soma fraction and axonal fraction 
18plex_TMTpro & run on Thermo Orbitrap Ascent w/FAIMS-Pro, hrMS2
data was created with MSstats, has p/q.values and log2FC for samples 
k-means clustering was performed on data, is in column "cluster"


# II. read in data & add protein organelle annotations
- facet wrap by annotations & save 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# 1.  read the data  
# ─────────────────────────────────────────────────────────────────────────────

diff136_df <- read.csv('/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/datasets/diff136_iNd35_Ctrl-ASAH1e1_axonalproteome_All_results_stats_localizationAdded.csv', header=T)


# ─────────────────────────────────────────────────────────────────────────────
# 2.  Transform p/q-values to -log10, and filter non-finite log2FCs
# ─────────────────────────────────────────────────────────────────────────────

log2fc_cols <- c(
  "log2FC_Ctrl_Soma.Ctrl",
  "log2FC_Ctrl_Axon.Ctrl",
  "log2FC_ASAH1.Ctrl",
  "log2FC_ASAH1_Soma.Ctrl",
  "log2FC_ASAH1_Axon.Ctrl",
  "log2FC_Ctrl_Axon.Ctrl_Soma",
  "log2FC_ASAH1.Ctrl_Soma",
  "log2FC_ASAH1_Soma.Ctrl_Soma",
  "log2FC_ASAH1_Axon.Ctrl_Soma",
  "log2FC_ASAH1.Ctrl_Axon",
  "log2FC_ASAH1_Soma.Ctrl_Axon",
  "log2FC_ASAH1_Axon.Ctrl_Axon",
  "log2FC_ASAH1_Soma.ASAH1",
  "log2FC_ASAH1_Axon.ASAH1",
  "log2FC_ASAH1_Axon.ASAH1_Soma"
)

diff136_df <- diff136_df %>%
  mutate(across(contains("p.val"), ~ ifelse(. > 0, -log10(.), NA), .names = "neg_log10_{.col}"),
         across(contains("q.val"), ~ ifelse(. > 0, -log10(.), NA), .names = "neg_log10_{.col}")) %>%
  filter(if_all(all_of(log2fc_cols), ~ is.finite(.)))

# ─────────────────────────────────────────────────────────────────────────────
# 3.  Add Subcellular Annotation
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1: Define contaminant list (optional, can be skipped if not relevant)
contaminant_genes <- c(
  "KRT8", "KRT19", "KRT9", "KRT10", "KRT2", "KRT18", "KRT1", "KRT5",
  "COL10A1", "COL13A1", "COL14A1", "COL18A1", "COL1A1", "COL1A2",
  "COL26A1", "COL2A1", "COL4A1", "COL4A2", "COL5A1", "COL6A1", "COL6A2"
)

# --- Step 2: Load and process subcellular annotation
cols_to_read <- c(1,2,3,4,5,6,7,9,10,12,15,16,17,18,19,20,21,22,23,25,26,27,28,29,30,31,32,33)
subcell_df <- read_csv("/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/neuroLSDgit/proteome/SubCellAnnotation.csv", show_col_types = FALSE)[, cols_to_read]
subcell_df[] <- lapply(subcell_df, as.character)

long_subcell_df <- subcell_df %>%
  pivot_longer(cols = everything(), names_to = "Localization", values_to = "Genes") %>%
  filter(!is.na(Genes) & Genes != "") %>%
  mutate(Genes = trimws(Genes)) %>%
  distinct()

binary_matrix <- long_subcell_df %>%
  mutate(value = TRUE) %>%
  pivot_wider(
    names_from = Localization,
    values_from = value,
    values_fill = list(value = FALSE)
  )

# --- Step 3: Merge with diff136_df by Gene.Symbol
diff136_df_annotated <- diff136_df %>%
  filter(!Gene.Symbol %in% contaminant_genes) %>%
  left_join(binary_matrix, by = c("Gene.Symbol" = "Genes"))

diff136_df_annotated <- diff136_df_annotated %>%
  mutate(across(where(is.logical) | where(is.numeric), ~ replace_na(., 0)))

# --- Step 4. save as csv file 
write.csv(diff136_df_annotated, 
          file = file.path(out_dir_diff136_dataframes,"diff136_iNd35_Ctrl-ASAH1e1_axonalproteome_annotated.csv"), row.names = FALSE)
```


### IIa. Heatmap (z-norm rows, cluster by cluster)
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Heatmap (z-norm rows, cluster by cluster)
# ─────────────────────────────────────────────────────────────────────────────
out_dir_diff136_heatmap  <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20250603_diff136_iN_d35_Ctrl-ASAH1E1_AxonalProteomics/diff136renv/heatmaps"
pdf_file <- file.path(out_dir_diff136_heatmap, "diff136_heatmap_axonalProt_ctrl_asah1.pdf")

# --- 1. set annotations and sample names 
expr_cols <- c(
  "Ctrl_1","Ctrl_2","Ctrl_3",
  "Ctrl_Soma_1","Ctrl_Soma_2","Ctrl_Soma_3",
  "Ctrl_Axon_1","Ctrl_Axon_2","Ctrl_Axon_3",
  "ASAH1_1","ASAH1_2","ASAH1_3",
  "ASAH1_Soma_1","ASAH1_Soma_2","ASAH1_Soma_3",
  "ASAH1_Axon_1","ASAH1_Axon_2","ASAH1_Axon_3"
)

# Matrix from expression columns
mat <- diff136_df %>%
  filter(!is.na(Gene.Symbol)) %>%
  mutate(Gene.Symbol = make.unique(as.character(Gene.Symbol))) %>%
  select(Gene.Symbol, cluster, all_of(expr_cols))

# Store and remove cluster for sorting
row_cluster <- mat$cluster
rownames(mat) <- mat$Gene.Symbol
mat <- mat[, expr_cols]

# Reorder rows by cluster
mat <- mat[order(row_cluster), , drop = FALSE]


# --- 2. set cluster colors and cluster annotations as rows
# Define full cluster levels explicitly
cluster_levels <- 0:6

# Force 'cluster' to be a factor with all levels, even if not present
row_annotation <- data.frame(
  cluster = factor(row_cluster[order(row_cluster)], levels = cluster_levels)
)
rownames(row_annotation) <- rownames(mat)

# Assign distinct colors (Set2 has up to 8 distinct hues)
cluster_palette <- setNames(
  RColorBrewer::brewer.pal(7, "Set2")[1:7],
  as.character(cluster_levels)
)

# Add to annotation color list
ann_colors <- list(
  Genotype = c(Ctrl = "grey40", ASAH1 = "dodgerblue3"),
  Fraction = c(`Whole-cell` = "#f1eef6",
               Soma         = "#bdc9e1",
               Axon         = "#74a9cf")
)

ann_colors$cluster <- cluster_palette


# --- 3. column annotations
annotation_col <- tibble(Sample = expr_cols) %>%
  mutate(
    Genotype = if_else(str_detect(Sample, "^Ctrl"), "Ctrl", "ASAH1"),
    Fraction = case_when(
      str_detect(Sample, "_Soma")  ~ "Soma",
      str_detect(Sample, "_Axon")  ~ "Axon",
      TRUE                         ~ "Whole-cell"
    )
  ) %>%
  column_to_rownames("Sample")



# --- 4. plot heatmap, save as pdf and store underlying dataframe as csv 
cairo_pdf(pdf_file, width = 9, height = 10)

pheatmap(
  mat,
  scale              = "row",
  color              = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col     = annotation_col,
  annotation_row     = row_annotation,
  annotation_colors  = ann_colors,
  breaks             = seq(-4, 4, length.out = 101),
  cluster_rows       = FALSE,
  cluster_cols       = TRUE,
  show_rownames      = FALSE,
  show_colnames      = TRUE,
  angle_col          = 90,
  main               = "diff136 day 35 axonal proteome: Ctrl vs ASAH1",
  fontsize           = 6,
  border_color       = NA
)

dev.off()

# Add cluster as first column to expression matrix
mat_with_cluster <- cbind(cluster = row_annotation$cluster, mat)

# Write to CSV with cluster
write.csv(mat_with_cluster,
          file = file.path(out_dir_diff136_heatmap,"diff136_iNd35_heatmap.csv"),
          row.names = TRUE)
```



### IIb.  Violin Plot of Cluster Abundance
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Violin Plot of Cluster Abundance
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. define & select sample labels for plotting  
sample_cols <- c(
  "log2FC_Ctrl_Soma.Ctrl",
  "log2FC_Ctrl_Axon.Ctrl",
  "log2FC_ASAH1_Soma.ASAH1",
  "log2FC_ASAH1_Axon.ASAH1"
)

# --- 2. Drop cluster 0 and NA
violin_df <- diff136_df_annotated %>%
  filter(!is.na(cluster) & cluster != 0) %>%
  pivot_longer(cols = all_of(sample_cols), names_to = "Sample", values_to = "log2FC") %>%
  filter(!is.na(log2FC))

# --- 3. Consistent y-axis scale
y_limits <- range(violin_df$log2FC, na.rm = TRUE)

# --- 4. plot facet wrap
diff136_clusterPlot <- ggplot(violin_df, aes(x = Sample, y = log2FC, fill = Sample)) +
  geom_jitter(
    aes(color = "red"), size = 0.5, alpha = 0.2, width = 0.2, height = 0
  ) +
  geom_violin(scale = "width", trim = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  #facet_wrap(~ cluster, scales = "fixed") +
  facet_wrap(~ cluster, scales = "free_y") +
  #coord_cartesian(ylim = y_limits) +
  scale_fill_manual(values = c(
    "log2FC_Ctrl_Soma.Ctrl" = "#f1eef6",
    "log2FC_Ctrl_Axon.Ctrl" = "#bdc9e1",
    "log2FC_ASAH1_Soma.ASAH1" = "#74a9cf",
    "log2FC_ASAH1_Axon.ASAH1" = "dodgerblue3"
  )) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none", 
        #panel.grid = element_blank()
        ) +
  labs(title = "log2FC by Sample per Cluster", x = NULL, y = "log2FC")

diff136_clusterPlot

# --- 5. save plot as PDF
ggsave(file.path(out_dir_diff136_heatmap, 
  "diff136_clusterPlot.pdf"), 
  diff136_clusterPlot, width = 6, height = 6, device = cairo_pdf)
```


## IIc. GO analysis of clusters of interest 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# GO term analysis based on heatmap clusters
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1. Define function for GO Term analysis 
plot_topGO_for_geneset <- function(gene_set, background_genes, tag = "clusterX", out_dir) {
  library(topGO)
  library(org.Hs.eg.db)
  library(tidyverse)

  # Step 1.1: Create geneList for topGO
  geneList <- factor(as.integer(background_genes %in% gene_set))
  names(geneList) <- background_genes

  # Step 1.2: Run topGO Fisher test (BP)
  GOdata <- new("topGOdata",
                ontology = "BP",
                allGenes = geneList,
                geneSel = function(x) x == 1,
                annot = annFUN.org,
                mapping = "org.Hs.eg.db",
                ID = "symbol")

  resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

  # Step 1.3: Generate GO table and extract raw p-values
  # calculate 30 top hits, but only plot top 10 
  go_table <- GenTable(GOdata, classicFisher = resultFisher, topNodes = 30)

  go_table <- go_table %>%
    mutate(
      raw_pval     = score(resultFisher)[GO.ID],
      Significant  = as.numeric(Significant),
      Expected     = as.numeric(Expected),
      Enrichment   = Significant / Expected,
      log10P       = -log10(raw_pval)
    ) %>%
    filter(!is.na(raw_pval)) %>%
    arrange(raw_pval) %>%
    slice_head(n = 10)

  # Step 1.4: Get associated genes per GO term
  go_genes <- genesInTerm(GOdata)
  go_table$Genes <- sapply(go_table$GO.ID, function(go_id) {
    intersect(go_genes[[go_id]], gene_set) %>% paste(collapse = ";")
  })

  # Step 1.5: Save GO term table + gene associations
  write_csv(go_table, file = file.path(out_dir, paste0("GO_results_", tag, ".csv")))

  # Step 1.6: Plot dot plot
  p <- ggplot(go_table, aes(x = Enrichment, y = reorder(Term, Enrichment))) +
    geom_point(aes(size = log10P, fill = Enrichment), shape = 21, color = "black", stroke = 0.25) +
    scale_fill_gradient(low = "white", high = "firebrick") +
    theme_bw(base_size = 8) +
    labs(
      x = "Fold Enrichment",
      y = "GO Term",
      size = expression(-log[10]*"P"),
      title = paste0("Top GO Terms (", tag, ")")
    ) +
    theme(panel.grid = element_blank())

  ggsave(file.path(out_dir, paste0("GO_dotplot_", tag, ".pdf")), p, width = 5, height = 3.5, device = cairo_pdf)

  return(p)
}



# --- Step 2. Define input clusters to analyse & background genes 
# Input: cluster genes
genes_interest <- diff136_df_annotated %>%
  #filter(cluster %in% c(4)) %>%
  filter(cluster %in% c(4,5)) %>%
  pull(Gene.Symbol) %>%
  unique()

# Background genes
background_genes <- unique(diff136_df_annotated$Gene.Symbol)


# --- Step 3. Run function 
# Run function
plot_topGO_for_geneset(
  gene_set = genes_interest,
  background_genes = background_genes,
  tag = "ASAH1_cluster4_5",
  out_dir = out_dir_diff136_GOcluster
)

```





# III. Correlation of Selected Annotations
calculate annotated-based correlation across all and also facet wrap them
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Correlation of Selected Annotations
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Copy out log2FC & annotated columns for wrangling
# Define log2FC columns (all that start with log2FC_)
log2fc_cols <- grep("^log2FC_", colnames(diff136_df_annotated), value = TRUE)

# Define annotation columns (binary organelle terms)
# assumes binary 0/1 format and not part of log2FC set
annotation_cols <- setdiff(colnames(diff136_df_annotated), log2fc_cols)
annotation_cols <- annotation_cols[sapply(diff136_df_annotated[, annotation_cols], function(x) all(x %in% c(0, 1, NA)))]

# Drop unwanted ones
annotation_cols <- setdiff(annotation_cols, c("Autophagy_det", "EndoDetailLoac"))



length(log2fc_cols)  # sanity check
length(annotation_cols)



# --- 2. Compute Spearman correlation between log2FC and annotations
org_corr_df <- map_dfr(annotation_cols, function(ann) {
  map_dfr(log2fc_cols, function(fc_col) {
    x <- diff136_df_annotated[[fc_col]]
    y <- as.numeric(diff136_df_annotated[[ann]])
    keep <- is.finite(x) & is.finite(y)
    if (sum(keep) < 5) return(NULL)

    cor_test <- suppressWarnings(cor.test(x[keep], y[keep], method = "spearman"))
    tibble(
      log2FC_Column = fc_col,
      Organelle     = ann,
      r             = unname(cor_test$estimate),
      p             = cor_test$p.value
    )
  })
})



# --- 3. Summary heatmap (log2FC × Organelle)
org_corr_mat <- org_corr_df %>%
  dplyr::select(log2FC_Column, Organelle, r) %>%
  tidyr::pivot_wider(names_from = Organelle, values_from = r) %>%
  tibble::column_to_rownames("log2FC_Column") %>%
  as.matrix()

# Extract rownames from org_corr_mat (log2FC columns)
annotation_row <- tibble(Sample = rownames(org_corr_mat)) %>%
  mutate(
    Genotype = if_else(str_detect(Sample, "^log2FC_Ctrl"), "Ctrl", "ASAH1"),
    Fraction = case_when(
      str_detect(Sample, "_Soma")  ~ "Soma",
      str_detect(Sample, "_Axon")  ~ "Axon",
      TRUE                         ~ "Whole-cell"
    )
  ) %>%
  column_to_rownames("Sample")


# Replace NA with 0 (or leave NA if you prefer gaps)
org_corr_mat[is.na(org_corr_mat)] <- 0

axonprotCorr <- pheatmap(org_corr_mat,
         color = rev(RColorBrewer::brewer.pal(11, "RdYlBu")),
         annotation_row = annotation_row,                
         annotation_colors = ann_colors,                
         na_col = "grey90",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize = 8,
         cellheight = 10,
         cellwidth = 10,
         border_color = NA, 
         main = "log2FC × Organelle Correlation")

axonprotCorr



# --- 4. Save heatmap as PDF and csv file 
pdf(file.path(out_dir_diff136_corrplots, "axonprotCorr.pdf"),
    width = 12, height = 16, useDingbats = FALSE)
print(axonprotCorr)
dev.off()

write.csv(org_corr_mat,
          file = file.path(out_dir_diff136_corrplots, "axonprotCorr.csv"),
          row.names = TRUE)



# --- 5. Faceted correlation plots (per-organelle, KO×KO)
# Function to plot KO×KO correlation for a given annotation
plot_ko_corr_matrix <- function(organelle_name, df = diff136_df_annotated) {
  gene_filter <- df[[organelle_name]] == 1
  mat <- df[gene_filter, log2fc_cols, drop = FALSE]

  if (nrow(mat) < 5) return(NULL)

  corr_mat <- cor(mat, use = "pairwise.complete.obs", method = "spearman")

  pheatmap(
    corr_mat,
    main = organelle_name,
    color = rev(brewer.pal(11, "RdYlBu")),
    silent = TRUE,
    cellheight = 15,
    cellwidth  = 15,
    na_col = "grey90",
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    border_color = NA,
    fontsize = 6
  )
}



# --- 6. Apply to all annotations
heatmap_list <- lapply(annotation_cols, function(org) {
  if (!org %in% colnames(diff136_df_annotated)) return(NULL)
  plot_ko_corr_matrix(org)
})
heatmap_list <- Filter(Negate(is.null), heatmap_list)



# --- 7. Save all in a faceted PDF
pdf(file.path(out_dir_diff136_corrplots, "diff136_log2FC_annotation_corr_facet.pdf"),
    width = 12, height = 16, useDingbats = FALSE)
print(axonprotCorr)
dev.off()



# --- 8. make subset of comparisons, plot heatmap and save pdf 
selected_samples <- c(
  "log2FC_ASAH1_Axon.ASAH1_Soma",
  "log2FC_ASAH1_Axon.ASAH1",
  "log2FC_ASAH1_Soma.ASAH1",
  "log2FC_ASAH1_Soma.Ctrl_Soma",
  "log2FC_ASAH1_Axon.Ctrl_Axon",
  "log2FC_ASAH1.Ctrl",
  "log2FC_Ctrl_Axon.Ctrl_Soma",
  "log2FC_Ctrl_Axon.Ctrl",
  "log2FC_Ctrl_Soma.Ctrl"
)

# Subset matrix and annotations
org_corr_subset <- org_corr_mat[rownames(org_corr_mat) %in% selected_samples, , drop = FALSE]
annotation_row_subset <- annotation_row[rownames(org_corr_subset), , drop = FALSE]

# Create heatmap
axonprotCorr_subset <- pheatmap(org_corr_subset,
  color = rev(RColorBrewer::brewer.pal(11, "RdYlBu")),
  annotation_row = annotation_row_subset,
  annotation_colors = ann_colors,
  na_col = "grey90",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize = 8,
  cellheight = 10,
  cellwidth = 10,
  border_color = NA,
  main = "log2FC × Organelle Correlation (Selected Samples)"
)

# Save to PDF
pdf(file.path(out_dir_diff136_corrplots, "axonprotCorr_subset.pdf"),
    width = 10, height = 12, useDingbats = FALSE)
print(axonprotCorr_subset)
dev.off()



# --- 9. Apply to all annotations
plot_ko_corr_matrix_select <- function(organelle_name, df = diff136_df_annotated, log2fc_subset = selected_samples) {
  gene_filter <- df[[organelle_name]] == 1
  mat <- df[gene_filter, log2fc_subset, drop = FALSE]

  if (nrow(mat) < 5) return(NULL)

  corr_mat <- cor(mat, use = "pairwise.complete.obs", method = "spearman")

  pheatmap(
    corr_mat,
    main = organelle_name,
    color = rev(brewer.pal(11, "RdYlBu")),
    silent = TRUE,
    cellheight = 15,
    cellwidth  = 15,
    na_col = "grey90",
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    border_color = NA,
    fontsize = 6
  )
}

# Build corr matrices for all annotations and save for later faceting 
heatmap_list_selected <- lapply(annotation_cols, function(org) {
  if (!org %in% colnames(diff136_df_annotated)) return(NULL)
  plot_ko_corr_matrix_select(org, log2fc_subset = selected_samples)
})
heatmap_list_selected <- Filter(Negate(is.null), heatmap_list_selected)

# Save to PDF
pdf(file.path(out_dir_diff136_corrplots, "diff136_selected_log2FC_annotation_corr_facet.pdf"), width = 35, height = 35, useDingbats = FALSE)
grid.arrange(grobs = lapply(heatmap_list_selected, `[[`, "gtable"), ncol = 4)
dev.off()


```


# IV. Fraction comparisons (WC/SOMA/PROJECTION)
## IVa. Barplots: General comparison annotation [mean(log2FC)]
### IVaa. Comparison across ALL sample permutations
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# barplots of mean(log2FC) of each organelle annotations across all sample comparisons
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1. Define relevant log2FC and annotation columns
log2fc_cols <- grep("^log2FC_", colnames(diff136_df_annotated), value = TRUE)

# Filter binary annotation columns (0/1 or NA only)
annotation_cols <- setdiff(colnames(diff136_df_annotated), log2fc_cols)
annotation_cols <- annotation_cols[
  sapply(diff136_df_annotated[, annotation_cols], function(x) all(x %in% c(0, 1, NA)))
]

# Drop unwanted annotations
annotation_cols <- setdiff(annotation_cols, c("Autophagy_det", "EndoDetailLoac"))


# --- Step 2. Reshape to long format for annotation × log2FC mapping
long_df <- diff136_df_annotated %>%
  dplyr::select(all_of(c("Gene.Symbol", log2fc_cols, annotation_cols))) %>%
  pivot_longer(cols = all_of(log2fc_cols), names_to = "Log2FC", values_to = "value") %>%
  pivot_longer(cols = all_of(annotation_cols), names_to = "Annotation", values_to = "is_annotated") %>%
  filter(is_annotated == 1 & is.finite(value))


# --- Step 3. Compute mean log2FC per organelle annotation
mean_log2fc_df <- long_df %>%
  group_by(Log2FC, Annotation) %>%
  summarise(mean_log2FC = mean(value, na.rm = TRUE), .groups = "drop")


# --- Step 4. Shared x-axis range calculation
xlims <- range(mean_log2fc_df$mean_log2FC, na.rm = TRUE)

# --- 4a. Facet plot across all log2FCs with fixed x-axis
loc_dist_overview_all <- ggplot(mean_log2fc_df, aes(x = reorder(Annotation, mean_log2FC), y = mean_log2FC)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip(ylim = xlims) +
  facet_wrap(~ Log2FC, ncol = 3) +
  labs(x = NULL, y = "Mean log2FC", title = "Mean log2FC per Organelle Annotation (All)") +
  theme_bw(base_size = 8)

ggsave(file.path(out_dir_diff136_organelle_dist_bar, "diff136_loc_dist_overview_all.pdf"),
  loc_dist_overview_all,
  width = 14, height = 12, device = cairo_pdf
)

# --- 4b. Filter for user-specified subset
subset_log2fc <- c(
  "log2FC_ASAH1_Axon.ASAH1_Soma",
  "log2FC_ASAH1_Axon.ASAH1",
  "log2FC_ASAH1_Soma.ASAH1",
  "log2FC_ASAH1_Soma.Ctrl_Soma",
  "log2FC_ASAH1_Axon.Ctrl_Axon",
  "log2FC_ASAH1.Ctrl",
  "log2FC_Ctrl_Axon.Ctrl_Soma",
  "log2FC_Ctrl_Axon.Ctrl",
  "log2FC_Ctrl_Soma.Ctrl"
)

mean_log2fc_subset_df <- mean_log2fc_df %>% filter(Log2FC %in% subset_log2fc)

loc_dist_overview_subset <- ggplot(mean_log2fc_subset_df, aes(x = reorder(Annotation, mean_log2FC), y = mean_log2FC)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip(ylim = xlims) +
  facet_wrap(~ Log2FC, ncol = 3) +
  labs(x = NULL, y = "Mean log2FC", title = "Mean log2FC per Organelle Annotation (Selected)") +
  theme_bw(base_size = 8)


# --- Step 5. Save to PDF
ggsave(file.path(out_dir_diff136_organelle_dist_bar, "diff136_loc_dist_overview_selected.pdf"),
  loc_dist_overview_subset,
  width = 14, height = 12, device = cairo_pdf
)
```

### IVab. Comparison across selected annotations
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# barplots of mean(log2FC) of each organelle annotations across all sample comparisons
# selected annotations  
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 6. Manual Subset of comparisons for function 
subset2_log2fc <- c(
  "log2FC_ASAH1_Soma.Ctrl_Soma",
  "log2FC_ASAH1_Axon.Ctrl_Axon"
)

mean2_log2fc_subset_df <- mean_log2fc_df %>% 
  filter(Annotation != "Nucleus") %>% 
  filter(Log2FC %in% subset2_log2fc)

loc_dist_overview_subset2 <- ggplot(mean2_log2fc_subset_df, aes(x = reorder(Annotation, mean_log2FC), y = mean_log2FC)) +
  geom_col(fill = "dodgerblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  #coord_flip(ylim = xlims) +
  coord_flip() +
  facet_wrap(~ Log2FC, ncol = 3) +
  labs(x = NULL, y = "Mean log2FC", title = "Mean log2FC per Organelle Annotation (Selected)") +
  theme_bw(base_size = 6) +
  theme(
   panel.grid = element_blank() 
  )


# --- Step 7. Save to PDF
ggsave(file.path(out_dir_diff136_organelle_dist_bar, "diff136_loc_dist_overview_selected_directComp.pdf"),
  loc_dist_overview_subset2,
  width = 3, height = 4, device = cairo_pdf
)

```

### IVac. Selected annotations & sorted 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# barplots of mean(log2FC) of each organelle annotations across all sample comparisons
# selected annotations  
# sorted in decreasing manner 
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 8a. Define order based on Axon comparison
axon_order <- mean2_log2fc_subset_df %>%
  filter(Log2FC == "log2FC_ASAH1_Axon.Ctrl_Axon") %>%
  arrange(mean_log2FC) %>%
  pull(Annotation)

# --- Step 8b. Apply this order as factor levels for all comparisons
mean2_log2fc_subset_df <- mean2_log2fc_subset_df %>%
  filter(Annotation != "Nucleus") %>% 
  mutate(Annotation = factor(Annotation, levels = axon_order))

# --- Step 8c. Plot with consistent order
loc_dist_overview_subset3 <- ggplot(
  mean2_log2fc_subset_df,
  aes(x = Annotation, y = mean_log2FC)
) +
  geom_col(fill = "dodgerblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  facet_wrap(~ Log2FC, ncol = 2) +
  labs(x = NULL, y = "Mean log2FC",
       title = "Mean log2FC per Organelle Annotation\n(Prioritized by Axon Order)") +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank())


# --- Step 9. Save to PDF
ggsave(file.path(out_dir_diff136_organelle_dist_bar, "diff136_loc_dist_overview_selected_directComp_decs.pdf"),
  loc_dist_overview_subset3,
  width = 3, height = 4, device = cairo_pdf
)

```


## IVb. Barplots: General comparison annotation WC-norm mean(log2FC)
### IVba. Barplots: ASAH1 [axon/WC] vs Ctrl [axon/WC]
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# barplots of mean(log2FC) of each organelle annotations across all sample comparisons
# internally normalized comparisons
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1. Manual Subset of internal norm comparisons
subset3_log2fc <- c(
  "log2FC_ASAH1_Axon.ASAH1",
  "log2FC_Ctrl_Axon.Ctrl"
)

mean3_log2fc_subset_df <- mean_log2fc_df %>% 
  filter(Annotation != "Nucleus") %>% 
  filter(Log2FC %in% subset3_log2fc)

# --- Step 2a. Define order based on Axon.ASAH1 comparison
axon_norm_order2 <- mean3_log2fc_subset_df %>%
  filter(Log2FC == "log2FC_ASAH1_Axon.ASAH1") %>%
  arrange(mean_log2FC) %>%
  pull(Annotation)

# --- Step 2b. Apply this order as factor levels for both comparisons
mean3_log2fc_subset_df <- mean3_log2fc_subset_df %>%
  mutate(Annotation = factor(Annotation, levels = axon_norm_order2))

# --- Step 3. Plot with consistent order 
loc_dist_overview_subset_norm <- ggplot(
  mean3_log2fc_subset_df,
  aes(x = Annotation, y = mean_log2FC)
) +
  geom_col(fill = "dodgerblue2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  facet_wrap(~ Log2FC, ncol = 2) +
  labs(x = NULL, y = "Mean log2FC",
       title = "Mean log2FC per Organelle Annotation\n(Internal Normalization)") +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank())

# --- Step 4. Save to PDF
ggsave(file.path(out_dir_diff136_organelle_dist_barnorm, "diff136_loc_dist_overview_selected_directComp_internalNorm.pdf"), loc_dist_overview_subset_norm, width = 3, height = 4, device = cairo_pdf)



# ─────────────────────────────────────────────────────────────────────────────
# barplots with internal norm comparisons + ratio-of-ratios
# ─────────────────────────────────────────────────────────────────────────────
# --- Step 5. Compute ratio-of-ratios
ratio_df <- mean3_log2fc_subset_df %>%
  dplyr::select(Log2FC, Annotation, mean_log2FC) %>%
  pivot_wider(names_from = Log2FC, values_from = mean_log2FC) %>%
  mutate(ratio_of_ratios = `log2FC_ASAH1_Axon.ASAH1` / `log2FC_Ctrl_Axon.Ctrl`,
         Log2FC = "RatioOfRatios") %>%
  dplyr::select(Annotation, Log2FC, mean_log2FC = ratio_of_ratios)

# --- Step 6. Combine with original df
combined_df <- bind_rows(mean3_log2fc_subset_df, ratio_df) %>%
  mutate(Annotation = factor(Annotation, levels = axon_norm_order2))

# --- Step 7. Plot with 3 facets
loc_dist_overview_subset_ratio <- ggplot(
  combined_df,
  aes(x = Annotation, y = mean_log2FC)
) +
  geom_col(fill = "dodgerblue2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  facet_wrap(~ Log2FC, ncol = 3) +
  labs(x = NULL, y = "Mean log2FC",
       title = "Mean log2FC per Organelle Annotation\n(Internal Norm + Ratio-of-Ratios)") +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank())

# --- Step 8. Save to PDF
ggsave(file.path(out_dir_diff136_organelle_dist_barnorm, "diff136_loc_dist_overview_selected_directComp_internalNorm_ratio.pdf"),loc_dist_overview_subset_ratio,width = 5, height = 4, device = cairo_pdf)

```


### IVbb. Resorted RoR, EulerTop5
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# IIIb-plus. Resort by RoR, save sorted plot, extract top-5 annotations,
# make Euler plots, and run GO on sig up/down for those annotations
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. RoR-sorted plot
ror_sort_order_vec <- ratio_df %>%
  arrange(desc(mean_log2FC)) %>%
  dplyr::pull(Annotation) %>%
  as.character() %>%
  unique()

# 1.1 mark which facets get the restricted range
combined_df_sortedByRoR <- combined_df %>%
  mutate(y_clamped = ifelse(Log2FC %in% c("log2FC_ASAH1_Axon.ASAH1",
                                          "log2FC_Ctrl_Axon.Ctrl"),
                            pmax(pmin(mean_log2FC, 0.5), -0.5),
                            mean_log2FC))

loc_dist_overview_subset_ratio_sorted <- ggplot(combined_df_sortedByRoR,aes(x = Annotation, y = y_clamped)) +
  geom_col(fill = "dodgerblue2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_x_discrete(limits = rev(ror_sort_order_vec), drop = FALSE) +  
  coord_flip() +
  facet_wrap(~ Log2FC, ncol = 3, scales = "free_y") +
  labs(x = NULL, y = "Mean log2FC",
       title = "Mean log2FC per Organelle Annotation (Internal Norm + Ratio-of-Ratios, RoR-sorted)") +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank())

ggsave(file.path(out_dir_diff136_organelle_dist_norm_ranked,"diff136_loc_dist_overview_selected_directComp_internalNorm_rorsorted_clamped.pdf"),loc_dist_overview_subset_ratio_sorted, width = 5, height = 4, device = cairo_pdf)


# 1.2 Build internal-norm-only df, ordered by RoR and clamped
comparisons_internal_norm <- c("log2FC_ASAH1_Axon.ASAH1", "log2FC_Ctrl_Axon.Ctrl")

combined_df_internalNorm_rorSorted <- combined_df %>%
  dplyr::filter(Log2FC %in% comparisons_internal_norm) %>%
  dplyr::mutate(
    Annotation = factor(as.character(Annotation), levels = ror_sort_order_vec),
    mean_log2FC = pmax(pmin(mean_log2FC, 0.6), -0.6)
  )

# 1.3 plot reordered RoR plot
loc_dist_overview_internalNorm_rorSorted <- ggplot(combined_df_internalNorm_rorSorted,aes(x = Annotation, y = mean_log2FC)) +
  geom_col(fill = "dodgerblue2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  # reverse limits so highest-RoR annotations appear at the top
  scale_x_discrete(limits = ror_sort_order_vec) +
  facet_wrap(~ Log2FC, ncol = 2, scales = "free_y") +
  labs(x = NULL, y = NULL,
       title = "Mean log2FC per Organelle Annotation\n(Internal Norm Only, RoR order, clamped)") +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank())

# 1.5 save plot
ggsave(file.path(out_dir_diff136_organelle_dist_norm_ranked,"diff136_loc_dist_overview_selected_directComp_internalNorm_clamped_rorSorted.pdf"),loc_dist_overview_internalNorm_rorSorted,width = 4, height = 4, device = cairo_pdf)



# --- 2. CSV of genes in top 5 annotations by RoR and Euler overlap plots
suppressPackageStartupMessages(require(eulerr))

top5_annotations_by_ror <- ratio_df %>%
  arrange(desc(mean_log2FC)) %>%
  slice_head(n = 5) %>%
  pull(Annotation) %>%
  as.character()

# 2.1 helper to get gene sets per annotation for internal-norm comparisons
get_internal_norm_gene_sets <- function(ann_term, df) {
  sub_df <- df %>% filter(.data[[ann_term]] == 1)
  asah1_pos <- sub_df %>%
    filter(is.finite(log2FC_ASAH1_Axon.ASAH1) & log2FC_ASAH1_Axon.ASAH1 > 0) %>%
    pull(Gene.Symbol) %>% unique()
  ctrl_pos <- sub_df %>%
    filter(is.finite(log2FC_Ctrl_Axon.Ctrl) & log2FC_Ctrl_Axon.Ctrl > 0) %>%
    pull(Gene.Symbol) %>% unique()
  list(ASAH1_only = setdiff(asah1_pos, ctrl_pos),
       Ctrl_only  = setdiff(ctrl_pos, asah1_pos),
       Common     = intersect(asah1_pos, ctrl_pos),
       ASAH1_all  = asah1_pos,
       Ctrl_all   = ctrl_pos)
}

# 2.2 build table
top5_gene_overlap_tbl <- purrr::map_dfr(top5_annotations_by_ror, function(ann_term) {
  gs <- get_internal_norm_gene_sets(ann_term, diff136_df_annotated)
  tibble(
    Annotation = ann_term,
    Set        = c("ASAH1_only", "Ctrl_only", "Common"),
    N          = c(length(gs$ASAH1_only), length(gs$Ctrl_only), length(gs$Common)),
    Genes      = c(paste(gs$ASAH1_only, collapse = ";"),
                   paste(gs$Ctrl_only,  collapse = ";"),
                   paste(gs$Common,     collapse = ";"))
  )
})

# 2.3 save table
csv_top5_genes_path <- file.path(out_dir_diff136_organelle_dist_norm_ranked, "top5_byRoR_gene_overlap_ASAH1_vs_Ctrl.csv")
write.csv(top5_gene_overlap_tbl, csv_top5_genes_path, row.names = FALSE)


# --- 3. Create & save Euler plots per annotation
invisible(lapply(top5_annotations_by_ror, function(ann_term) {
  gs <- get_internal_norm_gene_sets(ann_term, diff136_df_annotated)
  fit2 <- eulerr::euler(list(ASAH1 = gs$ASAH1_all, Ctrl = gs$Ctrl_all))
  p_euler <- plot(fit2, labels = TRUE, quantities = TRUE)
  grDevices::cairo_pdf(file.path(out_dir_diff136_organelle_dist_norm_ranked, paste0("euler_", ann_term, "_ASAH1_vs_Ctrl.pdf")),
                       width = 3.5, height = 3.0)
  print(p_euler)
  dev.off()
}))


# --- 4. GO on Euler overlap sets (Top5 RoR; only if set size >= 10)
# use in-memory overlap table if present, else read the saved CSV
go_overlap_tbl <- if (exists("top5_gene_overlap_tbl")) {
  message("Using in-memory top5_gene_overlap_tbl")
  top5_gene_overlap_tbl
} else {
  message("Loading top5_gene_overlap_tbl from CSV")
  read.csv(csv_top5_genes_path, stringsAsFactors = FALSE)
}

# make long table of unique genes per Annotation×Set
go_overlap_long <- go_overlap_tbl %>%
  mutate(Genes = ifelse(is.na(Genes), "", Genes)) %>%
  separate_rows(Genes, sep = ";") %>%
  mutate(Genes = trimws(Genes)) %>%
  filter(nzchar(Genes)) %>%
  distinct(Annotation, Set, Genes)

# manifest with counts, saved into the Euler outdir
go_overlap_counts <- go_overlap_long %>%
  count(Annotation, Set, name = "n_genes") %>%
  arrange(match(Annotation, top5_annotations_by_ror), Set)

write_csv(go_overlap_counts,
          file.path(out_dir_diff136_organelle_dist_norm_ranked, "GO_inputs_top5RoR_overlap_manifest_counts.csv"))

# helper to run GO for a given set if size >= 10, and save input CSV
.run_go_for_set <- function(ann_term, set_label) {
  genes_vec <- go_overlap_long %>%
    filter(Annotation == ann_term, Set == set_label) %>%
    pull(Genes) %>% unique()

  if (length(genes_vec) < 10) return(invisible(NULL))

  # save the exact gene input used
  write_csv(tibble(GeneSymbol = genes_vec),
            file.path(out_dir_diff136_organelle_dist_norm_ranked,
                      paste0("GO_inputs_Top5RoR_", ann_term, "_", set_label, ".csv")))

  # run GO; function is assumed pre-defined upstream
  plot_topGO_for_geneset(
    gene_set = genes_vec,
    background_genes = unique(diff136_df_annotated$Gene.Symbol),
    tag = paste0("Top5RoR_", ann_term, "_", set_label),
    out_dir = out_dir_diff136_organelle_dist_norm_ranked
  )

  invisible(NULL)
}

# run per-annotation and per-overlap set
sets_to_eval <- c("ASAH1_only", "Ctrl_only", "Common")
invisible(lapply(top5_annotations_by_ror, function(ann_term) {
  lapply(sets_to_eval, function(set_lab) .run_go_for_set(ann_term, set_lab))
}))


```

### IVbc. Sensitivity ranking with normalized comparisons (delta, Cohen's d, RoR)
```{r}
# • Δ mean log2FC – difference in normalized mean log2 fold-change between ASAH1 and Ctrl (\bar{x}{ASAH1} - \bar{x}{Ctrl}).
# • Cohen’s d – standardized effect size, Δ mean log2FC divided by pooled standard deviation across genes in the annotation.
#   => Positive Cohen’s d → mean log2FC in ASAH1 KO is higher than in Ctrl (effect skewed toward KO).
#   => Negative Cohen’s d → mean log2FC in Ctrl is higher than in ASAH1 KO (effect skewed toward control).
# • RoR – ratio-of-ratios from earlier analysis, magnitude of normalized fold-change difference relative to baseline.
# • Composite percentile rank – average of the percentile ranks for |Δ mean|, |Cohen’s d|, and |RoR|, giving a weighted consensus ranking.

# --- 1. Input setup
norm_cols_vec <- c("log2FC_ASAH1_Axon.ASAH1", "log2FC_Ctrl_Axon.Ctrl")
stopifnot(all(norm_cols_vec %in% colnames(diff136_df_annotated)))

# --- 2. Identify binary annotation columns
all_log2fc_cols <- grep("^log2FC_", colnames(diff136_df_annotated), value = TRUE)
annotation_cols_norm <- setdiff(colnames(diff136_df_annotated), all_log2fc_cols)
annotation_cols_norm <- annotation_cols_norm[
  sapply(diff136_df_annotated[, annotation_cols_norm], function(x) all(x %in% c(0,1,NA)))
]
annotation_cols_norm <- setdiff(annotation_cols_norm, c("Autophagy_det","EndoDetailLoac","Nucleus"))

# --- 3. Helper function to compute delta mean and Cohen's d per annotation
.compute_ann_stats <- function(ann) {
  sub <- diff136_df_annotated[diff136_df_annotated[[ann]] == 1, , drop = FALSE]
  a <- sub[[norm_cols_vec[1]]]; a <- a[is.finite(a)]
  c <- sub[[norm_cols_vec[2]]]; c <- c[is.finite(c)]
  n_a <- length(a); n_c <- length(c)

  m_a <- if (n_a) mean(a) else NA_real_
  m_c <- if (n_c) mean(c) else NA_real_
  sd_a <- if (n_a > 1) sd(a) else NA_real_
  sd_c <- if (n_c > 1) sd(c) else NA_real_

  delta_mean <- m_a - m_c

  # --- 3a. Cohen's d (pooled SD); NA if not enough genes
  d_val <- NA_real_
  if (n_a > 1 && n_c > 1) {
    s_pooled <- sqrt( ((n_a - 1)*sd_a^2 + (n_c - 1)*sd_c^2) / (n_a + n_c - 2) )
    if (is.finite(s_pooled) && s_pooled > 0) d_val <- delta_mean / s_pooled
  }

  tibble(
    Annotation = ann,
    n_genes = nrow(sub),
    mean_ASAH1_norm = m_a,
    mean_Ctrl_norm  = m_c,
    delta_mean      = delta_mean,
    cohen_d         = d_val
  )
}

# --- 4. Compute annotation stats for all annotations
ann_stats_tbl <- purrr::map_dfr(annotation_cols_norm, .compute_ann_stats)

# --- 5. Merge with RoR values from upstream
ror_tbl <- ratio_df %>%
  dplyr::filter(Log2FC == "RatioOfRatios") %>%
  dplyr::select(Annotation, ror_mean = mean_log2FC)

ann_stats_tbl <- ann_stats_tbl %>%
  left_join(ror_tbl, by = "Annotation")

# --- 6. Compute percentile ranks
rank_tbl <- ann_stats_tbl %>%
  mutate(
    abs_delta = abs(delta_mean),
    abs_d     = abs(cohen_d),
    abs_ror   = abs(ror_mean),

    pr_delta  = dplyr::percent_rank(replace_na(abs_delta, -Inf)),
    pr_d      = dplyr::percent_rank(replace_na(abs_d, -Inf)),
    pr_ror    = dplyr::percent_rank(replace_na(abs_ror, -Inf))
  )

# --- 7. Compute composite rank (mean of component percentiles)
rank_tbl <- rank_tbl %>%
  rowwise() %>%
  mutate(
    composite_pr = mean(c(pr_delta, pr_d, pr_ror)[is.finite(c(pr_delta, pr_d, pr_ror))])
  ) %>%
  ungroup() %>%
  arrange(desc(composite_pr))

# --- 8. Save ranking table
out_rank_dir <- file.path(out_dir_diff136_organelle_dist_norm_composite, "ranking_sensitivity")
dir.create(out_rank_dir, showWarnings = FALSE, recursive = TRUE)
readr::write_csv(rank_tbl, file.path(out_rank_dir, "RankTable_annotation_ranking_norm_delta_d_ror.csv"))

# --- 9a. Heatmap of ranking metrics
# order rows by ascending composite rank
ord_levels <- rev(rank_tbl$Annotation)

heatmap_mat <- rank_tbl %>%
  dplyr::select(Annotation, delta_mean, cohen_d, ror_mean, pr_delta, pr_d, pr_ror, composite_pr) %>%
  as.data.frame()
rownames(heatmap_mat) <- heatmap_mat$Annotation
heatmap_mat$Annotation <- NULL
heatmap_mat <- heatmap_mat[ord_levels, , drop = FALSE]

# make heatmap
pheatmap(
  mat = heatmap_mat,
  scale = "column",
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  #color = viridis(100),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = NA,
  fontsize_row = 6,
  fontsize_col = 6,
  cellwidth = 6,
  cellheight = 6,
  filename = file.path(out_rank_dir, "RankTable_annotation_ranking_heatmap.pdf"),
  width = 5,
  height = 6
)


# --- 9b. Scatter: x = pr_ror, y = composite_pr, size = pr_d, alpha = pr_delta, color = Annotation
pr_dot_df <- rank_tbl %>%
  dplyr::select(Annotation, pr_ror, pr_d, pr_delta, composite_pr) %>%
  mutate(Annotation = factor(Annotation, levels = ord_levels))

# palette for discrete Annotation colors
n_ann <- length(levels(pr_dot_df$Annotation))
ann_cols <- setNames(colorRampPalette(rev(brewer.pal(11, "RdBu")))(n_ann),
                     levels(pr_dot_df$Annotation))

# plot dotplot w/ annotation text
pr_dot <- ggplot(pr_dot_df, aes(x = pr_ror, y = composite_pr)) +
  geom_point(aes(size = pr_d, alpha = pr_delta, color = Annotation)) +
  geom_text_repel(aes(label = Annotation), size = 2, max.overlaps = Inf, show.legend = FALSE) +
  scale_color_manual(values = ann_cols, name = "Annotation") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_size(range = c(1.5, 6), limits = c(0, 1), name = "pr(d)") +
  scale_alpha(range = c(0.3, 1), limits = c(0, 1), name = "pr(Δ)") +
  labs(x = "pr(RoR)", y = "Composite percentile", title = "Percentile metrics by annotation") +
  theme_bw(base_size = 7) +
  theme(
    panel.grid = element_blank(), 
    legend.position = "bottom")

ggsave(file.path(out_rank_dir, "ranking_scatter_pr_metrics_labels.pdf"),
       pr_dot, width = 5, height = 4, device = cairo_pdf)

# --- 9c. Plot delta and Cohen's d using ascending composite order
ord_levels <- rev(rank_tbl$Annotation)

p_delta <- ggplot(rank_tbl, aes(x = factor(Annotation, levels = ord_levels), y = delta_mean)) +
  geom_col(fill = "grey60") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(x = NULL, y = "Δ mean log2FC (ASAH1−Ctrl)",
       title = "Annotation ranking by composite percentile (ascending)") +
  theme_bw(base_size = 6) + theme(panel.grid = element_blank())

p_d <- ggplot(rank_tbl, aes(x = factor(Annotation, levels = ord_levels), y = cohen_d)) +
  geom_col(fill = "dodgerblue3") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(x = NULL, y = "Cohen's d",
       title = "Annotation ranking by composite percentile (ascending)") +
  theme_bw(base_size = 6) + theme(panel.grid = element_blank())

p_comp_rank <- ggplot(rank_tbl, aes(x = factor(Annotation, levels = ord_levels), y = composite_pr)) +
  geom_col(fill = "grey30") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Composite percentile (0 to 1)",
    title = "Annotation ranking by composite percentile (ascending)"
  ) +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank())

ggsave(file.path(out_rank_dir, "ranking_delta_compositeOrder_asc.pdf"), p_delta, width = 4, height = 5, device = cairo_pdf)
ggsave(file.path(out_rank_dir, "ranking_cohensd_compositeOrder_asc.pdf"), p_d, width = 4, height = 5, device = cairo_pdf)
ggsave(file.path(out_rank_dir, "ranking_compositePercentile_asc.pdf"), p_comp_rank, width = 4, height = 5, device = cairo_pdf)

# --- 10. Save top-5 lists for concordance check
topk <- 5
top_by_delta <- rank_tbl %>% arrange(desc(abs_delta)) %>% slice_head(n = topk) %>% pull(Annotation)
top_by_d     <- rank_tbl %>% arrange(desc(abs_d))     %>% slice_head(n = topk) %>% pull(Annotation)
top_by_ror   <- rank_tbl %>% arrange(desc(abs_ror))   %>% slice_head(n = topk) %>% pull(Annotation)
top_by_comp  <- rank_tbl %>% slice_head(n = topk) %>% pull(Annotation)

concordance_tbl <- tibble(
  rank = 1:topk,
  by_delta = top_by_delta,
  by_cohens_d = top_by_d,
  by_ror = top_by_ror,
  by_composite = top_by_comp
)
readr::write_csv(concordance_tbl, file.path(out_rank_dir, "top5_concordance_delta_d_ror_composite.csv"))

# --- 11. Replot loc_dist_overview with ascending composite order
loc_dist_overview_internalNorm_compAsc <- loc_dist_overview_internalNorm_rorSorted +
  scale_x_discrete(limits = ord_levels)

ggsave(
  file.path(out_dir_diff136_organelle_dist_norm_composite,
            "diff136_loc_dist_overview_selected_directComp_internalNorm_clamped_compAsc.pdf"),
  loc_dist_overview_internalNorm_compAsc,
  width = 5, height = 4, device = cairo_pdf
)
```


### IVbd. Top5 COMPOSITE GO eval
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# GO for top-5 annotations by composite rank: build inputs (up/down) + run GO
# ─────────────────────────────────────────────────────────────────────────────

# --- 0. Preconditions
stopifnot(exists("rank_tbl"), "composite_pr" %in% colnames(rank_tbl))

# --- 1. Choose top-5 annotations by composite rank
top5_annotations_by_comp <- rank_tbl %>%
  arrange(desc(composite_pr)) %>%
  slice_head(n = 5) %>%
  pull(Annotation) %>%
  as.character()

# --- 2. Reuse sig list if present; fallback to CSV
go_sig_tbl <- if (exists("sig_hits_tbl")) {
  message("Using in-memory sig_hits_tbl")
  sig_hits_tbl
} else {
  fallback_path <- file.path(out_dir_diff136_organelle_dist, "diff136_sig_hits_annotation_geneList.csv")
  if (!file.exists(fallback_path)) stop("sig_hits_tbl not in memory and CSV not found at: ", fallback_path)
  message("Loading sig_hits_tbl from CSV fallback")
  read.csv(fallback_path, stringsAsFactors = FALSE)
}

# --- 3. Filter to top-5 by composite and up/down only
go_ann_to_use_comp <- intersect(unique(go_sig_tbl$Annotation), top5_annotations_by_comp)

sig_hits_top5_comp <- go_sig_tbl %>%
  dplyr::filter(Annotation %in% go_ann_to_use_comp,
                SigStatus %in% c("up","down")) %>%
  dplyr::mutate(GeneList = ifelse(is.na(GeneList), "", GeneList))

# --- 4. Build GO input tables (long and wide)
sig_hits_top5_comp_long <- sig_hits_top5_comp %>%
  tidyr::separate_rows(GeneList, sep = ";", convert = FALSE) %>%
  dplyr::rename(GeneSymbol = GeneList) %>%
  dplyr::filter(nzchar(GeneSymbol)) %>%
  dplyr::distinct(Annotation, SigStatus, GeneSymbol)

readr::write_csv(sig_hits_top5_comp_long, file.path(out_dir_diff136_organelle_dist_norm_composite, "GO_inputs_top5Composite_up_down_long.csv"))

sig_hits_top5_comp_wide <- sig_hits_top5_comp_long %>%
  dplyr::group_by(Annotation, SigStatus) %>%
  dplyr::summarise(GeneList = paste(sort(unique(GeneSymbol)), collapse = ";"), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = SigStatus, values_from = GeneList)

readr::write_csv(sig_hits_top5_comp_wide, file.path(out_dir_diff136_organelle_dist_norm_composite, "GO_inputs_top5Composite_up_down_wide.csv"))

# --- 5. Run GO per annotation × direction (writes plots + GO_results_Top5Comp_<ann>_<dir>.csv)
invisible(lapply(go_ann_to_use_comp, function(ann_term) {
  lapply(c("up","down"), function(dir_flag) {
    gene_vec <- sig_hits_top5_comp_long %>%
      dplyr::filter(Annotation == ann_term, SigStatus == dir_flag) %>%
      dplyr::pull(GeneSymbol) %>%
      unique()
    if (length(gene_vec) >= 5) {
      plot_topGO_for_geneset(
        gene_set = gene_vec,
        background_genes = unique(diff136_df_annotated$Gene.Symbol),
        tag = paste0("Top5Comp_", ann_term, "_", dir_flag),
        out_dir = out_dir_diff136_organelle_dist_norm_composite
      )
    }
    NULL
  })
}))

# --- 6. Optional: also run GO on the combined (up ∪ down) set per annotation
invisible(lapply(go_ann_to_use_comp, function(ann_term) {
  gene_vec <- sig_hits_top5_comp_long %>%
    dplyr::filter(Annotation == ann_term) %>%
    dplyr::pull(GeneSymbol) %>%
    unique()
  if (length(gene_vec) >= 5) {
    plot_topGO_for_geneset(
      gene_set = gene_vec,
      background_genes = unique(diff136_df_annotated$Gene.Symbol),
      tag = paste0("Top5Comp_", ann_term, "_combined"),
      out_dir = out_dir_diff136_organelle_dist_norm_composite
    )
  }
  NULL
}))

# --- 7. Summary for QC purposes
go_manifest_comp <- sig_hits_top5_comp_long %>%
  dplyr::count(Annotation, SigStatus, name = "n_genes") %>%
  tidyr::complete(Annotation = top5_annotations_by_comp, SigStatus = c("up","down"), fill = list(n_genes = 0)) %>%
  dplyr::arrange(match(Annotation, top5_annotations_by_comp), SigStatus)

readr::write_csv(go_manifest_comp, file.path(out_dir_diff136_organelle_dist_norm_composite, "GO_inputs_top5Composite_manifest_counts.csv"))
```


### IVbe. Top5 COMPOSITE — Euler overlaps + GO on unique/common
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Euler analysis & GO for top-5 annotations by composite rank 
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Top-5 by composite rank
top5_annotations_by_comp <- rank_tbl %>%
  arrange(desc(composite_pr)) %>%
  slice_head(n = 5) %>%
  pull(Annotation) %>%
  as.character()

# --- 2. Helper to get gene sets for internal-norm comparisons (reuse from above if already defined)
get_internal_norm_gene_sets <- function(ann_term, df) {
  sub_df <- df %>% filter(.data[[ann_term]] == 1)
  asah1_pos <- sub_df %>%
    filter(is.finite(log2FC_ASAH1_Axon.ASAH1) & log2FC_ASAH1_Axon.ASAH1 > 0) %>%
    pull(Gene.Symbol) %>% unique()
  ctrl_pos <- sub_df %>%
    filter(is.finite(log2FC_Ctrl_Axon.Ctrl) & log2FC_Ctrl_Axon.Ctrl > 0) %>%
    pull(Gene.Symbol) %>% unique()
  list(
    ASAH1_only = setdiff(asah1_pos, ctrl_pos),
    Ctrl_only  = setdiff(ctrl_pos, asah1_pos),
    Common     = intersect(asah1_pos, ctrl_pos),
    ASAH1_all  = asah1_pos,
    Ctrl_all   = ctrl_pos
  )
}

# --- 3. Build overlap table for composite top-5
top5_comp_gene_overlap_tbl <- purrr::map_dfr(top5_annotations_by_comp, function(ann_term) {
  gs <- get_internal_norm_gene_sets(ann_term, diff136_df_annotated)
  tibble(
    Annotation = ann_term,
    Set        = c("ASAH1_only", "Ctrl_only", "Common"),
    N          = c(length(gs$ASAH1_only), length(gs$Ctrl_only), length(gs$Common)),
    Genes      = c(paste(gs$ASAH1_only, collapse = ";"),
                   paste(gs$Ctrl_only,  collapse = ";"),
                   paste(gs$Common,     collapse = ";"))
  )
})

# --- 4. Save the composite overlap CSV
write.csv(top5_comp_gene_overlap_tbl,file = file.path(out_dir_diff136_organelle_dist_norm_composite_euler, "top5_byComposite_gene_overlap_ASAH1_vs_Ctrl.csv"),row.names = FALSE)

# --- 5. Euler plots per annotation
invisible(lapply(top5_annotations_by_comp, function(ann_term) {
  gs <- get_internal_norm_gene_sets(ann_term, diff136_df_annotated)
  fit2 <- eulerr::euler(list(ASAH1 = gs$ASAH1_all, Ctrl = gs$Ctrl_all))
  p_euler <- plot(fit2, labels = TRUE, quantities = TRUE)
  grDevices::cairo_pdf(file.path(out_dir_diff136_organelle_dist_norm_composite_euler, paste0("eulerComposite_", ann_term, "_ASAH1_vs_Ctrl.pdf")),
                       width = 3.5, height = 3.0)
  print(p_euler)
  dev.off()
}))

# --- 6. GO on Euler overlap sets (only if set size ≥ 10); inputs also saved in same dir
go_overlap_tbl_comp <- top5_comp_gene_overlap_tbl

go_overlap_long_comp <- go_overlap_tbl_comp %>%
  mutate(Genes = ifelse(is.na(Genes), "", Genes)) %>%
  separate_rows(Genes, sep = ";") %>%
  mutate(Genes = trimws(Genes)) %>%
  filter(nzchar(Genes)) %>%
  distinct(Annotation, Set, Genes)

# manifest for QC
go_overlap_counts_comp <- go_overlap_long_comp %>%
  count(Annotation, Set, name = "n_genes") %>%
  arrange(match(Annotation, top5_annotations_by_comp), Set)

write_csv(go_overlap_counts_comp,
          file.path(out_dir_diff136_organelle_dist_norm_composite_euler, "GO_inputs_top5Composite_overlap_manifest_counts.csv"))

.run_go_for_set_comp <- function(ann_term, set_label) {
  genes_vec <- go_overlap_long_comp %>%
    filter(Annotation == ann_term, Set == set_label) %>%
    pull(Genes) %>% unique()

  if (length(genes_vec) < 10) return(invisible(NULL))

  # save inputs actually used
  write_csv(tibble(GeneSymbol = genes_vec),
            file.path(out_dir_diff136_organelle_dist_norm_composite_euler, paste0("GO_inputs_Top5Composite_", ann_term, "_", set_label, ".csv")))

  plot_topGO_for_geneset(
    gene_set = genes_vec,
    background_genes = unique(diff136_df_annotated$Gene.Symbol),
    tag = paste0("Top5Composite_", ann_term, "_", set_label),
    out_dir = out_dir_diff136_organelle_dist_norm_composite_euler
  )
  invisible(NULL)
}

sets_to_eval <- c("ASAH1_only", "Ctrl_only", "Common")
invisible(lapply(top5_annotations_by_comp, function(ann_term) {
  lapply(sets_to_eval, function(set_lab) .run_go_for_set_comp(ann_term, set_lab))
}))

```



## IVc. Barplots of mean(log2FC) of synaptic annotations
```{r}
# --- Step 1. Define columns and color mapping
plot_cols <- c("log2FC_ASAH1_Axon.ASAH1_Soma", "log2FC_Ctrl_Axon.Ctrl_Soma")
condition_labels <- c("ASAH1", "Ctrl")
names(condition_labels) <- plot_cols

fill_colors <- c("ASAH1" = "dodgerblue", "Ctrl" = "grey70")

selected_annotations <- c("SVfusion", "Svendocytosis", "Svexocytosis", 
                          "SynapseSVs", "Presynaptic", "Postsynaptic","vATPase")

vatpase_genes <- c("ATP6V1A","ATP6V1B1","ATP6V1B2","ATP6V1C1","ATP6V1C2","ATP6V1D","ATP6V1E1","ATP6V1E2","ATP6V1F","ATP6V1G1","ATP6V1G2","ATP6V1G3","ATP6V1H","ATP6V0A1","ATP6V0A2","ATP6V0A3","ATP6V0A4",
  "ATP6V0C","ATP6V0B","ATP6V0D1","ATP6V0D2","ATP6V0E1","ATP6V0E2","RNASEK","ATP6AP1","ATP6AP2","DMXL1","DMXL2","WDR7","ROGDI")


# --- Step 2. Prepare data
summary_sv_df <- diff136_df_annotated %>%
  filter(Gene.Symbol %in% vatpase_genes | 
           rowSums(across(all_of(selected_annotations)), na.rm = TRUE) > 0) %>%
  dplyr::select(Gene.Symbol, all_of(plot_cols), all_of(selected_annotations)) %>%
  pivot_longer(cols = all_of(plot_cols), names_to = "Comparison", values_to = "log2FC") %>%
  pivot_longer(cols = all_of(selected_annotations), names_to = "Annotation", values_to = "is_annotated") %>%
  filter(is_annotated == 1, is.finite(log2FC)) %>%
  mutate(Condition = recode(Comparison, !!!condition_labels)) %>%
  group_by(Annotation, Condition) %>%
  summarise(
    mean = mean(log2FC, na.rm = TRUE),
    sem  = sd(log2FC, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )
# --- Step 2b. Prepare per-gene data for stats (before summarise)
sv_long_df <- diff136_df_annotated %>%
  filter(Gene.Symbol %in% vatpase_genes | 
           rowSums(across(all_of(selected_annotations)), na.rm = TRUE) > 0) %>%
  dplyr::select(Gene.Symbol, all_of(plot_cols), all_of(selected_annotations)) %>%
  pivot_longer(cols = all_of(plot_cols), names_to = "Comparison", values_to = "log2FC") %>%
  pivot_longer(cols = all_of(selected_annotations), names_to = "Annotation", values_to = "is_annotated") %>%
  filter(is_annotated == 1, is.finite(log2FC)) %>%
  mutate(Condition = recode(Comparison, !!!condition_labels))

# --- Step 2c. Pairwise tests Ctrl vs ASAH1 per annotation
stat_results <- sv_long_df %>%
  group_by(Annotation) %>%
  summarise(
    n_ASAH1 = sum(Condition == "ASAH1"),
    n_Ctrl  = sum(Condition == "Ctrl"),
    .test   = list({
      x <- log2FC[Condition == "ASAH1"]
      y <- log2FC[Condition == "Ctrl"]
      if (length(x) > 1 & length(y) > 1) {
        wt <- wilcox.test(x, y, exact = FALSE)  # Mann-Whitney
        tibble(
          method = wt$method,
          p_value = wt$p.value,
          median_ASAH1 = median(x, na.rm=TRUE),
          median_Ctrl  = median(y, na.rm=TRUE),
          delta_median = median(x, na.rm=TRUE) - median(y, na.rm=TRUE)
        )
      } else {
        tibble(method = NA, p_value = NA,
               median_ASAH1 = NA, median_Ctrl = NA, delta_median = NA)
      }
    }),
    .groups = "drop"
  ) %>%
  unnest(cols = .test) %>%
  mutate(
    p_adj_BH  = p.adjust(p_value, method = "BH"),
    p_adj_FDR = p.adjust(p_value, method = "fdr"),
    signif_label = case_when(
      p_value <= 0.001 ~ "***",
      p_value <= 0.01  ~ "**",
      p_value <= 0.05  ~ "*",
      TRUE             ~ "ns"
    )
  )

# --- Step 2d. Save stats
readr::write_csv(stat_results,file.path(out_dir_diff136_organelle_dist,"diff136_synaptic_annotation_ASAH1_vs_Ctrl_tests.csv"))


# --- Step 3. Plot means + SEM
barplot_SV_means_sem <- ggplot(summary_sv_df, aes(x = Annotation, y = mean, fill = Condition)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.8, color = "black") +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), 
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_manual(values = fill_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank()) +
  labs(
    title = "Mean log2FC ± SEM for Synaptic Annotations",
    x = NULL,
    y = "log2FC",
    fill = "Condition"
  )

barplot_SV_means_sem


# --- Step 4. Save PDF
ggsave(file.path(out_dir_diff136_organelle_dist, "diff136_vatpase_synaptic_barplot_means_sem.pdf"),
       barplot_SV_means_sem, width = 3, height = 3, device = cairo_pdf)


```


## IVd.  Comparison of Neuron Projection Proteome 
### IVda.  Absolute & RoR
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Analysis of neuronal projection proteome in Ctrl & ASAH1-/- neurons 
# Absolute (genotype vs WC) and Ratio of Ratio (RoR), norm. to WC in genotype
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1. Calculate RoR
diff136_df_annotated <- diff136_df_annotated %>%
  dplyr::mutate(RoR_ASAH1_Axon = `log2FC_ASAH1_Axon.ASAH1` - `log2FC_Ctrl_Axon.Ctrl`)

# --- Step 2. Threshold
fc_thresh <- 0.585

# --- Step 3. Call RoR significance
diff136_df_annotated <- diff136_df_annotated %>%
  dplyr::mutate(sig_RoR_ASAH1_Axon = case_when(
    is.na(RoR_ASAH1_Axon)        ~ NA_character_,
    RoR_ASAH1_Axon > fc_thresh   ~ "up",
    RoR_ASAH1_Axon < -fc_thresh  ~ "down",
    TRUE                         ~ "n.s."
  ),
  enriched_in_RoR = case_when(
    sig_RoR_ASAH1_Axon == "up"   ~ "KO",
    sig_RoR_ASAH1_Axon == "down" ~ "Ctrl",
    TRUE ~ NA_character_
  ))

# --- Step 4. Absolute direction in each genotype
diff136_df_annotated <- diff136_df_annotated %>%
  dplyr::mutate(
    abs_dir_ctrl = case_when(
      `log2FC_Ctrl_Axon.Ctrl`  >  fc_thresh ~ "accumulating",
      `log2FC_Ctrl_Axon.Ctrl`  < -fc_thresh ~ "decreased",
      TRUE ~ "n.s."
    ),
    abs_dir_ko = case_when(
      `log2FC_ASAH1_Axon.ASAH1` >  fc_thresh ~ "accumulating",
      `log2FC_ASAH1_Axon.ASAH1` < -fc_thresh ~ "decreased",
      TRUE ~ "n.s."
    )
  )

# --- Step 5. Define relevant columns
sig_cols_abs <- c("sig_ASAH1_Axon.ASAH1_Soma", "sig_Ctrl_Axon.Ctrl_Soma")
sig_cols_ror <- "sig_RoR_ASAH1_Axon"
annotation_cols <- setdiff(names(diff136_df_annotated),
                            c(sig_cols_abs, sig_cols_ror))
annotation_cols <- annotation_cols[sapply(diff136_df_annotated[, annotation_cols],
                                           function(x) all(x %in% c(0, 1, NA)))]
annotation_cols <- setdiff(annotation_cols, "Nucleus")

# --- Step 6. Define 4-color palette
fill_palette4 <- c(
  "Ctrl_down"   = "grey50",
  "Ctrl_up"     = "grey70",
  "ASAH1_down"  = "dodgerblue1",
  "ASAH1_up"    = "dodgerblue3"
)

# --- Step 7. Long-format Absolute with Call
sig_long_abs <- diff136_df_annotated %>%
  dplyr::select(Gene.Symbol, all_of(sig_cols_abs), all_of(annotation_cols)) %>%
  pivot_longer(cols = all_of(sig_cols_abs), names_to = "Comparison", values_to = "SigStatus") %>%
  pivot_longer(cols = all_of(annotation_cols), names_to = "Annotation", values_to = "is_annotated") %>%
  filter(is_annotated == 1, SigStatus %in% c("up", "down")) %>%
  mutate(Type = "Absolute",
         Call = case_when(
           Comparison == "sig_Ctrl_Axon.Ctrl_Soma"   & SigStatus == "up"   ~ "Ctrl_up",
           Comparison == "sig_Ctrl_Axon.Ctrl_Soma"   & SigStatus == "down" ~ "Ctrl_down",
           Comparison == "sig_ASAH1_Axon.ASAH1_Soma" & SigStatus == "up"   ~ "ASAH1_up",
           Comparison == "sig_ASAH1_Axon.ASAH1_Soma" & SigStatus == "down" ~ "ASAH1_down"
         )) %>%
  filter(!is.na(Call))

# --- Step 8. Long-format RoR with Call
sig_long_ror <- diff136_df_annotated %>%
  dplyr::select(Gene.Symbol, sig_RoR_ASAH1_Axon, enriched_in_RoR,
         abs_dir_ctrl, abs_dir_ko, all_of(annotation_cols)) %>%
  pivot_longer(cols = all_of(annotation_cols), names_to = "Annotation", values_to = "is_annotated") %>%
  filter(is_annotated == 1, sig_RoR_ASAH1_Axon %in% c("up", "down")) %>%
  transmute(
    Gene.Symbol,
    Comparison = "sig_RoR_ASAH1_Axon",
    SigStatus  = sig_RoR_ASAH1_Axon,
    Annotation,
    Type       = "RatioOfRatio",
    AbsDir_KO  = abs_dir_ko,
    AbsDir_Ctrl= abs_dir_ctrl,
    Call = case_when(
      SigStatus == "up"   & AbsDir_KO   == "accumulating" ~ "ASAH1_up",
      SigStatus == "up"   & AbsDir_KO   == "decreased"    ~ "ASAH1_down",
      SigStatus == "down" & AbsDir_Ctrl == "accumulating" ~ "Ctrl_up",
      SigStatus == "down" & AbsDir_Ctrl == "decreased"    ~ "Ctrl_down"
    )
  ) %>% filter(!is.na(Call))

# --- Step 9. Combine and summarise
sig_long_df <- bind_rows(sig_long_abs, sig_long_ror)

sig_summary <- sig_long_df %>%
  group_by(Type, Comparison, Annotation, Call) %>%
  summarise(n = n(), .groups = "drop")

sig_gene_summary <- sig_long_df %>%
  group_by(Type, Comparison, Annotation, Call) %>%
  summarise(n = n(),
            GeneList = paste(unique(Gene.Symbol), collapse = ";"),
            .groups = "drop")

# --- Step 10. Plot Absolute
sig_plot_abs <- ggplot(filter(sig_summary, Type == "Absolute"),
                       aes(x = reorder(Annotation, n), y = n, fill = Call)) +
  geom_col() + coord_flip() +
  facet_wrap(~ Comparison, ncol = 1) +
  scale_fill_manual(values = fill_palette4, drop = FALSE) +
  labs(x = NULL, y = "# Significant Hits", title = "Absolute Comparisons (4-way call)") +
  theme_bw(base_size = 8) + theme(panel.grid = element_blank())

ggsave(file.path(out_dir_diff136_organelle_dist_ror_norm,"diff136_sig_hits_absolute_barplot_4colors.pdf"),sig_plot_abs, width = 5, height = 8, device = cairo_pdf)

# --- Step 11. Plot RoR
sig_plot_ror <- ggplot(filter(sig_summary, Type == "RatioOfRatio"),
                       aes(x = reorder(Annotation, n), y = n, fill = Call)) +
  geom_col() + coord_flip() +
  facet_wrap(~ Comparison, ncol = 1) +
  scale_fill_manual(values = fill_palette4, drop = FALSE) +
  labs(x = NULL, y = "# Significant Hits", title = "RoR Comparisons (4-way call)") +
  theme_bw(base_size = 8) + theme(panel.grid = element_blank())

ggsave(file.path(out_dir_diff136_organelle_dist_ror_norm,"diff136_sig_hits_ror_barplot_4colors.pdf"), sig_plot_ror, width = 6, height = 8, device = cairo_pdf)

# --- Step 12. Save CSVs
write.csv(sig_gene_summary,file = file.path(out_dir_diff136_organelle_dist_ror_norm, "diff136_sig_hits_annotation_geneList_withSig.csv"),row.names = FALSE)

write.csv(sig_long_df,file = file.path(out_dir_diff136_organelle_dist_ror_norm, "diff136_sig_hits_long_withSig.csv"),row.names = FALSE)


```



### IVdb. GO eval of IDs in TOP 5 sig up / down categories 
```{r}
# --------------------------------------------------
# GO on RoR hits split into four Call classes
# calculates for: Ctrl_up, Ctrl_down, ASAH1_up, ASAH1_down
# --------------------------------------------------

# --- 1. Guard and background
stopifnot("Call" %in% names(sig_long_df))
background_genes_vec <- unique(diff136_df_annotated$Gene.Symbol)
min_genes_for_go <- 5

# --- 2. Count RoR hits per Annotation × Call, get top-5 per Call
ror_counts_by_call <- sig_long_df %>%
  dplyr::filter(Type == "RatioOfRatio", Annotation != "Nucleus") %>%
  dplyr::count(Call, Annotation, name = "n") %>%
  dplyr::arrange(Call, dplyr::desc(n)) %>%
  dplyr::group_by(Call) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::ungroup()

# --- 3. Long gene table for selected top-5 per Call
go_ror_long <- sig_long_df %>%
  dplyr::filter(Type == "RatioOfRatio",
                Annotation %in% ror_counts_by_call$Annotation,
                Call %in% ror_counts_by_call$Call) %>%
  dplyr::select(Call, Annotation, Gene.Symbol) %>%
  dplyr::distinct()

# --- 4. Save manifests
readr::write_csv(ror_counts_by_call,
                 file.path(out_dir_diff136_GOror, "RoR_top5_annotation_counts_byCall_manifest.csv"))
readr::write_csv(go_ror_long,
                 file.path(out_dir_diff136_GOror, "GO_inputs_RoR_top5_byCall_long.csv"))

# --- 5. Run GO for each Call × Annotation if set size >= threshold
calls_order <- c("Ctrl_up","Ctrl_down","ASAH1_up","ASAH1_down")
all_go_results <- list()

invisible(lapply(calls_order, function(call_flag) {
  ann_subset <- ror_counts_by_call %>%
    dplyr::filter(Call == call_flag) %>%
    dplyr::pull(Annotation) %>%
    unique()

  lapply(ann_subset, function(ann_term) {
    gene_vec <- go_ror_long %>%
      dplyr::filter(Call == call_flag, Annotation == ann_term) %>%
      dplyr::pull(Gene.Symbol) %>%
      unique()

    # Save exact input list
    readr::write_csv(tibble::tibble(GeneSymbol = gene_vec),
                     file.path(out_dir_diff136_GOror,
                               paste0("GO_inputs_RoR_top5_", call_flag, "_", ann_term, ".csv")))

    if (length(gene_vec) >= min_genes_for_go) {
      # 5.1 Build 0/1 vector over background
      geneList <- as.integer(background_genes_vec %in% gene_vec)
      names(geneList) <- background_genes_vec

      # 5.2 topGO Fisher test (BP)
      GOdata <- new("topGOdata",
                    ontology = "BP",
                    allGenes = geneList,
                    geneSel = function(x) x == 1,
                    annot = annFUN.org,
                    mapping = "org.Hs.eg.db",
                    ID = "symbol")
      resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

      # 5.3 Generate GO table with extra fields
      go_table <- GenTable(GOdata, classicFisher = resultFisher, topNodes = 30) %>%
        dplyr::mutate(
          raw_pval    = score(resultFisher)[GO.ID],
          Significant = as.numeric(Significant),
          Expected    = as.numeric(Expected),
          Enrichment  = Significant / Expected,
          log10P      = -log10(raw_pval)
        ) %>%
        dplyr::filter(!is.na(raw_pval)) %>%
        dplyr::arrange(raw_pval) %>%
        dplyr::slice_head(n = 10)

      # 5.4 Add associated genes per GO term
      go_genes <- genesInTerm(GOdata)
      go_table$Genes <- vapply(go_table$GO.ID, function(go_id) {
        paste(intersect(unname(go_genes[[go_id]]), gene_vec), collapse = ";")
      }, character(1))

      # 5.5 Save per run CSV and collect for master table
      tag_str <- paste0("RoR_top5_", call_flag, "_", ann_term)
      readr::write_csv(go_table, file.path(out_dir_diff136_GOror, paste0("GO_results_", tag_str, ".csv")))
      all_go_results[[length(all_go_results) + 1]] <<- dplyr::mutate(go_table,
        Call = call_flag, Annotation = ann_term, Tag = tag_str)
    }
    NULL
  })
}))

# --- 6. Save master GO CSV with Call and Annotation
if (length(all_go_results) > 0) {
  master_go <- dplyr::bind_rows(all_go_results)
  readr::write_csv(master_go,
                   file.path(out_dir_diff136_GOror, "GO_results_RoR_top5_byCall_MASTER.csv"))
}
```


# V. Rainbowplot of annotations markers 
## Va. Rainbowplot of Lysosome & vATPase Members vs nuclear markers 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Rainfall plot for highlighting protein localization
# Lysosome & vATPase Members vs nuclear markers 
# ─────────────────────────────────────────────────────────────────────────────
library(ggplot2)
library(ggdist)
#install.packages("gghalves")
library(gghalves)

# --- Step 0. Define vATPase and lysohydrolase gene lists
vatpase_genes <- c("ATP6V1A","ATP6V1B1","ATP6V1B2","ATP6V1C1","ATP6V1C2","ATP6V1D","ATP6V1E1","ATP6V1E2","ATP6V1F","ATP6V1G1","ATP6V1G2","ATP6V1G3","ATP6V1H","ATP6V0A1","ATP6V0A2","ATP6V0A3","ATP6V0A4",
  "ATP6V0C","ATP6V0B","ATP6V0D1","ATP6V0D2","ATP6V0E1","ATP6V0E2","RNASEK","ATP6AP1","ATP6AP2","DMXL1","DMXL2","WDR7","ROGDI")

lysohydrolases <- c(
  "ASAH1","GBA","ARSA","ARSB","GALNS","IDUA","GLA","NEU1","MAN2B1",
  "GLB1","HEXA","GNS","HEXB","CTSA","CTSZ","TPP1","NAGLU","CTSL",
  "CTSB","NAGA","LGMN"
)

# --- Step 1. Prepare scatter_df with categorical y-axis
scatter_df <- diff136_df_annotated %>%
  filter(
    Gene.Symbol %in% vatpase_genes |
    Gene.Symbol %in% lysohydrolases |
    Nucleus == 1 |
    Lysosome == 1
  ) %>%
  dplyr::mutate(
    y_group = case_when(
      Gene.Symbol %in% lysohydrolases ~ "Hydrolases",
      Gene.Symbol %in% vatpase_genes  ~ "vATPase",
      Lysosome == 1                   ~ "Lysosome",
      #Nucleus == 1                    ~ "Nucleus",
      TRUE                            ~ NA_character_
    )
  ) %>%
  filter(!is.na(y_group)) %>%
  mutate(
    # put Nucleus at bottom in the flipped plot
    y_group = factor(y_group, levels = c("Nucleus","Lysosome","Hydrolases","vATPase"))
  )

# --- Step 2a. Select top labels per group
n_per_group <- 5
label_df <- bind_rows(
  scatter_df %>%
    group_by(y_group) %>%
    slice_max(order_by = log2FC_Ctrl_Axon.Ctrl_Soma, n = n_per_group, with_ties = FALSE),
  scatter_df %>%
    group_by(y_group) %>%
    slice_min(order_by = log2FC_Ctrl_Axon.Ctrl_Soma, n = n_per_group, with_ties = FALSE)
) %>% ungroup()


# --- Step 3. Create plot
axonXsoma_loc_rainbowplot <- ggplot(
  scatter_df,
  aes(y = log2FC_Ctrl_Axon.Ctrl_Soma, x = y_group, fill = y_group)
) +
  gghalves::geom_half_point(
    side = "b",
    range_scale = 0.2,
    alpha = 0.2,
    size = 1,
    shape = 16
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  ggdist::stat_halfeye(
    adjust = 2,
    width = 0.6,
    .width = 0,
    justification = -0.2,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  # --- Step 3a. Label top pos/neg per annotation
  geom_text_repel(
    data = label_df,
    aes(label = Gene.Symbol),
    position = position_nudge(x = -0.25),
    size = 2,
    min.segment.length = 0,
    max.overlaps = Inf,
    box.padding = 0.2,
    point.padding = 0.1,
    segment.size = 0.2
  ) +
  scale_fill_manual(values = c(
    "Lysosome" = "grey70",
    "Hydrolases" = "firebrick",
    "vATPase" = "dodgerblue3"
  )) +
  coord_flip() +
  theme_bw(base_size = 6) +
  labs(y = "log2FC Control Projection vs Soma", x = NULL,
       title = "Localization Organelle Distribution in Axon vs Soma") +
  theme(panel.grid = element_blank(), legend.position = "none")

# --- Step 4. Save PDF
ggsave(file.path(out_dir_diff136_rainbow, "diff136_lysosome_hydrolases_vATPase_nucleus_scatter.pdf"),
  axonXsoma_loc_rainbowplot, width = 8, height = 3, device = cairo_pdf)


```

## Vb. Flexible "rainbowplot" for up to 3 annotation groups
```{r}
# --- Step  1. Function for plotting user-defined gene sets for axon-soma localization distribution
# Generates “rainbow” distribution plots for up to three annotation groups using Va-style aesthetics, with top/bottom N gene labels based on a reference log2FC.
# Takes a dataframe, annotation gene sets, and log2FC columns, and outputs one PDF plot per log2FC to the specified directory.


plot_annotation_rainbow3 <- function(
  df,
  gene_sets,
  log2fc_col,
  out_dir,
  reference_fc = NULL,
  n_top_labels = 5,
  color_map = c("Label1" = "grey70", "Label2" = "firebrick", "Label3" = "dodgerblue3")
) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  if (is.null(reference_fc)) reference_fc <- log2fc_col

  # normalize gene_sets to tibble Gene.Symbol,y_group
  if (is.character(gene_sets) && is.null(names(gene_sets))) {
    gene_sets <- setNames(gene_sets, gene_sets)
    gene_df <- purrr::imap_dfr(gene_sets, \(col_name, grp) tibble(Gene.Symbol = df$Gene.Symbol[df[[col_name]] == 1], y_group = grp))
  } else {
    gene_df <- purrr::imap_dfr(gene_sets, \(genes, grp) tibble(Gene.Symbol = genes, y_group = grp))
  }
  gene_df <- distinct(gene_df, Gene.Symbol, y_group)

  # enforce max 3 groups
  unique_groups <- unique(gene_df$y_group)
  if (length(unique_groups) > 3) {
    message("More than 3 groups supplied. Using the first 3: ", paste(unique_groups[1:3], collapse = ", "))
    gene_df <- filter(gene_df, y_group %in% unique_groups[1:3])
    unique_groups <- unique(gene_df$y_group)
  }

  # one row per gene; keep only needed columns
  df_gene <- df %>%
    group_by(Gene.Symbol) %>%
    summarise(!!log2fc_col := mean(.data[[log2fc_col]], na.rm = TRUE),
              !!reference_fc := mean(.data[[reference_fc]], na.rm = TRUE),
              .groups = "drop")

  fc_df <- inner_join(df_gene, gene_df, by = "Gene.Symbol") %>%
    filter(is.finite(.data[[log2fc_col]])) %>%
    mutate(y_group = factor(y_group, levels = unique_groups))

  if (nrow(fc_df) == 0) {
    warning("No data to plot after filtering."); return(invisible(NULL))
  }

  # top/bottom labels ranked by reference_fc
  ref_df <- fc_df
  label_df <- bind_rows(
    ref_df %>% group_by(y_group) %>% slice_max(order_by = .data[[reference_fc]], n = n_top_labels, with_ties = FALSE),
    ref_df %>% group_by(y_group) %>% slice_min(order_by = .data[[reference_fc]], n = n_top_labels, with_ties = FALSE)
  ) %>% ungroup() %>%
    transmute(Gene.Symbol, y_group, value = .data[[log2fc_col]])

  # colors
  # fixed palette values in order; names are ignored
  palette_vals <- unname(color_map)
  if (length(palette_vals) == 0) palette_vals <- c("grey70","firebrick","dodgerblue3")
  
  # assign colors by the order of groups being plotted (max 3, already enforced)
  present_colors <- setNames(
    rep_len(palette_vals, length(unique_groups))[seq_along(unique_groups)],
    as.character(unique_groups)
  )

  # plot
  p <- ggplot(fc_df, aes(y = .data[[log2fc_col]], x = y_group, fill = y_group)) +
    gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA) +
    ggrepel::geom_text_repel(
      data = label_df,
      aes(label = Gene.Symbol, y = value, x = y_group),
      position = position_nudge(x = -0.25),
      size = 2, min.segment.length = 0, max.overlaps = Inf,
      box.padding = 0.2, point.padding = 0.1, segment.size = 0.2
    ) +
    scale_fill_manual(values = present_colors) +
    coord_flip(clip = "off") +
    theme_bw(base_size = 6) +
    labs(y = log2fc_col, x = NULL,
         title = paste0("Localization Distribution: ", paste(as.character(unique_groups), collapse = " vs "))) +
    theme(panel.grid = element_blank(), legend.position = "none")

  clean_name <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
  out_file <- file.path(out_dir, paste0("rainbow3_", clean_name(log2fc_col), "_", paste(clean_name(as.character(unique_groups)), collapse = "_"), ".pdf"))
  ggsave(out_file, p, width = 8, height = 3, device = cairo_pdf)
}



# --- Step 2. Define annotations to be plotted ==> 3 MAX
# gene_sets <- c("Lysosome","EarlyEndosome","RecyclingEndosome","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis","Nucleus","Lysosome","EarlyEndosome","RecyclingEndosome","NeuroDev","SynapseSVs","Presynaptic", "Postsynaptic","vATPase")
gene_sets3 <- c("Lysosome","OXPHOS", "mtComplexI")
gene_sets3 <- c("SynapseSVs","Svexocytosis", "SVfusion")
gene_sets3 <- c("Endo_iN_curated_Hundley","EarlyEndosome", "RecyclingEndosome")
gene_sets3 <- c("OXPHOS","MitoIMS", "mtComplexI")

# --- Step 3. Define log2FC datasets to be plotted 
# log2fc_cols <- c(
#   "log2FC_ASAH1_Axon.ASAH1_Soma",
#   "log2FC_Ctrl_Axon.Ctrl_Soma",
#   "log2FC_ASAH1_Axon.Ctrl_Axon"
# )


# --- Step  4. Call function for Control Axon vs projections 
plot_annotation_rainbow3(
  df          = diff136_df_annotated,
  gene_sets   = gene_sets3,
  log2fc_col  = "log2FC_ASAH1_Axon.ASAH1_Soma",
  out_dir     = out_dir_diff136_rainbow_3anno,
  reference_fc = "log2FC_ASAH1_Axon.ASAH1_Soma",
  n_top_labels = 5)



# --- Step  5. Call function for KO vs Ctrl axon comparison
plot_annotation_rainbow3(
  df          = diff136_df_annotated,
  gene_sets   = gene_sets3,
  log2fc_col  = "log2FC_ASAH1_Axon.Ctrl_Axon",
  out_dir     = out_dir_diff136_rainbow_3anno,
  reference_fc = "log2FC_ASAH1_Axon.Ctrl_Axon",
  n_top_labels = 5)

```


### ?? RoR rainbowplot (3 annotations) — ASAH1/Ctrl in projections, WC-normalized
```{r}
# --- 1. Add RoR column (if not yet present)
diff136_df_annotated <- diff136_df_annotated %>%
  mutate(RoR_ASAH1_Axon = log2FC_ASAH1_Axon.ASAH1 - log2FC_Ctrl_Axon.Ctrl) %>% 
  mutate(RoR_ASAH1_Soma = log2FC_ASAH1_Soma.ASAH1 - log2FC_Ctrl_Soma.Ctrl)



plot_ror_rainbow3 <- function(
  df,
  gene_sets,                       # length 3; vector of annotation cols OR named list of gene vectors
  out_dir,
  ror_col = "RoR_ASAH1_Axon",      # log2FC_ASAH1_Axon.ASAH1 - log2FC_Ctrl_Axon.Ctrl
  n_top_labels = 5,
  color_map = c("Label1" = "grey70", "Label2" = "firebrick", "Label3" = "dodgerblue3"),
  xlim_override = NULL
) {
  # 0) Ensure RoR column exists
  if (!ror_col %in% names(df)) {
    stopifnot(all(c("log2FC_ASAH1_Axon.ASAH1","log2FC_Ctrl_Axon.Ctrl") %in% names(df)))
    df <- df %>%
      dplyr::mutate(RoR_ASAH1_Axon = log2FC_ASAH1_Axon.ASAH1 - log2FC_Ctrl_Axon.Ctrl)
  }

  # 1) Normalize gene_sets -> tibble(Gene.Symbol, y_group)
  if (is.character(gene_sets) && is.null(names(gene_sets))) {
    if (length(gene_sets) != 3) stop("Provide exactly three annotation columns.")
    gene_sets <- setNames(gene_sets, gene_sets)
    gene_df <- purrr::imap_dfr(gene_sets, \(col_name, grp)
      tibble::tibble(Gene.Symbol = df$Gene.Symbol[df[[col_name]] == 1], y_group = grp)
    )
  } else {
    if (length(gene_sets) != 3) stop("Provide exactly three annotation groups.")
    gene_df <- purrr::imap_dfr(gene_sets, \(genes, grp)
      tibble::tibble(Gene.Symbol = genes, y_group = grp)
    )
  }
  gene_df <- dplyr::distinct(gene_df, Gene.Symbol, y_group)
  group_levels <- unique(gene_df$y_group)
  if (length(group_levels) != 3) stop("After processing, there must be exactly three groups.")

  # 2) One value per gene for the RoR column
  df_gene <- df %>%
    dplyr::group_by(Gene.Symbol) %>%
    dplyr::summarise(!!ror_col := mean(.data[[ror_col]], na.rm = TRUE), .groups = "drop")

  # 3) Join and filter
  fc_df <- df_gene %>%
    dplyr::inner_join(gene_df, by = "Gene.Symbol") %>%
    dplyr::filter(is.finite(.data[[ror_col]])) %>%
    dplyr::mutate(y_group = factor(y_group, levels = group_levels))

  if (nrow(fc_df) == 0) {
    warning("No finite RoR values found for the provided groups."); return(invisible(NULL))
  }

  # 4) Label top/bottom per group (rank by RoR itself)
  label_df <- dplyr::bind_rows(
    fc_df %>% dplyr::group_by(y_group) %>% dplyr::slice_max(order_by = .data[[ror_col]], n = n_top_labels, with_ties = FALSE),
    fc_df %>% dplyr::group_by(y_group) %>% dplyr::slice_min(order_by = .data[[ror_col]], n = n_top_labels, with_ties = FALSE)
  ) %>% dplyr::ungroup() %>%
    dplyr::transmute(Gene.Symbol, y_group, value = .data[[ror_col]])

  # 5) Colors: use first 3 provided
  pal_vals <- unname(color_map)
  if (length(pal_vals) < 3) pal_vals <- c("grey70","firebrick","dodgerblue3")
  present_colors <- setNames(pal_vals[seq_len(3)], group_levels)

  # 6) Axis limits
  xlims <- if (is.null(xlim_override)) range(fc_df[[ror_col]], na.rm = TRUE) else xlim_override

  # 7) Plot (Vb optics)
  p <- ggplot2::ggplot(fc_df, ggplot2::aes(y = .data[[ror_col]], x = y_group, fill = y_group)) +
    gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
    ggplot2::geom_boxplot(width = 0.15, outlier.shape = NA) +
    ggrepel::geom_text_repel(
      data = label_df,
      ggplot2::aes(label = Gene.Symbol, y = value, x = y_group),
      position = ggplot2::position_nudge(x = -0.25),
      size = 2, min.segment.length = 0, max.overlaps = Inf,
      box.padding = 0.2, point.padding = 0.1, segment.size = 0.2
    ) +
    ggplot2::scale_fill_manual(values = present_colors) +
    ggplot2::coord_flip(ylim = xlims, clip = "off") +
    ggplot2::theme_bw(base_size = 6) +
    ggplot2::labs(
      y = "RoR (ASAH1 Axon − Ctrl Axon, WC-normalized)",
      x = NULL,
      title = paste0("RoR Localization: ", paste(as.character(group_levels), collapse = " vs "))
    ) +
    ggplot2::theme(panel.grid = element_blank(), legend.position = "none")

  # 8) Save
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  clean_name <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
  out_file <- file.path(out_dir, paste0(
    "rainbow3_RoR_ASAH1_over_Ctrl_projections_",
    paste(clean_name(as.character(group_levels)), collapse = "_"),
    ".pdf"
  ))
  grDevices::cairo_pdf(out_file, width = 8, height = 3); print(p); grDevices::dev.off()
}




# Ensure the two log2FC columns exist; the function will create RoR if missing
plot_ror_rainbow3(
  df        = diff136_df_annotated,
  gene_sets = c("MitoIMS","OXPHOS","mtComplexI"),  
  out_dir   = out_dir_diff136_rainbow_RoR,
  ror_col   = "RoR_ASAH1_Axon",    # ASAH1 Axon vs Ctrl Axon, each normalized to WC in genotype
  n_top_labels = 5
)


# Annotations:
# "Mito","MitoIMS","MitoMatrix","MitoMIM","MitoMOM","OXPHOS","mtComplexI","Golgi","Lysosome"               ,"ER","Cytoplasm","Nucleus","Autophagy","Autophagy_det","SynGO","Endo_iN_curated_Hundley" "SynapseSVs","Presynaptic","Postsynaptic","EndoDetailLoac","NeuroDev","EndoLyso","EarlyEndosome","RecyclingEndosome","SVfusion","Svendocytosis","Svexocytosis","vATPase"     




```



## IVb. Rainbowplot of User-select Annotations for axon--soma localization distribution
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Rainfall plot for highlighting protein localization
# user select input 
# ─────────────────────────────────────────────────────────────────────────────
# --- Step  1. Function for plotting user-defined gene sets for axon-soma localization distribution
#    saves single plots and faceted version of the three subcell comparisons
plot_annotation_distribution_batch <- function(df, gene_sets, log2fc_cols, out_dir) {
  if (is.character(gene_sets) && is.null(names(gene_sets))) {
    gene_sets <- setNames(gene_sets, gene_sets)
    gene_df <- purrr::imap_dfr(gene_sets, function(col, label) {
      genes <- df$Gene.Symbol[df[[col]] == 1]
      tibble(Gene.Symbol = genes, y_group = label)
    })
  } else {
    gene_df <- purrr::imap_dfr(gene_sets, ~ tibble(Gene.Symbol = .x, y_group = .y))
  }

  # Helper
  clean_name <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
  default_colors <- c("grey30", "#bdc9e1", "dodgerblue3", "#74a9cf", "#045a8d", "#7fcdbb", "#e34a33", "#fdcc8a")
  fill_colors <- setNames(rep_len(default_colors, length(gene_sets)), names(gene_sets))

  # Collect all for faceting
  full_plot_df <- purrr::map_dfr(log2fc_cols, function(fc) {
    df %>%
      right_join(gene_df, by = "Gene.Symbol") %>%
      filter(!is.na(y_group), is.finite(.data[[fc]])) %>%
      mutate(
        y_group = factor(y_group, levels = names(gene_sets)),
        Log2FC = fc
      )
  })

  # Get shared x-axis limits
  x_limits <- range(full_plot_df[[log2fc_cols[1]]], na.rm = TRUE)
  for (fc in log2fc_cols[-1]) {
    x_limits <- range(c(x_limits, full_plot_df[[fc]]), na.rm = TRUE)
  }

  # --- Save individual plots
  for (fc in log2fc_cols) {
    plot_df <- full_plot_df %>% filter(Log2FC == fc)

    p <- ggplot(plot_df, aes(y = .data[[fc]], x = y_group, fill = y_group)) +
      gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
      ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
      geom_boxplot(width = 0.15, outlier.shape = NA) +
      scale_fill_manual(values = fill_colors) +
      coord_flip(ylim = x_limits, clip = "off") +
      theme_bw(base_size = 6) +
      labs(y = fc, x = NULL, title = paste("Localization Distribution:", fc)) +
      theme(panel.grid = element_blank(), legend.position = "none")

    fname <- paste0("locdist_", clean_name(fc), "_", paste(clean_name(names(gene_sets)), collapse = "_"), ".pdf")
    ggsave(file.path(out_dir, fname), p, width = 4, height = 4, device = cairo_pdf)
  }

  # --- Faceted version
  facet_df <- full_plot_df %>%
    pivot_longer(cols = all_of(log2fc_cols), names_to = "log2fc_col", values_to = "value") %>%
    filter(!is.na(value))

  p_all <- ggplot(facet_df, aes(y = value, x = y_group, fill = y_group)) +
    gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA) +
    scale_fill_manual(values = fill_colors) +
    coord_flip(ylim = x_limits, clip = "off") +
    facet_wrap(~ log2fc_col, ncol = 1) +
    theme_bw(base_size = 6) +
    labs(y = "log2FC", x = NULL, title = "Localization Distribution Across Comparisons") +
    theme(panel.grid = element_blank(), legend.position = "none")

  fname_all <- paste0("facet_locdist_", paste(clean_name(log2fc_cols), collapse = "_"), "_", paste(clean_name(names(gene_sets)), collapse = "_"), ".pdf")
  ggsave(file.path(out_dir, fname_all), p_all, width = 4, height = 10, device = cairo_pdf)
}



# --- Step 2. Define annotations to be plotted 
#gene_sets <- c("SynGO", "Presynaptic", "Postsynaptic")
#gene_sets <- c("EndoLyso","EarlyEndosome","RecyclingEndosome")
#gene_sets <- c("OXPHOS","Lysosome","Endo_iN_curated_Hundley", "EarlyEndosome","RecyclingEndosome")
#gene_sets <- c("OXPHOS","Golgi","ER")
#gene_sets <- c("Nucleus","Lysosome","NeuroDev","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis")
gene_sets <- c("Lysosome","EarlyEndosome","RecyclingEndosome","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis")
gene_sets <- c("Nucleus","Lysosome","EarlyEndosome","RecyclingEndosome","NeuroDev","SynapseSVs","Presynaptic", "Postsynaptic")
gene_sets <- c("Mito","MitoIMS","MitoMatrix","MitoMIM","MitoMOM","OXPHOS","mtComplexI")

# --- Step 3. Define log2FC datasets to be plotted 
log2fc_cols <- c(
  "log2FC_ASAH1_Axon.ASAH1_Soma",
  "log2FC_Ctrl_Axon.Ctrl_Soma",
  "log2FC_ASAH1_Axon.Ctrl_Axon"
)

# --- Step 4. Call function and define output directory
plot_annotation_distribution_batch(
  df          = diff136_df_annotated,
  gene_sets   = gene_sets,
  log2fc_cols = log2fc_cols,
  out_dir     = out_dir_diff136_rainbow
)


# Annotations:
# "Mito","MitoIMS","MitoMatrix","MitoMIM","MitoMOM","OXPHOS","mtComplexI","Golgi","Lysosome"               ,"ER","Cytoplasm","Nucleus","Autophagy","Autophagy_det","SynGO","Endo_iN_curated_Hundley" "SynapseSVs","Presynaptic","Postsynaptic","EndoDetailLoac","NeuroDev","EndoLyso","EarlyEndosome","RecyclingEndosome","SVfusion","Svendocytosis","Svexocytosis"     


```



## IVc. Rainbowplot of User-select Annotations for axon--soma localization distribution
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Rainfall plot for highlighting protein localization
# user select input 
# ─────────────────────────────────────────────────────────────────────────────
# --- Step  1. Function for plotting user-defined gene sets for axon-soma localization distribution
#    saves single plots and faceted version of the three subcell comparisons
plot_annotation_distribution_batch <- function(df, gene_sets, log2fc_cols, out_dir, x_lim_override = NULL) {
  if (is.character(gene_sets) && is.null(names(gene_sets))) {
    gene_sets <- setNames(gene_sets, gene_sets)
    gene_df <- purrr::imap_dfr(gene_sets, function(col, label) {
      genes <- df$Gene.Symbol[df[[col]] == 1]
      tibble(Gene.Symbol = genes, y_group = label)
    })
  } else {
    gene_df <- purrr::imap_dfr(gene_sets, ~ tibble(Gene.Symbol = .x, y_group = .y))
  }

  # Helper
  clean_name <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
  default_colors <- c("grey30", "#bdc9e1", "dodgerblue3", "#74a9cf", "#045a8d", "#7fcdbb", "#e34a33", "#fdcc8a")
  fill_colors <- setNames(rep_len(default_colors, length(gene_sets)), names(gene_sets))

  # Collect all for faceting
  full_plot_df <- purrr::map_dfr(log2fc_cols, function(fc) {
    df %>%
      right_join(gene_df, by = "Gene.Symbol") %>%
      filter(!is.na(y_group), is.finite(.data[[fc]])) %>%
      mutate(
        y_group = factor(y_group, levels = names(gene_sets)),
        Log2FC = fc
      )
  })

  # Determine x-axis limits
  if (!is.null(x_lim_override)) {
    x_limits <- x_lim_override
  } else {
    x_limits <- range(full_plot_df[[log2fc_cols[1]]], na.rm = TRUE)
    for (fc in log2fc_cols[-1]) {
      x_limits <- range(c(x_limits, full_plot_df[[fc]]), na.rm = TRUE)
    }
  }

  # Individual plots
  for (fc in log2fc_cols) {
    plot_df <- full_plot_df %>% filter(Log2FC == fc)

    p <- ggplot(plot_df, aes(y = .data[[fc]], x = y_group, fill = y_group)) +
      gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
      ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
      geom_boxplot(width = 0.15, outlier.shape = NA) +
      scale_fill_manual(values = fill_colors) +
      coord_flip(ylim = x_limits, clip = "off") +
      theme_bw(base_size = 6) +
      labs(y = fc, x = NULL, title = paste("Localization Distribution:", fc)) +
      theme(panel.grid = element_blank(), legend.position = "none")

    fname <- paste0("locdist_", clean_name(fc), "_", paste(clean_name(names(gene_sets)), collapse = "_"), ".pdf")
    ggsave(file.path(out_dir, fname), p, width = 4, height = 4, device = cairo_pdf)
  }

  # Faceted version
  facet_df <- full_plot_df %>%
    pivot_longer(cols = all_of(log2fc_cols), names_to = "log2fc_col", values_to = "value") %>%
    filter(!is.na(value))

  p_all <- ggplot(facet_df, aes(y = value, x = y_group, fill = y_group)) +
    gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA) +
    scale_fill_manual(values = fill_colors) +
    coord_flip(ylim = x_limits, clip = "off") +
    facet_wrap(~ log2fc_col, ncol = 1) +
    theme_bw(base_size = 6) +
    labs(y = "log2FC", x = NULL, title = "Localization Distribution Across Comparisons") +
    theme(panel.grid = element_blank(), legend.position = "none")

  #fname_all <- paste0("facet_locdist_", paste(clean_name(log2fc_cols), collapse = "_"), "_", paste(clean_name(names(gene_sets)), collapse = "_"), ".pdf")
  fname_all <- paste0("facet_locdist_", length(log2fc_cols), "comparisons_", length(gene_sets), "annotations.pdf")
  ggsave(file.path(out_dir, fname_all), p_all, width = 4, height = 10, device = cairo_pdf)
}



# --- Step 2. Define annotations to be plotted 
#gene_sets <- c("SynGO", "Presynaptic", "Postsynaptic")
#gene_sets <- c("EndoLyso","EarlyEndosome","RecyclingEndosome")
#gene_sets <- c("OXPHOS","Lysosome","Endo_iN_curated_Hundley", "EarlyEndosome","RecyclingEndosome")
#gene_sets <- c("OXPHOS","Golgi","ER")
#gene_sets <- c("Nucleus","Lysosome","NeuroDev","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis")
gene_sets <- c("Lysosome","EarlyEndosome","RecyclingEndosome","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis")
gene_sets <- c("Nucleus","Lysosome","EarlyEndosome","RecyclingEndosome","NeuroDev","SynapseSVs","Presynaptic", "Postsynaptic")
gene_sets <- c("Mito","MitoIMS","MitoMatrix","MitoMIM","MitoMOM","OXPHOS","mtComplexI")

# --- Step 3. Define log2FC datasets to be plotted 
log2fc_cols <- c(
  "log2FC_ASAH1_Axon.ASAH1_Soma",
  "log2FC_Ctrl_Axon.Ctrl_Soma",
  "log2FC_ASAH1_Axon.Ctrl_Axon"
)


# --- Step 4. Call function and define output directory
plot_annotation_distribution_batch(
  df          = diff136_df_annotated,
  gene_sets   = gene_sets,
  log2fc_cols = log2fc_cols,
  out_dir     = out_dir_diff136_rainbow,
  x_lim_override = c(-2, 2) #do set manuarl x-lims if necessary
)


```




## IVc. RoR-based Rainbowplot of user-selected annotations 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Rainfall plot for highlighting protein localization
# RoR-based Rainfall Plot of Protein Localization
# user select input 
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Add RoR column (if not yet present)
diff136_df_annotated <- diff136_df_annotated %>%
  mutate(RoR_ASAH1_Axon = log2FC_ASAH1_Axon.ASAH1 - log2FC_Ctrl_Axon.Ctrl) %>% 
  mutate(RoR_ASAH1_Soma = log2FC_ASAH1_Soma.ASAH1 - log2FC_Ctrl_Soma.Ctrl)

# --- 2. Function: rainplot for RoR values by gene set
plot_ror_annotation_distribution <- function(df, gene_sets, ror_cols, out_dir, xlim = NULL) {
  # --- 2.1. Prepare gene set annotations
  if (is.character(gene_sets) && is.null(names(gene_sets))) {
    gene_sets <- setNames(gene_sets, gene_sets)
    gene_df <- purrr::imap_dfr(gene_sets, function(col, label) {
      genes <- df$Gene.Symbol[df[[col]] == 1]
      tibble(Gene.Symbol = genes, y_group = label)
    })
  } else {
    gene_df <- purrr::imap_dfr(gene_sets, ~ tibble(Gene.Symbol = .x, y_group = .y))
  }

  clean_name <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
  default_colors <- c("grey30", "#bdc9e1", "dodgerblue3", "#74a9cf", "#045a8d", "#7fcdbb", "#e34a33", "#fdcc8a")
  fill_colors <- setNames(rep_len(default_colors, length(gene_sets)), names(gene_sets))

  # --- 2.2. Build plotting data
  full_plot_df <- purrr::map_dfr(ror_cols, function(fc) {
    df %>%
      right_join(gene_df, by = "Gene.Symbol") %>%
      filter(!is.na(y_group), is.finite(.data[[fc]])) %>%
      mutate(
        y_group = factor(y_group, levels = names(gene_sets)),
        Log2FC = fc
      )
  })

  # --- 2.3. Compute or apply x-axis limits
  facet_df <- full_plot_df %>%
    pivot_longer(cols = all_of(ror_cols), names_to = "log2fc_col", values_to = "value") %>%
    filter(!is.na(value))

  x_limits <- if (!is.null(xlim)) xlim else range(facet_df$value, na.rm = TRUE)

  # --- 2.4. Individual plots
  for (fc in ror_cols) {
    plot_df <- facet_df %>% filter(log2fc_col == fc)

    p <- ggplot(plot_df, aes(y = value, x = y_group, fill = y_group)) +
      gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
      ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
      geom_boxplot(width = 0.15, outlier.shape = NA) +
      scale_fill_manual(values = fill_colors) +
      coord_flip(ylim = x_limits, clip = "off") +
      theme_bw(base_size = 6) +
      labs(y = "RoR (ASAH1 vs Ctrl)", x = NULL, title = paste("RoR Localization:", fc)) +
      theme(panel.grid = element_blank(), legend.position = "none")

    ggsave(
      file.path(out_dir, paste0("ror_locdist_", clean_name(fc), "_", paste(clean_name(names(gene_sets)), collapse = "_"), ".pdf")),
      p, width = 4, height = 2, device = cairo_pdf
    )
  }

  # --- 2.5. Faceted plot
  p_all <- ggplot(facet_df, aes(y = value, x = y_group, fill = y_group)) +
    gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
    geom_boxplot(width = 0.15, outlier.shape = NA) +
    scale_fill_manual(values = fill_colors) +
    coord_flip(ylim = x_limits, clip = "on") +
    facet_wrap(~ log2fc_col, ncol = 1) +
    theme_bw(base_size = 6) +
    labs(y = "RoR (log2FC ΔAxon/WC)", x = NULL, title = "RoR Localization Across Comparisons") +
    theme(panel.grid = element_blank(), legend.position = "none")

  ggsave(
    file.path(out_dir, paste0("facet_ror_locdist_", paste(clean_name(ror_cols), collapse = "_"), "_", paste(clean_name(names(gene_sets)), collapse = "_"), ".pdf")),
    p_all, width = 6, height = 5, device = cairo_pdf
  )
}


# --- 3. Define input annotations 
# gene_sets <- c("SynGO", "Presynaptic", "Postsynaptic")
# gene_sets <- c("EndoLyso","EarlyEndosome","RecyclingEndosome")
# gene_sets <- c("OXPHOS","Lysosome","Endo_iN_curated_Hundley", "EarlyEndosome","RecyclingEndosome")
# gene_sets <- c("OXPHOS","Golgi","ER")
# gene_sets <- c("Nucleus","Lysosome","NeuroDev","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis")
# gene_sets <- c("Lysosome","EarlyEndosome","RecyclingEndosome","SynapseSVs","SVfusion", "Svendocytosis","Svexocytosis")
# gene_sets <- c("Nucleus","Lysosome","EarlyEndosome","RecyclingEndosome","NeuroDev","SynapseSVs","Presynaptic", "Postsynaptic")
gene_sets <- c("Lysosome", "EarlyEndosome", "RecyclingEndosome", "SynapseSVs", "SVfusion", "Svendocytosis", "Svexocytosis")
gene_sets <- c("Mito","MitoIMS","MitoMatrix","MitoMIM","MitoMOM","OXPHOS","mtComplexI")


# --- 4. Define dataframes to plot (either Axon, Soma or both)
# single comparison 
# ror_cols <- c("RoR_ASAH1_Axon")
# ror_cols <- c("RoR_ASAH1_Soma")

# double comparison
ror_cols <- c("RoR_ASAH1_Axon", "RoR_ASAH1_Soma")


# --- 5. Run function
plot_ror_annotation_distribution(
  df        = diff136_df_annotated,
  gene_sets = gene_sets,
  ror_cols  = ror_cols,
  out_dir   = out_dir_diff136_rainbow_RoR,
  xlim      = c(-3, 3) #do set manuarl x-lims if necessary
)



# Annotations:
# "Mito","MitoIMS","MitoMatrix","MitoMIM","MitoMOM","OXPHOS","mtComplexI","Golgi","Lysosome"               ,"ER","Cytoplasm","Nucleus","Autophagy","Autophagy_det","SynGO","Endo_iN_curated_Hundley" "SynapseSVs","Presynaptic","Postsynaptic","EndoDetailLoac","NeuroDev","EndoLyso","EarlyEndosome","RecyclingEndosome","SVfusion","Svendocytosis","Svexocytosis","vATPase"     


```




