---
title: "HeLaLysoIpASAH1"
output: html_document
date: "2024-12-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load packages
```{r}
# --- Core tidyverse packages
library(plyr)
library(stringr)
library(tidyverse)
library(dplyr)
library(tibble)
library(purrr)
library(rlang)
library(readr)


# --- Grammar of graphics & plotting extensions
library(ggrepel)
library(scales)
library(NatParksPalettes)
library(viridis)
library(RColorBrewer)
library(ggVennDiagram)
library(pheatmap)
library(gplots)
library(png)
library(plotly)
library(patchwork)
library(cowplot)
library(gridExtra)
library(grid)

# --- Data handling
library(data.table)

# --- Specialized / utility packages
library(tidyplots)
library(gptstudio)
library(rlang)
library(devtools)
library(Cairo)


# --- GO Term analysis ---
#BiocManager::install("topGO")
library(topGO)
#BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)
#BiocManager::install("clusterProfiler")
library(clusterProfiler)


library(ggplot2)
library(ggdist)
library(gghalves)
library(ggrepel)
library(dplyr)
library(tibble)


# Optional install commands
# devtools::install_github("jbengler/tidyplots")
# install.packages("magick")
```

# output dirs overview:
```{r}
out_dir_lysoip_dataframes <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/dataframes"
dir.create(out_dir_lysoip_dataframes, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_heatmap  <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/heatmaps"
dir.create(out_dir_lysoip_heatmap, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_corrplots <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/Organelle_CrossCorr"
dir.create(out_dir_lysoip_corrplots, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_organelle_dist <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/Organelle_Distribution"
dir.create(out_dir_lysoip_organelle_dist, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_organelle_rainfall <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/Rainfall"
dir.create(out_dir_lysoip_organelle_rainfall, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_venn <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/venn"
dir.create(out_dir_lysoip_venn, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_venn_coenrichment  <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/venn/coenrichment"
dir.create(out_dir_lysoip_venn_coenrichment, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_autophagy <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/autophagy"
dir.create(out_dir_lysoip_autophagy, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_GO <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/GO"
dir.create(out_dir_lysoip_GO, recursive = TRUE, showWarnings = FALSE)

out_dir_lysoip_GOcoenrichment <- "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/GO/coenrichment"
dir.create(out_dir_lysoip_GOcoenrichment, recursive = TRUE, showWarnings = FALSE)



```

# I. Background HeLa LysoIP experiments
Dataset contains data from isolated lysosomes from HeLa cell lines
Genotypes: HeLa untagged, HeLa TMEM192-3xHA Ctrl, HeLa TMEM192-3xHA ASAH1-/- #6
Samples (all in triplicates): untagged, Ctrl and ASAH1 
9plex_TMTpro & run on Thermo Orbitrap Ascent w/FAIMS-Pro, hrMS2
data was created with MSstats, has p/q.values and log2FC for samples 
k-means clustering was performed on data, is in column "cluster"



# II. Data import, add annotations
```{r}
# read in df
helalysoip_df <- read.csv(
  "/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/Eval/datasets/20241130_HeLaLysoIP_ASAH1_All_results_stats_localizationAdded.csv",
  header = TRUE
)

# Transform p/q-values to -log10, and filter non-finite log2FCs
log2fc_cols_hela <- c(
"log2FC_CtrlLysoIP.untagged",
"log2FC_ASAH1LysoIP.untagged",
"log2FC_ASAH1LysoIP.CtrlLysoIP"
)

helalysoip_df <- helalysoip_df %>%
  mutate(across(contains("p.val"), ~ ifelse(. > 0, -log10(.), NA), .names = "neg_log10_{.col}"),
         across(contains("q.val"), ~ ifelse(. > 0, -log10(.), NA), .names = "neg_log10_{.col}")) %>%
  filter(if_all(all_of(log2fc_cols_hela), ~ is.finite(.)))


# 1. Define lysohydrolase list
lysohydrolases <- c(
  "ASAH1", "GBA", "ARSA", "ARSB", "GALNS", "IDUA", "GLA", "NEU1", "MAN2B1",
  "GLB1", "HEXA", "GNS", "HEXB", "CTSA", "CTSZ", "TPP1", "NAGLU", "CTSL",
  "CTSB", "NAGA", "LGMN"
)

# 2. Process subcellular annotation
cols_to_read <- c(1,2,3,4,5,6,7,9,10,12,15,16,17,18,19,20,21,22,23,25,26,27,28,29,30,31,32,33)
subcell_df <- read.csv("/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/neuroLSDgit/proteome/SubCellAnnotation.csv", stringsAsFactors = FALSE)[, cols_to_read]
subcell_df[] <- lapply(subcell_df, as.character)

long_subcell_df <- subcell_df %>%
  pivot_longer(cols = everything(), names_to = "Localization", values_to = "Genes") %>%
  filter(!is.na(Genes) & Genes != "") %>%
  mutate(Genes = trimws(Genes)) %>%
  distinct()

binary_matrix_helalysoip <- long_subcell_df %>%
  mutate(value = TRUE) %>%
  pivot_wider(names_from = Localization, values_from = value, values_fill = list(value = FALSE))

# 3. Annotate helalysoip_df directly
helalysoip_annotated_df <- helalysoip_df %>%
  mutate(LysoHydrolase = Gene.Symbol %in% lysohydrolases) %>%
  left_join(binary_matrix_helalysoip, by = c("Gene.Symbol" = "Genes")) %>%
  mutate(across(where(is.logical) | where(is.numeric), ~ replace_na(., FALSE)))

# save as csv file 
write.csv(helalysoip_annotated_df, file = file.path(out_dir_lysoip_dataframes, "helalysoip_annotated.csv"), row.names = FALSE)
```

# III. Heatmaps of dataframe w/ clusters of interst 
```{r}
# z-score heatmap (rows = gene, columns = sample)

# --- 1. Expression and clustering setup
expr_cols_lysoip <- c(
  "untagged_1", "untagged_2", "untagged_3",
  "CtrlLysoIP_1", "CtrlLysoIP_2", "CtrlLysoIP_3",
  "ASAH1LysoIP_1", "ASAH1LysoIP_2", "ASAH1LysoIP_3"
)

mat_helalysoip <- helalysoip_annotated_df %>%
  filter(!is.na(Gene.Symbol)) %>%
  mutate(Gene.Symbol = make.unique(as.character(Gene.Symbol))) %>%
  dplyr::select(Gene.Symbol, cluster, all_of(expr_cols_lysoip))

row_cluster <- mat_helalysoip$cluster
rownames(mat_helalysoip) <- mat_helalysoip$Gene.Symbol
mat_helalysoip <- mat_helalysoip[, expr_cols_lysoip]
mat_helalysoip <- mat_helalysoip[order(row_cluster), , drop = FALSE]

# --- 2. Row (cluster) annotation
cluster_levels_helalysoip <- sort(unique(na.omit(as.integer(row_cluster))))
row_annotation_helalysoip <- data.frame(
  cluster = factor(row_cluster[order(row_cluster)], levels = cluster_levels_helalysoip)
)
rownames(row_annotation_helalysoip) <- rownames(mat_helalysoip)

cluster_palette_helalysoip <- setNames(
  RColorBrewer::brewer.pal(length(cluster_levels_helalysoip), "Set2")[1:length(cluster_levels_helalysoip)],
  as.character(cluster_levels_helalysoip)
)

ann_colors_helalysoip <- list(
  Genotype = c(Ctrl = "grey40", ASAH1 = "dodgerblue3"),
  cluster = cluster_palette_helalysoip
)

# --- 3. Column (sample) annotation
annotation_col_helalysoip <- tibble(Sample = expr_cols_lysoip) %>%
  mutate(
    Genotype = if_else(str_detect(Sample, "^Ctrl"), "Ctrl", "ASAH1"),
    Fraction = "LysoIP"
  ) %>%
  column_to_rownames("Sample")

ann_colors_helalysoip$Fraction <- c(
  `Whole-cell` = "#bdbdbd",
  LysoIP = "#9ecae1"  # or another distinct color
)

# --- 4. Save heatmap and matrix
pdf_file_helalysoip <- file.path(out_dir_lysoip_heatmap, "helalysoip_heatmap.pdf")

cairo_pdf(pdf_file_helalysoip, width = 3, height = 5)
pheatmap(
  mat_helalysoip,
  scale              = "row",
  color              = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col     = annotation_col_helalysoip,
  annotation_row     = row_annotation_helalysoip,
  annotation_colors  = ann_colors_helalysoip,
  breaks             = seq(-4, 4, length.out = 101),
  cluster_rows       = FALSE,
  cluster_cols       = TRUE,
  show_rownames      = FALSE,
  show_colnames      = TRUE,
  main               = "HeLa LysoIP: Ctrl vs ASAH1",
  fontsize           = 6,
  border_color       = NA
)
dev.off()

# Save matrix with clusters
mat_helalysoip_out <- cbind(cluster = row_annotation_helalysoip$cluster, mat_helalysoip)
write.csv(mat_helalysoip_out,
          file = file.path(out_dir_lysoip_heatmap, "helalysoip_heatmap_data.csv"),
          row.names = TRUE)


```



### IIIa. Scaled intensity Violinplot based on heatmap clusters 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Violin Plot of scaled intensities of mat_helalysoip_out: Group Means by Cluster
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1.  Step: Row-wise Scaling (Z-score)
# Scale intensities row-wise (z-score across all replicates)
mat_scaled <- mat_helalysoip_out %>%
  dplyr::select(untagged_1:ASAH1LysoIP_3) %>%
  as.matrix() %>%
  t() %>%
  scale(center = TRUE, scale = TRUE) %>%
  t() %>%
  as.data.frame()

# --- Add cluster info back ---
mat_scaled$cluster <- mat_helalysoip_out$cluster



# --- Step 2. Rebuild long-format dataframe for plotting
# Compute group-wise scaled means
violin_df_lysoip_mean_scaled <- mat_scaled %>%
  mutate(
    untagged     = rowMeans(dplyr::select(., c("untagged_1", "untagged_2", "untagged_3")), na.rm = TRUE),
    CtrlLysoIP   = rowMeans(dplyr::select(., c("CtrlLysoIP_1", "CtrlLysoIP_2", "CtrlLysoIP_3")), na.rm = TRUE),
    ASAH1LysoIP  = rowMeans(dplyr::select(., c("ASAH1LysoIP_1", "ASAH1LysoIP_2", "ASAH1LysoIP_3")), na.rm = TRUE)
  ) %>%
  dplyr::select(cluster, untagged, CtrlLysoIP, ASAH1LysoIP) %>%
  pivot_longer(cols = c("untagged", "CtrlLysoIP", "ASAH1LysoIP"),
               names_to = "Sample", values_to = "scaled_intensity") %>%
  filter(!is.na(cluster) & cluster != 0 & is.finite(scaled_intensity)) %>%
  mutate(cluster = factor(cluster),
         Sample = factor(Sample, levels = c("untagged", "CtrlLysoIP", "ASAH1LysoIP")))


# --- Step 3. Plot scaled violin plot
# Scaled violin plot
violin_scaled_plot <- ggplot(violin_df_lysoip_mean_scaled, aes(x = Sample, y = scaled_intensity, fill = Sample)) +
  geom_jitter(color = "red", size = 0.5, alpha = 0.2, width = 0.2) +
  geom_violin(scale = "width", trim = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_manual(values = c(
    "untagged"    = "#bdc9e1",
    "CtrlLysoIP"  = "#74a9cf",
    "ASAH1LysoIP" = "dodgerblue3"
  )) +
  theme_bw(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none") +
  labs(title = "LysoIP: Scaled Intensities by Sample per Cluster", x = NULL, y = "Scaled Intensity")



# --- Step 4. Save to PDF
ggsave(file.path(out_dir_lysoip_heatmap, "LysoIP_clusterViolinPlot_scaled.pdf"),
       violin_scaled_plot, width = 6, height = 6, device = cairo_pdf)



```

### IIIb. Cluster 2 (lyso enriched) with hydrolases  
```{r}
# --- Target list to overlay
lysohydrolases <- c(
  "ASAH1","GBA","ARSA","ARSB","GALNS","IDUA","GLA","NEU1","MAN2B1",
  "GLB1","HEXA","GNS","HEXB","CTSA","CTSZ","TPP1","NAGLU","CTSL",
  "CTSB","NAGA","LGMN"
)

# --- Step 1. Row-wise scaling (unchanged) + keep gene IDs
mat_scaled <- mat_helalysoip_out %>%
  dplyr::select(untagged_1:ASAH1LysoIP_3) %>%
  as.matrix() %>%
  t() %>%
  scale(center = TRUE, scale = TRUE) %>%
  t() %>%
  as.data.frame()

mat_scaled$cluster <- mat_helalysoip_out$cluster
mat_scaled$Gene.Symbol <- rownames(mat_helalysoip_out)

# --- Step 2. Build long df for CLUSTER 2 ONLY and flag lysohydrolases
violin_df_lysoip_mean_scaled_c2 <- mat_scaled %>%
  filter(cluster == 2) %>%
  mutate(
    untagged     = rowMeans(dplyr::select(., c("untagged_1", "untagged_2", "untagged_3")), na.rm = TRUE),
    CtrlLysoIP   = rowMeans(dplyr::select(., c("CtrlLysoIP_1", "CtrlLysoIP_2", "CtrlLysoIP_3")), na.rm = TRUE),
    ASAH1LysoIP  = rowMeans(dplyr::select(., c("ASAH1LysoIP_1", "ASAH1LysoIP_2", "ASAH1LysoIP_3")), na.rm = TRUE)
  ) %>%
  dplyr::select(Gene.Symbol, cluster, untagged, CtrlLysoIP, ASAH1LysoIP) %>%
  pivot_longer(cols = c("untagged", "CtrlLysoIP", "ASAH1LysoIP"),
               names_to = "Sample", values_to = "scaled_intensity") %>%
  filter(is.finite(scaled_intensity)) %>%
  mutate(
    cluster = factor(cluster),
    Sample  = factor(Sample, levels = c("untagged", "CtrlLysoIP", "ASAH1LysoIP")),
    is_lysohydrolase = Gene.Symbol %in% lysohydrolases
  )

# --- Step 3. Plot: cluster 2 only, overlay lysohydrolases
violin_scaled_plot_c2_hydro <- ggplot(violin_df_lysoip_mean_scaled_c2, aes(x = Sample, y = scaled_intensity)) +
  geom_violin(aes(fill = Sample), scale = "width", trim = FALSE) +
  geom_jitter(data = dplyr::filter(violin_df_lysoip_mean_scaled_c2, !is_lysohydrolase), width = 0.2, height = 0, size =1, alpha = 0.15, color = "red") +
  geom_point(data = dplyr::filter(violin_df_lysoip_mean_scaled_c2, is_lysohydrolase), aes(fill = "red"), shape = 21, size = 3, color = "black", stroke = 0.25) +
  # ggrepel::geom_text_repel(
  #   data = dplyr::filter(violin_df_lysoip_mean_scaled_c2, is_lysohydrolase),
  #   aes(label = Gene.Symbol),
  #   size = 2, max.overlaps = Inf, box.padding = 0.2, point.padding = 0.1, segment.size = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_manual(values = c(
    "untagged"    = "#bdc9e1",
    "CtrlLysoIP"  = "#74a9cf",
    "ASAH1LysoIP" = "dodgerblue3"
  )) +
  theme_bw(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  ) +
  labs(title = "LysoIP (Cluster 2): Scaled Intensities with Lysohydrolase Overlay",
       x = NULL, y = "Scaled Intensity")

# --- Step 4. Save
ggsave(file.path(out_dir_lysoip_heatmap, "LysoIP_cluster2_Violin_scaled_lysohydrolaseOverlay.pdf"), violin_scaled_plot_c2_hydro, width = 3, height = 5, device = cairo_pdf)
```


# IV. Violins of clusters of interest 
### IVa. Violin Plot of Cluster Abundance
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Violin Plot of Cluster Abundance
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Define sample columns
sample_cols_lysoip <- c(
"log2FC_CtrlLysoIP.untagged",
"log2FC_ASAH1LysoIP.untagged",
"log2FC_ASAH1LysoIP.CtrlLysoIP" 
)

# --- 2. Reshape and clean
violin_df_lysoip <- helalysoip_df %>%
  filter(!is.na(cluster) & cluster != 0) %>%
  pivot_longer(cols = all_of(sample_cols_lysoip), names_to = "Sample", values_to = "log2FC") %>%
  filter(!is.na(log2FC)) %>%
  mutate(cluster = factor(cluster),
         Sample = factor(Sample, levels = sample_cols_lysoip))  # enforce consistent order


# --- 3. Define plot
lysoip_clusterPlot <- ggplot(violin_df_lysoip, aes(x = Sample, y = log2FC, fill = Sample)) +
  geom_jitter(aes(color = "red"), size = 0.5, alpha = 0.2, width = 0.2, height = 0) +
  geom_violin(scale = "width", trim = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_manual(values = c(
    "log2FC_CtrlLysoIP.untagged"     = "#bdc9e1",
    "log2FC_ASAH1LysoIP.untagged"    = "#74a9cf",
    "log2FC_ASAH1LysoIP.CtrlLysoIP"  = "dodgerblue3"
  )) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none") +
  labs(title = "LysoIP: log2FC by Sample per Cluster", x = NULL, y = "log2FC")

# --- 4. Save to PDF
ggsave(file.path(out_dir_lysoip_heatmap, "LysoIP_clusterViolinPlot_log2FC.pdf"),
       lysoip_clusterPlot, width = 6, height = 6, device = cairo_pdf)

```


### IVb. Violin Plot of Ctrl vs ASAH1-/-
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Violin Plot of detailed endo annotations in ctrl vs ASAH1-/-
# ─────────────────────────────────────────────────────────────────────────────

## --- 1. Define organelle labels and order
organelles <- c("Lysosome","Endo_iN_curated_Hundley","Golgi","ER","Mito")
org_mapping <- tibble(
  Organelle = factor(organelles, levels = organelles),
  Org_Label = c("Lyso","Endo","Golgi","ER","Mito"),
  Org_Number = 1:5
)

# --- 2. Subset data for Ctrl and ASAH1 comparisons
df_det_ctrl <- helalysoip_annotated_df %>%
  filter(if_any(all_of(organelles), ~ .x)) %>%
  dplyr::select(Gene.Symbol, all_of(organelles), log2FC_CtrlLysoIP.untagged) %>%
  pivot_longer(cols = all_of(organelles), names_to = "Organelle", values_to = "flag") %>%
  filter(flag) %>%
  mutate(Sample = "Ctrl vs Untagged", log2FC = log2FC_CtrlLysoIP.untagged)

df_det_asah1 <- helalysoip_annotated_df %>%
  filter(if_any(all_of(organelles), ~ .x)) %>%
  dplyr::select(Gene.Symbol, all_of(organelles), log2FC_ASAH1LysoIP.untagged) %>%
  pivot_longer(cols = all_of(organelles), names_to = "Organelle", values_to = "flag") %>%
  filter(flag) %>%
  mutate(Sample = "ASAH1 vs Untagged", log2FC = log2FC_ASAH1LysoIP.untagged)

# --- 3. Combine and annotate
violin_det_df <- bind_rows(df_det_ctrl, df_det_asah1) %>%
  mutate(Sample = factor(Sample, levels = c("Ctrl vs Untagged","ASAH1 vs Untagged")),
         Organelle = factor(Organelle, levels = organelles)) %>%
  left_join(org_mapping, by = "Organelle")

# --- 4. Compute mean log2FC per annotation and sample, save CSV
mean_log2fc_det_df <- violin_det_df %>%
  group_by(Organelle, Org_Label, Sample) %>%
  summarise(mean_log2FC = round(mean(log2FC, na.rm = TRUE), 2), .groups = "drop")
write.csv(mean_log2fc_det_df, file.path(out_dir_lysoip_organelle_dist, "LysoIP_mean_log2FC.csv"), row.names = FALSE)

# --- 5. Plot
p_violin_det <- ggplot(violin_det_df, aes(x = interaction(Organelle, Sample), y = log2FC, fill = Sample)) +
  geom_violin(scale = "width", trim = FALSE, width = 0.9) +
  geom_jitter(width = 0.15, height = 0, size = 0.3, alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_fill_manual(values = c("Ctrl vs Untagged" = "grey70", "ASAH1 vs Untagged" = "dodgerblue3")) +
  theme_bw(base_size = 6) +
  theme(panel.grid = element_blank(), axis.title.x = element_blank(), axis.text.x = element_text(angle = 90), legend.position = "none") +
  labs(y = expression(log[2]*"FC"), title = "Organelle Enrichment HeLa Ctrl vs ASAH1-/-")

# --- 6. Save
ggsave(file.path(out_dir_lysoip_organelle_dist, "LysoIP_OrganelleEnrichment_Violin.pdf"), p_violin_det, width = 3, height = 5, device = cairo_pdf)



```


### IVc. Violin Plot of select annotations in ctrl vs ASAH1-/-
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Violin Plot of select annotations in ctrl vs ASAH1-/-
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Define setup
organelles <- c("Lysosome","EarlyEndosome","Endo_iN_curated_Hundley","Golgi","RecyclingEndosome","ER","Mito")

org_mapping_detailed <- tibble(
  Organelle = factor(organelles, levels = organelles),
  Org_Label = c("Lysosome","EarlyEndosome","Endo_iN_curated_Hundley","Golgi","RecyclingEndosome","ER","Mito"),
  Org_Number = 1:7
)

org_labels <- org_mapping_detailed$Org_Label
org_order  <- org_mapping_detailed$Org_Number

# --- 2. Create data subset for both comparisons
df_ctrl <- helalysoip_annotated_df %>%
  filter(if_any(all_of(organelles), ~ .x)) %>%
  dplyr::select(Gene.Symbol, all_of(organelles), log2FC_CtrlLysoIP.untagged) %>%
  pivot_longer(cols = all_of(organelles), names_to = "Organelle", values_to = "flag") %>%
  filter(flag) %>%
  mutate(Sample = "Ctrl vs Untagged", log2FC = log2FC_CtrlLysoIP.untagged)

df_asah1 <- helalysoip_annotated_df %>%
  filter(if_any(all_of(organelles), ~ .x)) %>%
  dplyr::select(Gene.Symbol, all_of(organelles), log2FC_ASAH1LysoIP.untagged) %>%
  pivot_longer(cols = all_of(organelles), names_to = "Organelle", values_to = "flag") %>%
  filter(flag) %>%
  mutate(Sample = "ASAH1 vs Untagged", log2FC = log2FC_ASAH1LysoIP.untagged)

# --- 3. Combine dataframes and annotate
violin_df <- bind_rows(df_ctrl, df_asah1) %>%
  mutate(Sample = factor(Sample, levels = c("Ctrl vs Untagged","ASAH1 vs Untagged")),
         Organelle = factor(Organelle, levels = organelles)) %>%
  left_join(org_mapping_detailed, by = "Organelle")

# --- 4. Compute mean log2FC per annotation and sample → save CSV
mean_log2fc_df <- violin_df %>%
  group_by(Organelle, Org_Label, Sample) %>%
  summarise(mean_log2FC = round(mean(log2FC, na.rm = TRUE), 2), .groups = "drop")

write.csv(mean_log2fc_df,file.path(out_dir_lysoip_organelle_dist, "LysoIP_detailedEndo_mean_log2FC.csv"),row.names = FALSE)

# --- 5. Plot
n_slots <- length(organelles) * 2
xs <- seq_len(n_slots)

p_violin <- ggplot(violin_df, aes(x = interaction(Organelle, Sample), y = log2FC, fill = Sample)) +
  geom_violin(scale = "width", trim = FALSE, width = 0.9) +
  geom_boxplot(width = 0.2, fill = "white", outlier.shape = NA, color = "black", linewidth = 0.3) +
  geom_jitter(width = 0.15, height = 0, size = 0.3, alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_fill_manual(values = c("Ctrl vs Untagged" = "grey70", "ASAH1 vs Untagged" = "dodgerblue3")) +
  scale_x_discrete(labels = setNames(
    rep(org_labels, times = 2),
    interaction(rep(org_mapping_detailed$Organelle, times = 2),
                rep(c("Ctrl vs Untagged","ASAH1 vs Untagged"), each = length(organelles)))
  )) +
  theme_bw(base_size = 10) +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 0),
        legend.position = "none") +
  labs(y = expression(log[2]*"FC"), title = "Organelle Enrichment HeLa Ctrl vs ASAH1-/-") +
  annotate("text", x = xs, y = -4.5, label = rep(org_order, times = 2),
           color = rep(c("darkorange","firebrick"), each = length(organelles)), size = 3) +
  annotate("text", x = mean(1:length(organelles)), y = -5.2, label = "Ctrl vs Untagged", size = 3.2) +
  annotate("text", x = mean((length(organelles)+1):n_slots), y = -5.2, label = "ASAH1 vs Untagged", size = 3.2)

# --- 6. Save
ggsave(file.path(out_dir_lysoip_organelle_dist, "LysoIP_OrganelleEnrichment_detailed_ViolinCombo.pdf"), p_violin, width = 5, height = 7, device = cairo_pdf)


```


### IVd. Co-purification / co-enrichment relative to lysosome mean per genotype
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Co-purification / co-enrichment relative to lysosome mean per genotype
# ─────────────────────────────────────────────────────────────────────────────

# --- 0. Define organelle annotations to evaluate (must be logical columns in helalysoip_annotated_df)
organelle_coenrich <- c("Lysosome", "Endo_iN_curated_Hundley", "Golgi", "ER", "Mito",
                        "EarlyEndosome", "RecyclingEndosome")
organelle_coenrich <- organelle_coenrich[organelle_coenrich %in% names(helalysoip_annotated_df)]

# --- 1. Helper to compute lysosome mean and co-enrichment flags for one genotype
compute_coenrichment_one <- function(df_in, sample_col, sample_label) {
  stopifnot(sample_col %in% names(df_in))
  # lysosome mean for this genotype
  lysosome_mean <- df_in %>%
    filter(Lysosome == TRUE) %>%
    pull(all_of(sample_col)) %>%
    { mean(., na.rm = TRUE) }

  # gene-level long df of organelle memberships
  long_membership <- df_in %>%
    dplyr::select(Gene.Symbol, all_of(sample_col), all_of(organelle_coenrich)) %>%
    pivot_longer(cols = all_of(organelle_coenrich),
                 names_to = "Organelle", values_to = "flag") %>%
    filter(flag) %>%
    mutate(Sample = sample_label,
           log2FC_value = .data[[sample_col]],
           lysosome_mean = lysosome_mean,
           sig_coenriched = is.finite(log2FC_value) & (log2FC_value >= lysosome_mean)) %>%
    dplyr::select(Gene.Symbol, Sample, Organelle, log2FC_value, lysosome_mean, sig_coenriched)

  return(long_membership)
}

# --- 2. Run for Ctrl vs Untagged and ASAH1 vs Untagged
coenrich_ctrl_df <- compute_coenrichment_one(
  df_in = helalysoip_annotated_df,
  sample_col = "log2FC_CtrlLysoIP.untagged",
  sample_label = "Ctrl vs Untagged"
)

coenrich_asah1_df <- compute_coenrichment_one(
  df_in = helalysoip_annotated_df,
  sample_col = "log2FC_ASAH1LysoIP.untagged",
  sample_label = "ASAH1 vs Untagged"
)

coenrich_genelevel_df <- bind_rows(coenrich_ctrl_df, coenrich_asah1_df)

# --- 3. Write protein-level CSV
write.csv(coenrich_genelevel_df,
          file = file.path(out_dir_lysoip_organelle_dist, "LysoIP_coenrichment_genelevel.csv"),
          row.names = FALSE)

# --- 4. Summarize counts per annotation and genotype
coenrich_summary_df <- coenrich_genelevel_df %>%
  group_by(Organelle, Sample) %>%
  summarise(
    n_organelle_genes = n(),
    n_sig_coenriched = sum(sig_coenriched, na.rm = TRUE),
    frac_sig_coenriched = ifelse(n_organelle_genes > 0, n_sig_coenriched / n_organelle_genes, NA_real_),
    lysosome_mean_ref = unique(lysosome_mean),
    .groups = "drop"
  ) %>%
  arrange(Organelle, Sample)

# --- 5. Determine organelle order (ascending by total sig across genotypes)
totals_order <- coenrich_summary_df %>%
  group_by(Organelle) %>%
  summarise(total_sig = sum(n_sig_coenriched, na.rm = TRUE), .groups = "drop") %>%
  arrange(total_sig)
organelle_order <- totals_order$Organelle


# --- 6. Write summary CSV
# add per-organelle gene lists and preserve order for plotting
genes_per_group_df <- coenrich_genelevel_df %>%
  filter(sig_coenriched) %>%
  group_by(Organelle, Sample) %>%
  summarise(genes_sig_coenriched = paste(sort(unique(as.character(Gene.Symbol))), collapse = ";"),
            .groups = "drop")

coenrich_summary_df <- coenrich_summary_df %>%
  left_join(genes_per_group_df, by = c("Organelle","Sample")) %>%
  mutate(genes_sig_coenriched = replace_na(genes_sig_coenriched, "")) %>%
  mutate(Organelle = factor(Organelle, levels = organelle_order)) %>%
  arrange(Organelle, Sample)

# save csv
write.csv(coenrich_summary_df, file.path(out_dir_lysoip_organelle_dist, "LysoIP_coenrichment_summary.csv"),row.names = FALSE)

# 6b. Direct Ctrl vs ASAH1 comparisons (paired Wilcoxon) for plotted organelles
ctrl_col  <- "log2FC_CtrlLysoIP.untagged"
asah1_col <- "log2FC_ASAH1LysoIP.untagged"

organelle_coenrich <- c("Lysosome","Endo_iN_curated_Hundley","Golgi","ER","Mito","EarlyEndosome","RecyclingEndosome")
organelle_coenrich <- organelle_coenrich[organelle_coenrich %in% names(helalysoip_annotated_df)]

compute_org_stats <- function(df_in, organelle_name) {
  df_org <- df_in %>%
    filter(.data[[organelle_name]] == TRUE) %>%
    transmute(Gene.Symbol = as.character(Gene.Symbol),
              ctrl = .data[[ctrl_col]],
              asah1 = .data[[asah1_col]]) %>%
    filter(is.finite(ctrl) & is.finite(asah1))

  if (nrow(df_org) == 0) {
    return(tibble(Organelle = organelle_name, n_genes = 0,
                  median_ctrl = NA_real_, median_asah1 = NA_real_, median_delta = NA_real_,
                  wilcox_p = NA_real_))
  }

  wt <- suppressWarnings(wilcox.test(df_org$ctrl, df_org$asah1, paired = TRUE, exact = FALSE))
  delta <- df_org$asah1 - df_org$ctrl

  tibble(Organelle = organelle_name,
         n_genes = nrow(df_org),
         median_ctrl = median(df_org$ctrl, na.rm = TRUE),
         median_asah1 = median(df_org$asah1, na.rm = TRUE),
         median_delta = median(delta, na.rm = TRUE),
         wilcox_p = wt$p.value)
}

# helper to add stars
p_to_stars <- function(p) {
  ifelse(is.na(p), "",
    ifelse(p < 0.001, "***",
      ifelse(p < 0.01, "**",
        ifelse(p < 0.05, "*", ""))))
}

# add stats to csv file as new columns
org_stats_df <- bind_rows(lapply(organelle_coenrich, function(an) compute_org_stats(helalysoip_annotated_df, an))) %>%
  mutate(wilcox_p_BH = p.adjust(wilcox_p, method = "BH"),
         wilcox_FDR  = wilcox_p_BH,
         sig_raw     = p_to_stars(wilcox_p),
         sig_BH      = p_to_stars(wilcox_p_BH),
         sig_FDR     = p_to_stars(wilcox_FDR))

# save csv
write.csv(org_stats_df,file.path(out_dir_lysoip_organelle_dist, "LysoIP_ctrl_vs_ASAH1_comparisons.csv"),row.names = FALSE)


# --- 7. Stacked bar plot: counts of sig. co-enriched per annotation
coenrich_bar_plot <- ggplot(coenrich_summary_df, aes(x = n_sig_coenriched, y = Organelle, fill = Sample)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.1) +
  geom_text(aes(label = ifelse(n_sig_coenriched > 0, n_sig_coenriched, "")), position = position_stack(vjust = 0.5), size = 2.8) +
  scale_fill_manual(values = c("Ctrl vs Untagged" = "grey70", "ASAH1 vs Untagged" = "dodgerblue3")) +
  labs(x = NULL, y = "# sig. co-enriched (log2FC ≥ mean Lyso)", title = "Co-enrichment relative to Lysosome mean (per genotype)") +
  theme_bw(base_size = 6) +
  theme(
    panel.grid = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1), 
    legend.title = element_blank())


# --- 8. Save plot as PDF 
ggsave(file.path(out_dir_lysoip_organelle_dist, "LysoIP_coenrichment_stackedbar_vertical_sorted_labeled.pdf"), coenrich_bar_plot, width = 6, height = 4, device = cairo_pdf)



```


### IVe. Overlap of sig co-enriched IDs across annotations + Venns
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Overlap of sig co-enriched IDs across annotations + Venns
# ─────────────────────────────────────────────────────────────────────────────

# --- 1.  Targets (exclude Lysosome itself)
venn_ann6 <- c("Endo_iN_curated_Hundley","Golgi","ER","Mito","EarlyEndosome","RecyclingEndosome")

# --- 2. Sets of sig_coenriched genes per genotype (coerce to character)
get_sig_sets <- function(genelevel_df, sample_label, ann_vec) {
  genelevel_df %>%
    filter(Sample == sample_label, sig_coenriched, Organelle %in% ann_vec) %>%
    group_by(Organelle) %>%
    summarise(Genes = list(unique(as.character(Gene.Symbol))), .groups = "drop") %>%
    { setNames(.$Genes, .$Organelle) }
}

ctrl_sets  <- get_sig_sets(coenrich_genelevel_df, "Ctrl vs Untagged",  venn_ann6)
asah1_sets <- get_sig_sets(coenrich_genelevel_df, "ASAH1 vs Untagged", venn_ann6)

# --- 3. Helpers: size filter, sanitize, Venn
min_size <- 5

filter_by_size <- function(setlist, k = min_size) {
  if (is.null(setlist)) return(list())
  keep <- vapply(setlist, function(v) length(unique(as.character(v))) >= k, logical(1))
  setlist[keep]
}

sanitize_sets <- function(setlist) {
  if (is.null(setlist)) return(list())
  setlist <- lapply(setlist, function(v) unique(as.character(v[!is.na(v) & nzchar(v)])))
  setlist[vapply(setlist, length, integer(1)) > 0]
}

plot_venn <- function(setlist, fname, width = 16, height = 14, label_size_pt = 6) {
  setlist <- sanitize_sets(setlist)
  if (length(setlist) < 2) return(invisible(NULL))
  p <- ggVennDiagram::ggVennDiagram(setlist, show_intersect = FALSE, set_color = "black", set_size = NA,
                                    label = "both", label_alpha = 0, label_geom = "label",
                                    label_color = "black", label_size = label_size_pt,
                                    edge_lty = "solid", edge_size = 0) +
       scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") + theme_void()
  ggsave(file.path(out_dir_lysoip_venn_coenrichment, fname), p, width = width, height = height, units = "cm", dpi = 600, device = cairo_pdf)
}

# --- 4. Apply size filter per genotype
ctrl_sets5  <- filter_by_size(ctrl_sets,  min_size)
asah1_sets5 <- filter_by_size(asah1_sets, min_size)

# Log which annotations passed size filter
write.csv(tibble(Genotype = "Ctrl", Annotation = names(ctrl_sets5), n = vapply(ctrl_sets5, length, 1L)), file.path(out_dir_lysoip_venn_coenrichment, "kept_annotations_CTRL_min5.csv"), row.names = FALSE)
write.csv(tibble(Genotype = "ASAH1", Annotation = names(asah1_sets5), n = vapply(asah1_sets5, length, 1L)), file.path(out_dir_lysoip_venn_coenrichment, "kept_annotations_ASAH1_min5.csv"), row.names = FALSE)


# --- 5. Per-annotation Ctrl vs ASAH1 overlap (requires ≥5 on both sides)
anns_to_compare <- intersect(names(ctrl_sets5), names(asah1_sets5))
if (length(anns_to_compare) > 0) {
  per_ann_summary <- purrr::map_dfr(anns_to_compare, function(an) {
    ctrl_vec  <- unique(as.character(ctrl_sets5[[an]]))
    asah1_vec <- unique(as.character(asah1_sets5[[an]]))
    shared    <- intersect(ctrl_vec, asah1_vec)
    union_set <- union(ctrl_vec, asah1_vec)
    jacc      <- ifelse(length(union_set) > 0, length(shared) / length(union_set), NA_real_)
    base      <- gsub("[^A-Za-z0-9]+","_", an)

    # save gene lists overlap and unique for each annotations 
    pad_to <- function(x, n) { c(x, rep(NA_character_, n - length(x))) }
      ctrl_only  <- sort(unique(setdiff(ctrl_vec, asah1_vec)))
      asah1_only <- sort(unique(setdiff(asah1_vec, ctrl_vec)))
      shared     <- sort(unique(intersect(ctrl_vec, asah1_vec)))
      
      # make one CSV per plot: ASAH1_unique, Shared, Control_unique
      nmax <- max(length(asah1_only), length(shared), length(ctrl_only))
      venn_table <- tibble(
        ASAH1_unique   = pad_to(asah1_only, nmax),
        Shared         = pad_to(shared,     nmax),
        Control_unique = pad_to(ctrl_only,  nmax)
      )
      
      csv_name <- paste0("Venn_", base, "_Ctrl_vs_ASAH1_min5.csv")
      write.csv(venn_table, file.path(out_dir_lysoip_venn_coenrichment, csv_name), row.names = FALSE)
    # Venn
    two_set <- list("Ctrl sig" = ctrl_vec, "ASAH1 sig" = asah1_vec)
    plot_venn(two_set, paste0("Venn_", base, "_Ctrl_vs_ASAH1_min5.pdf"), width = 12, height = 10)

    tibble(Annotation = an, n_ctrl = length(ctrl_vec), n_asah1 = length(asah1_vec), n_shared = length(shared), jaccard_ctrl_asah1 = jacc)
  })

  write.csv(per_ann_summary, file.path(out_dir_lysoip_venn_coenrichment, "per_annotation_Ctrl_vs_ASAH1_overlap_summary_min5.csv"), row.names = FALSE)
} else {
  message("No annotations with ≥", min_size, " sig IDs in both Ctrl and ASAH1.")
}
```



### IVf. GO enrichment of Endo annotations sig. Control
```{r}
# pick target annotations
target_ann <- c("Endo_iN_curated_Hundley","EarlyEndosome")

# helper: run topGO for a set of symbols
run_topGO <- function(sym_vec, universe_vec, ann_label, ont = c("BP","CC","MF"), min_genes = 10) {
  if (length(sym_vec) < min_genes) {
    message("Skip ", ann_label, ": n=", length(sym_vec), " < ", min_genes)
    return(invisible(NULL))
  }

  geneList <- factor(as.integer(universe_vec %in% sym_vec))
  names(geneList) <- universe_vec

  for (o in ont) {
    GOdata <- new("topGOdata",
                  ontology = o,
                  allGenes = geneList,
                  geneSel = function(x) x == 1,
                  annot = annFUN.org,
                  mapping = "org.Hs.eg.db",
                  ID = "symbol")

    resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

    go_table <- GenTable(GOdata, classicFisher = resultFisher, topNodes = 30)
    go_table <- go_table %>%
      mutate(raw_pval   = score(resultFisher)[GO.ID],
             log10P     = -log10(raw_pval),
             Significant = as.numeric(Significant),
             Expected    = as.numeric(Expected),
             Enrichment  = Significant / Expected) %>%
      filter(!is.na(raw_pval)) %>%
      arrange(raw_pval)

    # genes per term
    go_genes <- genesInTerm(GOdata)
    go_table$Genes <- sapply(go_table$GO.ID, function(go_id) {
      intersect(go_genes[[go_id]], sym_vec) %>% paste(collapse = ";")
    })

    # save
    clean_ann <- gsub("[^A-Za-z0-9]+","_", ann_label)
    write.csv(go_table, file.path(out_dir_lysoip_GOcoenrichment, paste0("GO_", o, "_CTRL_", clean_ann, ".csv")), row.names = FALSE)

    p <- ggplot(go_table[1:min(10,nrow(go_table)), ], aes(x = Enrichment, y = reorder(Term, Enrichment))) +
      geom_point(aes(size = log10P, fill = Enrichment), shape = 21, color = "black", stroke = 0.25) +
      scale_fill_gradient(low = "white", high = "firebrick") +
      theme_bw(base_size = 8) +
      labs(x = "Fold Enrichment", y = "GO Term", size = expression(-log[10]*"P"), title = paste0("GO-", o, " (Ctrl sig) — ", ann_label)) +
      theme(panel.grid = element_blank())
    ggsave(file.path(out_dir_lysoip_GOcoenrichment, paste0("GO_", o, "_CTRL_", clean_ann, ".pdf")), p, width = 5, height = 3.5, device = cairo_pdf)
  }
}

# --- extract universe (all genes across coenrich_summary_df)
universe_genes <- coenrich_summary_df %>%
  pull(genes_sig_coenriched) %>%
  str_split(";") %>%
  unlist() %>%
  unique()

# --- loop over target annotations
for (ann in target_ann) {
  ctrl_syms <- coenrich_summary_df %>%
    filter(Sample == "Ctrl vs Untagged", Organelle == ann) %>%
    pull(genes_sig_coenriched) %>%
    str_split(";") %>%
    unlist() %>%
    unique()

  run_topGO(ctrl_syms, universe_genes, ann)
}

```




# V. Focused lyso evaluations
### Va. Abundnance of lysosomal hydrolases
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Hydrolase abundance in lysosome Ctrl vs ASAH1-/-
# ─────────────────────────────────────────────────────────────────────────────

# --- add lyso hydrolases to df
helalysoip_annotated_df$LysoHydrolase <- helalysoip_annotated_df$Gene.Symbol %in% lysohydrolases


# --- Step 1: Prepare data for slope plot
hydro_df <- helalysoip_annotated_df %>% 
  filter(LysoHydrolase == TRUE) %>%
  dplyr::select(Gene.Symbol,
         Ctrl = log2FC_CtrlLysoIP.untagged,
         ASAH1 = log2FC_ASAH1LysoIP.untagged) %>%
  pivot_longer(cols = c("Ctrl", "ASAH1"), names_to = "Sample", values_to = "log2FC") %>%
  mutate(Gene.Symbol = factor(Gene.Symbol, levels = unique(Gene.Symbol)))  # maintain order

# Ensure factor levels enforce desired order
hydro_df <- hydro_df %>%
  mutate(Sample = factor(Sample, levels = c("Ctrl", "ASAH1")))

# --- Step 2: Plot slope plot
p_slope <- ggplot(hydro_df, aes(x = Sample, y = log2FC, group = Gene.Symbol)) +
  geom_line(color = "grey90", alpha = 0.8) +
  geom_point(size = 3, aes(fill = Sample), shape = 21, color = "black") +
  scale_fill_manual(values = c("Ctrl" = "grey70", "ASAH1" = "dodgerblue3")) +
  theme_bw(base_size = 6) +
  labs(x = NULL, y = expression(log[2]*"FC"),
       title = "Lysohydrolase Abundance in Ctrl vs ASAH1-/-") +
  theme(
    panel.grid = element_blank(),
    aspect.ratio = 1.75,
    legend.position = "none"
  )

# --- Step 3: Save as square PDF
ggsave(filename = file.path(out_dir_lysoip_organelle_dist,"lysohydrolase_slopeplot.pdf"),plot = p_slope, width = 2, height = 4, device = cairo_pdf)



# ─────────────────────────────────────────────────────────────────────────────
# Hydrolase abundance in lysosome Ctrl vs ASAH1-/-
# heatmap 
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1: Prepare matrix for heatmap
heatmap_matrix_lysohydro <- helalysoip_annotated_df %>%
  filter(LysoHydrolase == TRUE) %>%
  dplyr::select(Gene.Symbol,
         Ctrl = log2FC_CtrlLysoIP.untagged,
         ASAH1 = log2FC_ASAH1LysoIP.untagged) %>%
  column_to_rownames("Gene.Symbol") %>%
  as.matrix()


# --- Step 2: Plot heatmap with centered scale
pdf(file = file.path(out_dir_lysoip_organelle_dist, "lysohydrolase_heatmap.pdf"),
    width = 2, height = 5, useDingbats = FALSE)

pheatmap(
  heatmap_matrix_lysohydro,
  color = c(colorRampPalette(c("dodgerblue3", "white"))(50),
            colorRampPalette(c("white", "firebrick"))(50)),
  breaks = c(seq(-2, 0, length.out = 51),
             seq(0, 4, length.out = 51)[-1]),  # avoid duplicate 0
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  na_col = "grey70",
  cellheight = 15,
  cellwidth = 15,
  border_color = "black",
  lwd = 0.25,
  main = "Lysosomal Hydrolase log2FC (Ctrl vs ASAH1−/−)"
)
dev.off()
```


### Vb. Rainfallplots of ASAH1 vs Control LysoIP for dplyr::select annotations 
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Rainfall plot: ASAH1LysoIP vs CtrlLysoIP (highlight top 5 by log2FC)
# ─────────────────────────────────────────────────────────────────────────────

# --- 1. Define gene sets
vatpase_genes <- c(
  "ATP6V1A","ATP6V1B1","ATP6V1B2","ATP6V1C1","ATP6V1C2","ATP6V1D","ATP6V1E1","ATP6V1E2","ATP6V1F",
  "ATP6V1G1","ATP6V1G2","ATP6V1G3","ATP6V1H","ATP6V0A1","ATP6V0A2","ATP6V0A3","ATP6V0A4",
  "ATP6V0C","ATP6V0B","ATP6V0D1","ATP6V0D2","ATP6V0E1","ATP6V0E2","RNASEK",
  "ATP6AP1","ATP6AP2","DMXL1","DMXL2","WDR7","ROGDI", "PODXL"
)

lysohydrolases <- c(
  "ASAH1","GBA","ARSA","ARSB","GALNS","IDUA","GLA","NEU1","MAN2B1",
  "GLB1","HEXA","GNS","HEXB","CTSA","CTSZ","TPP1","NAGLU","CTSL",
  "CTSB","NAGA","LGMN"
)

# --- 2. Subset for lysosome / hydrolases / vATPase
rain_df <- helalysoip_annotated_df %>%
  filter(
    Gene.Symbol %in% vatpase_genes |
    Gene.Symbol %in% lysohydrolases |
    Lysosome == TRUE
  ) %>%
  mutate(
    y_group = case_when(
      Gene.Symbol %in% lysohydrolases ~ "Hydrolases",
      Gene.Symbol %in% vatpase_genes  ~ "vATPase",
      Lysosome == TRUE                ~ "Lysosome",
      TRUE ~ NA_character_
    ),
    log2fc = log2FC_ASAH1LysoIP.CtrlLysoIP
  ) %>%
  filter(!is.na(y_group), is.finite(log2fc)) %>%
  mutate(
    y_group = factor(y_group, levels = c("Lysosome","Hydrolases","vATPase"))
  )

# --- 3. Pick top 5 by absolute FC within each group
label_df <- rain_df %>%
  group_by(y_group) %>%
  slice_max(order_by = abs(log2fc), n = 5, with_ties = FALSE) %>%
  ungroup()


# # --- 4. Plot (old)
# rainfall_plot <- ggplot(
#   rain_df,
#   aes(y = log2fc, x = y_group, fill = y_group)
# ) +
#   gghalves::geom_half_point(
#     side = "b",
#     range_scale = 0.2,
#     alpha = 0.2,
#     size = 1,
#     shape = 16
#   ) +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
#   ggdist::stat_halfeye(
#     adjust = 2,
#     width = 0.6,
#     .width = 0,
#     justification = -0.2,
#     point_colour = NA
#   ) +
#   geom_boxplot(width = 0.15, outlier.shape = NA) +
#   geom_text_repel(
#     data = label_df,
#     aes(label = Gene.Symbol),
#     position = position_nudge(x = -0.25),
#     size = 2,
#     min.segment.length = 0,
#     max.overlaps = Inf,
#     box.padding = 0.2,
#     point.padding = 0.1,
#     segment.size = 0.2
#   ) +
#   scale_fill_manual(values = c(
#     "Lysosome"   = "grey70",
#     "Hydrolases" = "firebrick",
#     "vATPase"    = "dodgerblue3"
#   )) +
#   coord_flip() +
#   theme_bw(base_size = 6) +
#   labs(y = "log2FC ASAH1LysoIP vs CtrlLysoIP", x = NULL,
#        title = "ASAH1 vs Ctrl: Lysosome / Hydrolases / vATPase") +
#   theme(panel.grid = element_blank(), legend.position = "none")

# # --- 4b. ∆log2FC (norm) (Δlog2FC = ASAH1_vs_untagged − Ctrl_vs_untagged)
# rain_df_delta <- helalysoip_annotated_df %>%
#   filter(Gene.Symbol %in% vatpase_genes | Gene.Symbol %in% lysohydrolases | Lysosome == TRUE) %>%
#   mutate(
#     y_group = case_when(
#       Gene.Symbol %in% lysohydrolases ~ "Hydrolases",
#       Gene.Symbol %in% vatpase_genes  ~ "vATPase",
#       Lysosome == TRUE                ~ "Lysosome",
#       TRUE ~ NA_character_
#     ),
#     delta_log2fc = log2FC_ASAH1LysoIP.untagged - log2FC_CtrlLysoIP.untagged 
#   ) %>%
#   filter(!is.na(y_group), is.finite(delta_log2fc)) %>%
#   mutate(y_group = factor(y_group, levels = c("Lysosome","Hydrolases","vATPase")))

# # top 5 by Δlog2FC within each group
# label_df_delta <- rain_df_delta %>%
#   group_by(y_group) %>%
#   slice_max(order_by = abs(delta_log2fc), n = 5, with_ties = FALSE) %>%
#   ungroup()

# # plot
# rainfall_plot_delta <- ggplot(rain_df_delta, aes(y = delta_log2fc, x = y_group, fill = y_group)) +
#   gghalves::geom_half_point(side = "b", range_scale = 0.2, alpha = 0.2, size = 1, shape = 16) +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
#   ggdist::stat_halfeye(adjust = 2, width = 0.6, .width = 0, justification = -0.2, point_colour = NA) +
#   geom_boxplot(width = 0.15, outlier.shape = NA) +
#   geom_text_repel(data = label_df_delta, aes(label = Gene.Symbol), position = position_nudge(x = -0.25),
#                   size = 2, min.segment.length = 0, max.overlaps = Inf, box.padding = 0.2,
#                   point.padding = 0.1, segment.size = 0.2) +
#   scale_fill_manual(values = c("Lysosome" = "grey70", "Hydrolases" = "firebrick", "vATPase" = "dodgerblue3")) +
#   coord_flip() +
#   theme_bw(base_size = 6) +
#   labs(y = "∆FC (ASAH1−Ctrl)",
#        x = NULL,
#        title = "Δlog2FC ASAH1 vs Ctrl — Lysosome / Hydrolases / vATPase") +
#   theme(panel.grid = element_blank(), legend.position = "none")


# # --- 5. Save
# ggsave(file.path(out_dir_lysoip_organelle_rainfall, "LysoIP_rainfall_deltalog2FC_LysoHydroVATPase.pdf"), rainfall_plot_delta, width = 4, height = 2, device = cairo_pdf)
# ggsave(filename = file.path(out_dir_lysoip_organelle_rainfall, "LysoIP_rainfall_ASAH1vsCtrl_LysoHydroVATPase.pdf"),plot = rainfall_plot, width = 4, height = 2, device = cairo_pdf)




# --- 4. Plot
rainfall_plot <- ggplot(
  rain_df,
  aes(y = log2fc, x = y_group, fill = y_group)
) +
  geom_jitter(
    position = position_jitter(width = 0.15, height = 0),
    alpha = 0.2,
    size = 1,
    shape = 16
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  ggdist::stat_halfeye(
    adjust = 2,
    width = 0.6,
    .width = 0,
    justification = -0.2,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_text_repel(
    data = label_df,
    aes(label = Gene.Symbol),
    position = position_nudge(x = -0.25),
    size = 2,
    min.segment.length = 0,
    max.overlaps = Inf,
    box.padding = 0.2,
    point.padding = 0.1,
    segment.size = 0.2
  ) +
  scale_fill_manual(values = c(
    "Lysosome"   = "grey70",
    "Hydrolases" = "firebrick",
    "vATPase"    = "dodgerblue3"
  )) +
  coord_flip() +
  theme_bw(base_size = 6) +
  labs(y = "log2FC ASAH1LysoIP vs CtrlLysoIP", x = NULL,
       title = "ASAH1 vs Ctrl: Lysosome / Hydrolases / vATPase") +
  theme(panel.grid = element_blank(), legend.position = "none")

# --- 4b. ∆log2FC (norm) (Δlog2FC = ASAH1_vs_untagged − Ctrl_vs_untagged)
rainfall_plot_delta <- ggplot(rain_df_delta, aes(y = delta_log2fc, x = y_group, fill = y_group)) +
  geom_jitter(
    position = position_jitter(width = 0.15, height = 0),
    alpha = 0.2,
    size = 1,
    shape = 16
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  ggdist::stat_halfeye(
    adjust = 2,
    width = 0.6,
    .width = 0,
    justification = -0.2,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_text_repel(
    data = label_df_delta,
    aes(label = Gene.Symbol),
    position = position_nudge(x = -0.1),
    size = 2,
    min.segment.length = 0,
    max.overlaps = Inf,
    box.padding = 0,        # changed from 0.1 to 0
    point.padding = 0,      # changed from 0.05 to 0
    segment.size = 0.1,
    force = 100,             
    direction = "x"
  ) +
  scale_fill_manual(values = c("Lysosome" = "grey70", "Hydrolases" = "firebrick", "vATPase" = "dodgerblue3")) +
  coord_flip() +
  theme_bw(base_size = 6) +
  labs(y = "∆FC (ASAH1−Ctrl)",
       x = NULL,
       title = "Δlog2FC ASAH1 vs Ctrl — Lysosome / Hydrolases / vATPase") +
  theme(panel.grid = element_blank(), legend.position = "none")

# --- 5. Save
ggsave(file.path(out_dir_lysoip_organelle_rainfall, "LysoIP_rainfall_deltalog2FC_LysoHydroVATPase.pdf"), rainfall_plot_delta, width = 4, height = 2, device = cairo_pdf)
ggsave(filename = file.path(out_dir_lysoip_organelle_rainfall, "LysoIP_rainfall_ASAH1vsCtrl_LysoHydroVATPase.pdf"), plot = rainfall_plot, width = 4, height = 2, device = cairo_pdf)

```




# VI. Venn Diagramm of lyso, endo and iN-endo list to see overlap
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Venn for all proteins detected
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1: read in .csv file containing the genes which are sig down in proteomics
lysoIPvenn <- read.csv('/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/MSstats/OrganelleEval/EndoLysoIP_venn.csv', 
                   header=T, sep = ",", , stringsAsFactors = FALSE)


# --- Step 2: Venn of all three datasets #####
ggVennDiagram(
  lysoIPvenn,
  category.names = c("Lyso-enriched", "Endo-enriched", "iNendo-enriched"),
  show_intersect = FALSE,
  set_color = "black",
  set_size = NA,
  label = c("both"),
  label_alpha = 0,
  label_geom = c("label"),
  label_color = "black",
  label_size = NA,
  label_percent_digit = 0,
  label_txtWidth = 40,
  edge_lty = "solid",
  edge_size = 0,
) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")


# --- Step 3: Save Venn diagram as .pdf for further processing in illustrator for figures 
ggsave(file.path(out_dir_lysoip_venn,'EndoLysoVennDiagramm.pdf'), width =20, height = 20, units = "cm", dpi=600)



# ─────────────────────────────────────────────────────────────────────────────
# Venn of Lyso and iN endo list
# ─────────────────────────────────────────────────────────────────────────────
# --- Step 4: make subset of IP data to only contain lyso and Endocurated list 
lysoIPvenn_subest <- lysoIPvenn[,c("Lyso", "EndoCurated")]

ggVennDiagram(
  lysoIPvenn_subest,
  category.names = c("Lyso-enriched", "iNendo-enriched"),
  show_intersect = FALSE,
  set_color = "black",
  set_size = NA,
  label = c("both"),
  label_alpha = 0,
  label_geom = c("label"),
  label_color = "black",
  label_size = NA,
  label_percent_digit = 0,
  label_txtWidth = 40,
  edge_lty = "solid",
  edge_size = 0,
) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
  coord_flip() ## flip plot on side 


# --- Step 5: save Venn diagram as .pdf for further processing in illustrator for figures 
ggsave(file.path(out_dir_lysoip_venn, 'EndoCuratedLysoVennDiagramm.pdf'), width =27, height = 20, units = "cm", dpi=600)


# --- Step 6: Extract overlapping protein names #### 
## parse and save overlapping protein names that between the lists 
# Convert columns to sets (remove NA values)
lyso_set <- na.omit(lysoIPvenn$Lyso)
endosome_set <- na.omit(lysoIPvenn$Endosome)
endocurated_set <- na.omit(lysoIPvenn$EndoCurated)


# --- Step 7: Compute pairwise intersections
overlap_lyso_endosome <- intersect(lyso_set, endosome_set)
overlap_lyso_endocurated <- intersect(lyso_set, endocurated_set)
overlap_endosome_endocurated <- intersect(endosome_set, endocurated_set)

# --- Step 8: Save the results to CSV files
write.csv(data.frame(Overlap_Lyso_Endosome = overlap_lyso_endosome),
          '/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/MSstats/OrganelleEval/Overlap_Lyso_Endosome.csv', row.names = FALSE)
write.csv(data.frame(Overlap_Lyso_EndoCurated = overlap_lyso_endocurated),
          '/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/MSstats/OrganelleEval/Overlap_Lyso_EndoCurated.csv', row.names = FALSE)
write.csv(data.frame(Overlap_Endosome_EndoCurated = overlap_endosome_endocurated),
          '/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/MSstats/OrganelleEval/Overlap_Endosome_EndoCurated.csv', row.names = FALSE)




# ─────────────────────────────────────────────────────────────────────────────
# Analysis for enriched protein (log2FC >0.5) in Ctrl over untagged
# ─────────────────────────────────────────────────────────────────────────────
# --- Step 9: Read in .csv file containing the genes which are sig down in proteomics
lysoIPvennCtrl <- read.csv('/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/MSstats/OrganelleEval/EndoLysoIP_vennCtrl.csv', 
                   header=T, sep = ",", , stringsAsFactors = FALSE)


# --- Step 10: Make subset of IP data to only contain lyso and Endocurated list & plot
lysoIPvennCtrl_subest <- lysoIPvennCtrl[,c("Lyso", "EndoCurated")]

ggVennDiagram(
  lysoIPvennCtrl_subest,
  category.names = c("Lyso-enriched", "iNendo-enriched"),
  show_intersect = FALSE,
  set_color = "black",
  set_size = NA,
  label = c("both"),
  label_alpha = 0,
  label_geom = c("label"),
  label_color = "black",
  label_size = NA,
  label_percent_digit = 0,
  label_txtWidth = 40,
  edge_lty = "solid",
  edge_size = 0,
) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
  coord_flip() ## flip plot on side 


# --- Step 11: Save Venn diagram as .pdf for further processing in illustrator for figures 
ggsave(file.path(out_dir_lysoip_venn,'EndoCuratedLysoVennDiagrammCtrl.pdf'), width =27, height = 20, units = "cm", dpi=600)


# --- Step 12: Ctrl over untagged: Extract overlapping protein names 
## parse and save overlapping protein names that between the lists 

# Convert columns to sets (remove NA values)
lyso_Ctrl_set <- na.omit(lysoIPvennCtrl$Lyso)
endosome_Ctrl_set <- na.omit(lysoIPvennCtrl$Endosome)
endocurated_Ctrl_set <- na.omit(lysoIPvennCtrl$EndoCurated)

# Compute pairwise intersections
overlap_Ctrl_lyso_endosome <- intersect(lyso_Ctrl_set, endosome_Ctrl_set)
overlap_Ctrl_lyso_endocurated <- intersect(lyso_Ctrl_set, endocurated_Ctrl_set)
overlap_Ctrl_endosome_endocurated <- intersect(endosome_Ctrl_set, endocurated_Ctrl_set)

# Save the results to CSV files
write.csv(data.frame(Overlap_Ctrl_Lyso_Endosome = overlap_Ctrl_lyso_endosome),file=file.path(out_dir_lysoip_venn,
          'Overlap_Ctrl_Lyso_Endosome.csv'), row.names = FALSE)
write.csv(data.frame(Overlap_Ctrl_Lyso_EndoCurated = overlap_Ctrl_lyso_endocurated),file=file.path(out_dir_lysoip_venn,
          'Overlap_Ctrl_Lyso_EndoCurated.csv'), row.names = FALSE)
write.csv(data.frame(Overlap_Ctrl_Endosome_EndoCurated = overlap_Ctrl_endosome_endocurated),file=file.path(out_dir_lysoip_venn,
          'Overlap_Ctrl_Endosome_EndoCurated.csv'), row.names = FALSE)





# ─────────────────────────────────────────────────────────────────────────────
# Analysis for enriched protein (log2FC >0.5) in ASAH1 over untagged 
# ─────────────────────────────────────────────────────────────────────────────
# --- Step 13: Read in .csv file containing the genes which are sig down in proteomics
lysoIPvennASAH1 <- read.csv('/Users/felix/Documents/PostDoc/Harvard/03_LSD-PD/Proteomics/20241119_HeLa_Ctrl-ASAH1_LysoIP/MSstats/OrganelleEval/EndoLysoIP_vennASAH1.csv', 
                   header=T, sep = ",", , stringsAsFactors = FALSE)


# --- Step 14: Make subset of IP data to only contain lyso and Endocurated list & plot
lysoIPvennASAH1_subest <- lysoIPvennASAH1[,c("Lyso", "EndoCurated")]

ggVennDiagram(
  lysoIPvennASAH1_subest,
  category.names = c("Lyso-enriched", "iNendo-enriched"),
  show_intersect = FALSE,
  set_color = "black",
  set_size = NA,
  label = c("both"),
  label_alpha = 0,
  label_geom = c("label"),
  label_color = "black",
  label_size = NA,
  label_percent_digit = 0,
  label_txtWidth = 40,
  edge_lty = "solid",
  edge_size = 0,
) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
  coord_flip() ## flip plot on side 


# --- Step 15: Save Venn diagram as .pdf for further processing in illustrator for figures 
ggsave(file.path(out_dir_lysoip_venn,'EndoCuratedLysoVennDiagrammASAH1.pdf'), width =27, height = 20, units = "cm", dpi=600)


# --- Step 16: Ctrl over untagged: Extract overlapping protein names 
## parse and save overlapping protein names that between the lists 
# Convert columns to sets (remove NA values)
lyso_ASAH1_set <- na.omit(lysoIPvennASAH1$Lyso)
endosome_ASAH1_set <- na.omit(lysoIPvennASAH1$Endosome)
endocurated_ASAH1_set <- na.omit(lysoIPvennASAH1$EndoCurated)

# Compute pairwise intersections
overlap_ASAH1_lyso_endosome <- intersect(lyso_ASAH1_set, endosome_ASAH1_set)
overlap_ASAH1_lyso_endocurated <- intersect(lyso_ASAH1_set, endocurated_ASAH1_set)
overlap_ASAH1_endosome_endocurated <- intersect(endosome_ASAH1_set, endocurated_ASAH1_set)

# Save the results to CSV files
write.csv(data.frame(Overlap_ASAH1_Lyso_Endosome = overlap_ASAH1_lyso_endosome),file=file.path(out_dir_lysoip_venn,
          'Overlap_ASAH1_Lyso_Endosome.csv'), row.names = FALSE)
write.csv(data.frame(Overlap_ASAH1_Lyso_EndoCurated = overlap_ASAH1_lyso_endocurated),file=file.path(out_dir_lysoip_venn,
          '/Overlap_ASAH1_Lyso_EndoCurated.csv'), row.names = FALSE)
write.csv(data.frame(Overlap_ASAH1_Endosome_EndoCurated = overlap_ASAH1_endosome_endocurated),file=file.path(out_dir_lysoip_venn,
          'Overlap_ASAH1_Endosome_EndoCurated.csv'), row.names = FALSE)



```


## VII. Autophagy investigation in more detail — stacked bars + UpSet
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# VII. Autophagy investigation in more detail — stacked bars + UpSet
# ─────────────────────────────────────────────────────────────────────────────
library(ComplexUpset)

# --- 0.  define annotation lists:
autophagy_receptors <- c(
  "AMBRA1", "BCL2L13", "BECN1", "BNIP3L", "CALCOCO2", "CCPG1", "CDK5RAP3", "FUNDC1", 
  "LGALS3", "LGALS8", "NBR1", "NCOA4", "NLRX1", "OPTN", "RETREG1", "SQSTM1", 
  "STBD1", "TAX1BP1", "TEX264", "TOLLIP", "TRIM5"
)
autophagy_core <- c(
  "ATG12", "ATG13", "ATG14", "ATG16L1", "ATG3", "ATG4A", "ATG4B", "ATG4C", "ATG5", 
  "ATG7", "ATG9A", "GABARAP", "GABARAPL2", "MAP1LC3A", "MAP1LC3B", "MTOR", "PIK3C3", 
  "PIK3R4", "PLEKHM1", "RAB5A", "RAB7A", "RB1CC1", "RUBCN", "STX17", "ULK1", 
  "ULK2", "UVRAG", "WDR45", "WDR45B", "WIPI1", "WIPI2", "ZFYVE1"
)
golgi_trans <- c(
  "PACS1", "VPS45", "VPS41", "GGA3", "ARFIP2", "GGA1", "GOLPH3L", "GOPC", 
  "AP4M1", "GGA2", "OSBPL11", "GOLGA1", "FAM91A1", "C17orf75", "OSBPL9", "GOLGA4", 
  "VPS51", "TBC1D23", "VPS52", "CCDC91", "ARFIP1", "AP4B1", "AP4E1", "IMPAD1"
)
golgi_trans <- c(
  "PACS1", "VPS45", "VPS41", "GGA3", "ARFIP2", "GGA1", "GOLPH3L", "GOPC", 
  "AP4M1", "GGA2", "OSBPL11", "GOLGA1", "FAM91A1", "C17orf75", "OSBPL9", "GOLGA4", 
  "VPS51", "TBC1D23", "VPS52", "CCDC91", "ARFIP1", "AP4B1", "AP4E1", "IMPAD1"
)

golgi_cis <- c(
  "TMED10", "TRAPPC3", "TMED1", "TRAPPC10", "COPG2", "TRAPPC5", "COPB1", "COPB2", 
  "GOLM1", "COPG1", "B3GAT3", "TRAPPC11", "TRAPPC6B", "TMED2", "GOSR2", "TRAPPC8", 
  "TRAPPC4", "GBF1", "TMED5", "TMED9", "GOLGA5", "SEB22B", "NUCB1", "GOLGA2"
)
ergic <- c(
  "TMED10", "TMED1", "ATP6AP1", "RAB1B", "RAB2A", "TAOK2", "NXF1", "TMEM181", 
  "STX16", "RAB2B", "RAB6B", "USP46", "SURF4", "GLE1", "GORASP1", "LMAN2L", 
  "ECPAS", "TMED2", "UGGT1", "TRAPPC12", "DNM2", "SEH1L", "MARCKS", "GBF1", "RDX", 
  "BCAP31", "DST", "ERGIC3", "TMED9", "LMAN1", "SEB22B", "LMAN2", "STX5"
)

annotation_lists <- list(
  autophagy_receptors = autophagy_receptors,
  autophagy_core = autophagy_core,
  golgi_trans = golgi_trans,
  golgi_cis = golgi_cis,
  ergic = ergic,
  ER = helalysoip_annotated_df %>% filter(ER) %>% pull(Gene.Symbol)  # if ER is already a logical column
)


# --- 1. Build annotation membership table over the analysis universe
autophagy_universe <- unique(unlist(annotation_lists, use.names = FALSE))

autophagy_df <- helalysoip_annotated_df %>%
  filter(Gene.Symbol %in% autophagy_universe) %>%
  transmute(
    Gene.Symbol = as.character(Gene.Symbol),
    log2FC_Ctrl  = log2FC_CtrlLysoIP.untagged,
    log2FC_ASAH1 = log2FC_ASAH1LysoIP.untagged
  )

# --- 2. Long table of gene × annotation
autophagy_long <- map_dfr(names(annotation_lists), function(an) {
  tibble(Annotation = an, Gene.Symbol = annotation_lists[[an]])
}) %>%
  distinct() %>%
  inner_join(autophagy_df, by = "Gene.Symbol")

# --- 3. Status per genotype by sign of log2FC vs untagged
autophagy_long <- autophagy_long %>%
  mutate(
    Ctrl_status  = case_when(is.finite(log2FC_Ctrl)  & log2FC_Ctrl  > 0 ~ "up",
                             is.finite(log2FC_Ctrl)  & log2FC_Ctrl  < 0 ~ "down",
                             TRUE ~ "ns"),
    ASAH1_status = case_when(is.finite(log2FC_ASAH1) & log2FC_ASAH1 > 0 ~ "up",
                             is.finite(log2FC_ASAH1) & log2FC_ASAH1 < 0 ~ "down",
                             TRUE ~ "ns")
  )

# --- 4. Prepare data for stacked barplot: counts of up/down per genotype per annotation
stack_counts <- bind_rows(
  autophagy_long %>% filter(Ctrl_status %in% c("up","down"))  %>% 
    transmute(Annotation, Group = paste("Ctrl",  Ctrl_status),  Gene.Symbol),
  autophagy_long %>% filter(ASAH1_status %in% c("up","down")) %>% 
    transmute(Annotation, Group = paste("ASAH1", ASAH1_status), Gene.Symbol)
) %>%
  distinct() %>%
  count(Annotation, Group, name = "n") %>%
  group_by(Annotation) %>%
  mutate(total = sum(n)) %>%
  ungroup()

# wide summary table
stack_counts_wide <- stack_counts %>%
  tidyr::pivot_wider(id_cols = Annotation, names_from = Group, values_from = n, values_fill = 0) %>%
  dplyr::mutate(total = rowSums(across(where(is.numeric))))


# --- 4b. Stacked barplot
ann_order <- stack_counts %>%
  dplyr::group_by(Annotation) %>%
  dplyr::summarise(total = sum(n), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(total)) %>%
  dplyr::pull(Annotation)

stack_plot <- ggplot(stack_counts, aes(y = factor(Annotation, levels = ann_order), x = n, fill = Group)) +
  geom_col(width = 0.8,color = "black", linewidth = 0.1) +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), size = 2.4) +
  scale_fill_manual(values = c("Ctrl up" = "grey50", "Ctrl down" = "grey70", "ASAH1 up" = "dodgerblue3", "ASAH1 down" = "dodgerblue")) +
  labs(x = NULL, y = NULL, title = "Autophagy annotations: up/down per genotype (LysoIP vs untagged)") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), legend.title = element_blank())

ggsave(file.path(out_dir_lysoip_autophagy, "autophagy_updown_stackedbar.pdf"), stack_plot, width = 5, height = 3.5, device = cairo_pdf)


# --- 5. Build gene-level membership table
sig_gene_summary_autophagy <- autophagy_long %>%
  dplyr::select(Gene.Symbol, Annotation, Ctrl_status, ASAH1_status) %>%
  distinct()


# --- 6. Plot intersections via Upsetplot (complexheatmap)
pdf(file.path(out_dir_lysoip_autophagy, "autophagy_upset_CtrlASAH1_updown.pdf"), width = 6, height = 5, useDingbats = FALSE)

ComplexUpset::upset(
  status_wide,
  c("Ctrl_up","Ctrl_down","ASAH1_up","ASAH1_down"),
  base_annotations = list(
    'Intersection size' = intersection_size(text = list(aes(label = after_stat(count))))
  ),
  width_ratio = 0.4
)

dev.off()


# --- 7. GO on ASAH1_up ∩ Ctrl_down overlap IDs (from status_wide)
# 7.1 Set target IDs list and background genes 
asah1up_ctrldown_genes <- rownames(status_wide)[status_wide$ASAH1_up == 1 & status_wide$Ctrl_down == 1]
asah1up_ctrldown_genes <- sort(unique(as.character(asah1up_ctrldown_genes)))
background_genes <- sort(unique(as.character(helalysoip_annotated_df$Gene.Symbol)))

# 7.2 Failsafe export of input list
write.csv(data.frame(Symbol = asah1up_ctrldown_genes), file.path(out_dir_lysoip_autophagy, "GO_input_ASAH1up_Ctrldown.csv"), row.names = FALSE)

# 7.3 Helper (topGO over BP/CC/MF) — saves CSV and dotplot PDFs with labeled points
run_topGO_set <- function(sym_vec, universe_vec, tag, ont = c("BP","CC","MF"), min_genes = 10, out_dir = out_dir_lysoip_autophagy) {
  sym_vec <- unique(sym_vec); universe_vec <- unique(universe_vec)
  if (length(sym_vec) < min_genes) { message("Skip ", tag, ": n=", length(sym_vec), " < ", min_genes); return(invisible(NULL)) }
  geneList <- factor(as.integer(universe_vec %in% sym_vec)); names(geneList) <- universe_vec
  for (o in ont) {
    GOdata <- new("topGOdata", ontology = o, allGenes = geneList, geneSel = function(x) x == 1, annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")
    resF <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
    go_tab <- GenTable(GOdata, classicFisher = resF, topNodes = 30) |>
      dplyr::mutate(raw_pval = score(resF)[GO.ID], log10P = -log10(raw_pval), Significant = as.numeric(Significant), Expected = as.numeric(Expected), Enrichment = Significant / Expected) |>
      dplyr::filter(!is.na(raw_pval)) |>
      dplyr::arrange(raw_pval)
    # genes per term
    gin <- genesInTerm(GOdata)
    go_tab$Genes <- vapply(go_tab$GO.ID, function(g) paste(intersect(gin[[g]], sym_vec), collapse = ";"), character(1))
    # save table + dotplot
    clean_tag <- gsub("[^A-Za-z0-9]+","_", tag)
    write.csv(go_tab, file.path(out_dir, paste0("GO_", o, "_", clean_tag, ".csv")), row.names = FALSE)
    p <- ggplot(go_tab[seq_len(min(10, nrow(go_tab))),], aes(x = Enrichment, y = reorder(Term, Enrichment))) +
      geom_point(aes(size = log10P, fill = Enrichment), shape = 21, color = "black", stroke = 0.25) +
      scale_fill_gradient(low = "white", high = "firebrick") +
      theme_bw(base_size = 8) + theme(panel.grid = element_blank()) +
      labs(x = "Fold Enrichment", y = "GO Term", size = expression(-log[10]*"P"), title = paste0("GO-", o, " (", tag, ")"))
    ggsave(file.path(out_dir, paste0("GO_", o, "_", clean_tag, ".pdf")), p, width = 5, height = 3.5, device = cairo_pdf)
  }
}

# 7.4 Run helper function with custom background (LysoIPset)
run_topGO_set(asah1up_ctrldown_genes, background_genes, tag = "ASAH1_up_AND_Ctrl_down")



# --- 8. save gene-level table
write.csv(sig_gene_summary_autophagy, file = file.path(out_dir_lysoip_autophagy, "sig_hits_annotation_geneList.csv"),row.names = FALSE)
write.csv(status_wide, file.path(out_dir_lysoip_autophagy, "autophagy_updown_membership.csv"), row.names = FALSE)
write_csv(stack_counts, file.path(out_dir_lysoip_autophagy, "sig_hits_autophagy_summary.csv"))
write_csv(stack_counts_wide, file.path(out_dir_lysoip_autophagy, "autophagy_updown_counts_wide.csv"))


```



## VIII. GO term analysis of LysoIP data (based on clusters)
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# GO Term Analysis for LysoIP Clusters (ASAH1 vs Ctrl)
# ─────────────────────────────────────────────────────────────────────────────

# --- Step 1. Define function for GO Term analysis 
plot_topGO_for_geneset <- function(gene_set, background_genes, tag = "clusterX", out_dir) {
  # library(topGO)
  # library(org.Hs.eg.db)
  # library(tidyverse)

  # Step 1: Create geneList for topGO
  geneList <- factor(as.integer(background_genes %in% gene_set))
  names(geneList) <- background_genes

  # Step 2: Run topGO Fisher test (BP)
  GOdata <- new("topGOdata",
                ontology = "BP",
                allGenes = geneList,
                geneSel = function(x) x == 1,
                annot = annFUN.org,
                mapping = "org.Hs.eg.db",
                ID = "symbol")

  resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

  # Step 3: Generate GO table and extract raw p-values
  # calculate 30 top hits, but only plot top 10 
  go_table <- GenTable(GOdata, classicFisher = resultFisher, topNodes = 30)

  go_table <- go_table %>%
    mutate(
      raw_pval     = score(resultFisher)[GO.ID],
      Significant  = as.numeric(Significant),
      Expected     = as.numeric(Expected),
      Enrichment   = Significant / Expected,
      log10P       = -log10(raw_pval)
    ) %>%
    filter(!is.na(raw_pval)) %>%
    arrange(raw_pval) %>%
    slice_head(n = 10)

  # Step 4: Get associated genes per GO term
  go_genes <- genesInTerm(GOdata)
  go_table$Genes <- sapply(go_table$GO.ID, function(go_id) {
    intersect(go_genes[[go_id]], gene_set) %>% paste(collapse = ";")
  })

  # Step 5: Save GO term table + gene associations
  write_csv(go_table, file = file.path(out_dir, paste0("GO_results_", tag, ".csv")))

  # Step 6: Plot dot plot
  p <- ggplot(go_table, aes(x = Enrichment, y = reorder(Term, Enrichment))) +
    geom_point(aes(size = log10P, fill = Enrichment), shape = 21, color = "black", stroke = 0.25) +
    scale_fill_gradient(low = "white", high = "firebrick") +
    theme_bw(base_size = 8) +
    labs(
      x = "Fold Enrichment",
      y = "GO Term",
      size = expression(-log[10]*"P"),
      title = paste0("Top GO Terms (", tag, ")")
    ) +
    theme(panel.grid = element_blank())

  ggsave(file.path(out_dir, paste0("GO_dotplot_", tag, ".pdf")), p, width = 5, height = 3.5, device = cairo_pdf)

  return(p)
}



# --- Step 2: Define target gene sets based on cluster assignment
genes_lysoip_cluster_up <- helalysoip_annotated_df %>%
  filter(cluster %in% c(1)) %>%
  #filter(cluster %in% c(3, 5)) %>%
  pull(Gene.Symbol) %>%
  unique()

# --- Step 3: Define background gene set
background_genes_lysoip <- helalysoip_annotated_df %>%
  pull(Gene.Symbol) %>%
  unique()

# --- Step 4: Run GO Term analysis & plotting
plot_topGO_for_geneset(
  gene_set = genes_lysoip_cluster_up,
  background_genes = background_genes_lysoip,
  tag = "LysoIP_cluster1",
  out_dir = out_dir_lysoip_GO
)

```



# VII. PPI eval on LysoIP data 
## Base: identification of lost / gained PPIs comparing Control and ASAH1 KO LysoIPs
```{r}
# =========================================================================
# HeLa LysoIP PPI Analysis: Control vs ASAH1 KO
# Identify lost and gained PPIs in ASAH1 knockout lysosomes
# =========================================================================

library(tidyverse)
library(pheatmap)
library(RColorBrewer)

# --- Setup ---
out_dir_lysoip_ppi <- file.path(out_dir_lysoip_dataframes, "../PPI")
dir.create(file.path(out_dir_lysoip_ppi, "plots"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(out_dir_lysoip_ppi, "tables"), recursive = TRUE, showWarnings = FALSE)

# Color palettes
state_colors <- c(retained = "grey70", lost = "#f39dcfff", gained = "dodgerblue4")

# =========================================================================
# STEP 1: Define lysosomal proteome baseline (Control LysoIP)
# =========================================================================

message("\n=== Step 1: Defining lysosomal proteome baseline ===")

# Step 1a. Define thresholds
enrichment_thresh <- 0.25   # log2FC threshold for lysosomal proteins
depletion_thresh <- -0.25   # log2FC threshold for background/contaminants

# Step 1b. Lysosomal proteins: enriched in CtrlLysoIP vs untagged
baseline_lysosomal_proteins <- helalysoip_annotated_df %>%
  dplyr::filter(
    log2FC_CtrlLysoIP.untagged > enrichment_thresh,
    is.finite(log2FC_CtrlLysoIP.untagged)
  ) %>%
  dplyr::pull(Gene.Symbol) %>%
  unique()

# Step 1c. Create baseline dataframe were we remove nuclear contaminants. (annotation based)
lysoip_baseline_df_clean <- helalysoip_annotated_df %>%
  dplyr::filter(Gene.Symbol %in% baseline_lysosomal_proteins) %>%
  dplyr::filter(is.na(Nucleus) | Nucleus == FALSE) %>%  # Remove nuclear contaminants
  dplyr::mutate(in_baseline = TRUE)

# Update baseline proteins list
baseline_lysosomal_proteins_clean <- lysoip_baseline_df_clean %>%
  dplyr::pull(Gene.Symbol) %>%
  unique()

message("Original lysosomal proteins (enriched): ", length(baseline_lysosomal_proteins))
message("Clean lysosomal proteins (nuclear removed): ", length(baseline_lysosomal_proteins_clean))
message("Nuclear contaminants removed: ", length(baseline_lysosomal_proteins) - length(baseline_lysosomal_proteins_clean))

# Save results
write_csv(lysoip_baseline_df_clean, file.path(out_dir_lysoip_ppi, "tables", "lysoip_baseline_proteome_clean.csv"))

# =========================================================================
# STEP 2: Map baseline PPIs using baseline
# =========================================================================

message("\n=== Step 2: Mapping baseline PPIs ===")

# Step 2d. Filter to CLEAN lysosomal baseline proteins
lysoip_baseline_edges_clean <- ppi_evidence %>%
  dplyr::filter(
    Gene1 %in% baseline_lysosomal_proteins_clean,
    Gene2 %in% baseline_lysosomal_proteins_clean
  )

message("PPI edges within CLEAN lysosomal baseline: ", nrow(lysoip_baseline_edges_clean))
message("Edges removed by nuclear filtering: ", nrow(lysoip_baseline_edges) - nrow(lysoip_baseline_edges_clean))

# save csv of clean baseline PPIs
write_csv(lysoip_baseline_edges_clean, file.path(out_dir_lysoip_ppi, "tables", "lysoip_baseline_edges_clean.csv"))

# =========================================================================
# STEP 3: Assess PPI state changes using baseline
# =========================================================================

message("\n=== Step 3: Computing PPI state changes with clean baseline ===")

drop_thresh <- -0.25   # Partner lost if log2FC < -0.25
gain_thresh <- 0.25    # Partner gained if log2FC > 0.25

# Step 3a. Match CLEAN edges to protein abundance changes (ASAH1 vs Ctrl)
lysoip_edge_states_clean <- lysoip_baseline_edges_clean %>%
  dplyr::left_join(
    helalysoip_annotated_df %>% dplyr::select(Gene.Symbol, log2FC_ASAH1LysoIP.CtrlLysoIP),
    by = c("Gene1" = "Gene.Symbol")
  ) %>%
  dplyr::rename(dA = log2FC_ASAH1LysoIP.CtrlLysoIP) %>%
  dplyr::left_join(
    helalysoip_annotated_df %>% dplyr::select(Gene.Symbol, log2FC_ASAH1LysoIP.CtrlLysoIP),
    by = c("Gene2" = "Gene.Symbol")
  ) %>%
  dplyr::rename(dB = log2FC_ASAH1LysoIP.CtrlLysoIP) %>%
  dplyr::mutate(
    state = case_when(
      # Edge lost: either partner dropped significantly
      (!is.na(dA) & dA < drop_thresh) | (!is.na(dB) & dB < drop_thresh) ~ "lost",
      # Edge gained: both partners increased significantly
      (!is.na(dA) & dA > gain_thresh) & (!is.na(dB) & dB > gain_thresh) ~ "gained",
      # Edge retained: no significant change
      !is.na(dA) | !is.na(dB) ~ "retained",
      TRUE ~ "unknown"
    )
  )

# Save csv file of PPI edge changes
write_csv(lysoip_edge_states_clean, file.path(out_dir_lysoip_ppi, "tables", "lysoip_edge_states_clean.csv"))

# Step 3b. Summary statistics
edge_summary_clean <- lysoip_edge_states_clean %>%
  dplyr::count(state, name = "n_edges") %>%
  dplyr::mutate(frac = n_edges / sum(n_edges))

message("\nCLEAN PPI edge state summary:")
print(edge_summary_clean)

# Compare with original
message("\nComparison with original (uncleaned) baseline:")
message("Original edges: ", nrow(lysoip_edge_states))
message("Clean edges: ", nrow(lysoip_edge_states_clean))

# save csv file for clean summary stats
write_csv(edge_summary_clean, file.path(out_dir_lysoip_ppi, "tables", "edge_state_summary_clean_ControlToASAH1.csv"))



# =========================================================================
# STEP 4: Visualizations
# =========================================================================

message("\n=== Step 4: Creating visualizations ===")

# Step 4a. Barplot of edge states
p_edge_states <- ggplot(edge_summary, aes(x = state, y = n_edges, fill = state)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.7) +
  geom_text(aes(label = n_edges), vjust = -0.5, size = 3) +
  scale_fill_manual(values = state_colors, name = "Edge State") +
  labs(title = "PPI Edge States: Control → ASAH1 KO", x = NULL, y = "Number of Edges") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), legend.position = "none")

ggsave(file.path(out_dir_lysoip_ppi, "plots", "bars_edge_states.pdf"), p_edge_states, width = 4, height = 4, device = cairo_pdf)


# Step 4b. Scatter plot: Partner A vs Partner B fold changes
scatter_data <- lysoip_edge_states %>%
  dplyr::filter(state %in% c("lost", "gained", "retained")) %>%
  dplyr::mutate(state = factor(state, levels = c("retained", "lost", "gained")))

p_scatter <- ggplot(scatter_data, aes(x = dA, y = dB, color = state, alpha = state)) +
  geom_point(size = 1.5, shape = 16) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black", linewidth = 0.3) +
  scale_color_manual(values = state_colors, name = "Edge State") +
  scale_alpha_manual(values = c(retained = 0.3, lost = 0.8, gained = 0.8), guide = "none") +
  labs(title = "PPI Partner Abundance Changes (ASAH1 vs Ctrl)", x = "log2FC Partner A", y = "log2FC Partner B") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), legend.position = "right")

ggsave(file.path(out_dir_lysoip_ppi, "plots", "scatter_partner_changes.pdf"), p_scatter, width = 5, height = 4, device = cairo_pdf)


# Step 4c. Network visualization (if igraph available)
if (requireNamespace("igraph", quietly = TRUE)) {
  library(igraph)
  
  # Build network from lost edges
  lost_edges <- lysoip_edge_states %>%
    dplyr::filter(state == "lost") %>%
    dplyr::select(Gene1, Gene2)
  
  if (nrow(lost_edges) > 0) {
    net_lost <- igraph::graph_from_data_frame(lost_edges, directed = FALSE)
    
    # Identify hubs (proteins with many lost interactions)
    deg_lost <- igraph::degree(net_lost)
    hubs_lost <- names(deg_lost)[deg_lost >= quantile(deg_lost, 0.9)]
    
    message("\nTop hubs with lost interactions:")
    print(sort(deg_lost[hubs_lost], decreasing = TRUE))
    
    # Save hub list
    hub_df <- tibble(Gene = names(deg_lost), degree = deg_lost) %>%
      dplyr::arrange(desc(degree))
    write_csv(hub_df, file.path(out_dir_lysoip_ppi, "tables", "lost_interaction_hubs.csv"))
  }
  
  # Build network from gained edges
  gained_edges <- lysoip_edge_states %>%
    dplyr::filter(state == "gained") %>%
    dplyr::select(Gene1, Gene2)
  
  if (nrow(gained_edges) > 0) {
    net_gained <- igraph::graph_from_data_frame(gained_edges, directed = FALSE)
    
    deg_gained <- igraph::degree(net_gained)
    hubs_gained <- names(deg_gained)[deg_gained >= quantile(deg_gained, 0.9)]
    
    message("\nTop hubs with gained interactions:")
    print(sort(deg_gained[hubs_gained], decreasing = TRUE))
    
    hub_gained_df <- tibble(Gene = names(deg_gained), degree = deg_gained) %>%
      dplyr::arrange(desc(degree))
    write_csv(hub_gained_df, file.path(out_dir_lysoip_ppi, "tables", "gained_interaction_hubs.csv"))
  }
}


# =========================================================================
# STEP 5: Lysohydrolase-specific analysis
# =========================================================================

message("\n=== Step 5: Lysohydrolase-specific PPI analysis ===")

# Focus on edges involving lysohydrolases
hydrolase_edges <- lysoip_edge_states %>%
  dplyr::filter(Gene1 %in% lysohydrolases | Gene2 %in% lysohydrolases)

hydrolase_summary <- hydrolase_edges %>%
  dplyr::count(state, name = "n_edges")

message("\nLysohydrolase-involved edges:")
print(hydrolase_summary)

write_csv(hydrolase_edges, file.path(out_dir_lysoip_ppi, "tables", "lysohydrolase_edges.csv"))

# Barplot
p_hydrolase <- ggplot(hydrolase_summary, aes(x = state, y = n_edges, fill = state)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.7) +
  geom_text(aes(label = n_edges), vjust = -0.5, size = 3) +
  scale_fill_manual(values = state_colors, name = "Edge State") +
  labs(title = "Lysohydrolase PPI Changes (ASAH1 KO)", x = NULL, y = "Number of Edges") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), legend.position = "none")

ggsave(file.path(out_dir_lysoip_ppi, "plots", "bars_lysohydrolase_edges.pdf"), p_hydrolase, width = 4, height = 4, device = cairo_pdf)



# =========================================================================
# STEP 6: Annotation-level analysis of gained/lost PPIs
# =========================================================================

message("\n=== Step 6: Annotation-level PPI analysis ===")

# Step 6a. Define annotations of interest (must be logical columns in helalysoip_annotated_df)
annotations <- c("Mito", "ER", "Lysosome", "Endo_iN_curated_Hundley","EarlyEndosome","RecyclingEndosome", "Autophagy", "vATPase")

# "LysoHydrolase", "Mito", "MitoIMS", "MitoMatrix","MitoMIM", "MitoMOM", "OXPHOS", "mtComplexI","Golgi", "Lysosome", "ER", "Cytoplasm","Nucleus", "Autophagy", "Autophagy_det", "SynGO","Endo_iN_curated_Hundley", "SynapseSVs", "Presynaptic", "Postsynaptic","EndoDetailLoac", "NeuroDev", "EndoLyso", "EarlyEndosome","RecyclingEndosome", "SVfusion", "Svendocytosis", "Svexocytosis","vATPase" 

# Step 6b. Helper function to count edges per annotation
count_edges_by_annotation <- function(edges_df, annotation_col, annotation_name) {
  edges_df %>%
    filter(
      Gene1 %in% (helalysoip_annotated_df %>% filter(!!sym(annotation_col) == TRUE) %>% pull(Gene.Symbol)) |
      Gene2 %in% (helalysoip_annotated_df %>% filter(!!sym(annotation_col) == TRUE) %>% pull(Gene.Symbol))
    ) %>%
    count(state, name = "n_edges") %>%
    mutate(annotation = annotation_name)
}

# Step 6c. Count edges for each annotation
annotation_edge_counts <- map_dfr(annotations, ~ count_edges_by_annotation(lysoip_edge_states, .x, .x))

# Summarize results
annotation_edge_summary <- annotation_edge_counts %>%
  group_by(annotation, state) %>%
  summarize(n_edges = sum(n_edges), .groups = "drop") %>%
  mutate(state = factor(state, levels = c("lost", "gained", "retained")))

# Save annotation-level summary
write_csv(annotation_edge_summary, file.path(out_dir_lysoip_ppi, "tables", "annotation_edge_summary.csv"))


# Step 6d. Plot annotation-level PPI changes
# Bar plot of gained/lost edges per annotation
p_annotation_edges <- ggplot(annotation_edge_summary, aes(x = annotation, y = n_edges, fill = state)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.3) +
  geom_text(aes(label = n_edges), position = position_dodge(width = 0.9), vjust = -0.5, size = 2.5) +
  scale_fill_manual(values = state_colors, name = "Edge State") +
  labs(title = "Annotation-level PPI Changes (Lyso IP Control -> ASAH1 KO)", x = "Annotation", y = "Number of Edges") +
  theme_bw(base_size = 6) +
  theme(
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black", fill = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # Ensure consistent dimensions with space for n=0
  facet_wrap(~ state, ncol = 1)  # Remove free_y to ensure the same y-axis scale across all facets

# Save the plot
ggsave(file.path(out_dir_lysoip_ppi, "plots", "annotation_edge_changes.pdf"), p_annotation_edges, width = 3, height = 6, device = cairo_pdf)

```


### Trimer analysis of lysosomal PPIs 
```{r}
# =========================================================================
#  Trimer identification and vulnerability analysis 
# =========================================================================

# --- Step 0: Create output directories ---
dir.create(file.path(out_dir_lysoip_ppi, "plots/trimers"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(out_dir_lysoip_ppi, "tables/trimers"), recursive = TRUE, showWarnings = FALSE)

# --- Step 1: Filter to lysosomal high-confidence PPIs only ---
conf_threshold <- 0.80

# Get lysosomal genes
lysosomal_genes_clean <- unique(c(lysoip_edge_states_clean$Gene1, lysoip_edge_states_clean$Gene2))
lysosomal_genes_clean <- lysosomal_genes_clean[!is.na(lysosomal_genes_clean)]

message("Number of CLEAN lysosomal genes: ", length(lysosomal_genes_clean))

# Filter high-confidence edges to lysosomal proteome
high_conf_edges_clean <- lysoip_baseline_edges_clean %>%
  dplyr::filter(
    Predictome_conf >= conf_threshold,
    Gene1 %in% lysosomal_genes_clean,
    Gene2 %in% lysosomal_genes_clean
  ) %>%
  dplyr::select(key, Gene1, Gene2, Predictome_conf)

message("High-confidence CLEAN lysosomal edges (≥", conf_threshold, "): ", nrow(high_conf_edges_clean))

# Reduce threshold if needed
if (nrow(high_conf_edges_clean) < 100) {
  conf_threshold <- 0.75
  high_conf_edges_clean <- lysoip_baseline_edges_clean %>%
    dplyr::filter(
      Predictome_conf >= conf_threshold,
      Gene1 %in% lysosomal_genes_clean,
      Gene2 %in% lysosomal_genes_clean
    ) %>%
    dplyr::select(key, Gene1, Gene2, Predictome_conf)
  message("Reduced threshold to ", conf_threshold, ": ", nrow(high_conf_edges_clean), " edges")
}

if (nrow(high_conf_edges_clean) < 50) {
  conf_threshold <- 0.65
  high_conf_edges_clean <- lysoip_baseline_edges_clean %>%
    dplyr::filter(
      Predictome_conf >= conf_threshold,
      Gene1 %in% lysosomal_genes_clean,
      Gene2 %in% lysosomal_genes_clean
    ) %>%
    dplyr::select(key, Gene1, Gene2, Predictome_conf)
  message("Further reduced threshold to ", conf_threshold, ": ", nrow(high_conf_edges_clean), " edges")
}

# Save high confidence dimers
write_csv(high_conf_edges_clean, file.path(out_dir_lysoip_ppi, "tables", "high_confidence_dimers_clean.csv"))

# --- Step 2: Find all potential trimers (A-B, B-C, A-C all present) ---
# Create adjacency list
edges_df <- high_conf_edges %>%
  dplyr::select(Gene1, Gene2) %>%
  dplyr::distinct()

message("Starting trimer search...")

# Find trimers: for each pair A-B, find all C where B-C and A-C both exist
trimers_list <- list()

for (i in seq_len(nrow(edges_df))) {
  if (i %% 100 == 0) message("  Processed ", i, "/", nrow(edges_df), " edges...")
  
  A <- edges_df$Gene1[i]
  B <- edges_df$Gene2[i]
  
  # Find all C connected to B
  C_via_B <- edges_df %>%
    dplyr::filter((Gene1 == B & Gene2 != A) | (Gene2 == B & Gene1 != A)) %>%
    dplyr::mutate(C = ifelse(Gene1 == B, Gene2, Gene1)) %>%
    dplyr::pull(C) %>%
    unique()
  
  if (length(C_via_B) == 0) next
  
  # Check if A-C edge exists for each C
  for (C in C_via_B) {
    ac_exists <- any(
      (edges_df$Gene1 == A & edges_df$Gene2 == C) |
      (edges_df$Gene1 == C & edges_df$Gene2 == A)
    )
    
    if (ac_exists) {
      # Found a trimer! Sort genes to avoid duplicates
      trimer <- sort(c(A, B, C))
      trimer_id <- paste(trimer, collapse = "::")
      trimers_list[[trimer_id]] <- trimer
    }
  }
}

message("Trimer search complete!")

# Convert to dataframe
trimers_baseline <- tibble::tibble(
  trimer_id = names(trimers_list),
  GeneA = sapply(trimers_list, `[`, 1),
  GeneB = sapply(trimers_list, `[`, 2),
  GeneC = sapply(trimers_list, `[`, 3)
)

message("Total trimers found in lysosomal baseline: ", nrow(trimers_baseline))

# Save baseline trimers
write_csv(trimers_baseline, file.path(out_dir_lysoip_ppi, "tables/trimers", "baseline_trimers.csv"))

# --- Step 3: Add organelle annotations to trimers ---
# Use the same annotation approach from the original code
annotations <- c("Mito", "ER", "Lysosome", "Endo_iN_curated_Hundley","EarlyEndosome","RecyclingEndosome", "Autophagy", "vATPase")

# Create gene-to-annotation mapping from helalysoip_annotated_df
gene_to_org <- lysoip_baseline_df_clean %>%
  dplyr::select(Gene.Symbol, all_of(annotations))

trimers_annotated <- trimers_baseline %>%
  dplyr::left_join(gene_to_org, by = c("GeneA" = "Gene.Symbol")) %>%
  dplyr::rename_with(~paste0(.x, "_A"), all_of(annotations)) %>%
  dplyr::left_join(gene_to_org, by = c("GeneB" = "Gene.Symbol")) %>%
  dplyr::rename_with(~paste0(.x, "_B"), all_of(annotations)) %>%
  dplyr::left_join(gene_to_org, by = c("GeneC" = "Gene.Symbol")) %>%
  dplyr::rename_with(~paste0(.x, "_C"), all_of(annotations))

# Determine shared organelle(s) - all three proteins must be in same organelle
trimers_annotated <- trimers_annotated %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    shared_organelle = {
      org_A <- annotations[which(c_across(paste0(annotations, "_A")))]
      org_B <- annotations[which(c_across(paste0(annotations, "_B")))]
      org_C <- annotations[which(c_across(paste0(annotations, "_C")))]
      
      shared <- Reduce(intersect, list(org_A, org_B, org_C))
      if (length(shared) == 0) "Other" else paste(shared, collapse = ";")
    }
  ) %>%
  dplyr::ungroup()

# Summary by organelle
trimer_org_summary <- trimers_annotated %>%
  tidyr::separate_rows(shared_organelle, sep = ";") %>%
  dplyr::count(shared_organelle, name = "n_trimers") %>%
  dplyr::arrange(desc(n_trimers))

message("Trimers by organelle:")
print(trimer_org_summary)

# save csv
write_csv(trimer_org_summary, file.path(out_dir_lysoip_ppi, "tables/trimers", "baseline_trimers_by_organelle.csv"))

# --- Step 4: Assess trimer integrity for Control vs ASAH1 ---
# Prepare trimer edges
trimers_with_keys <- trimers_annotated %>%
  dplyr::filter(!is.na(GeneA) & !is.na(GeneB) & !is.na(GeneC)) %>%
  dplyr::mutate(
    key_AB = .make_undirected_key(GeneA, GeneB),
    key_BC = .make_undirected_key(GeneB, GeneC),
    key_AC = .make_undirected_key(GeneA, GeneC)
  ) %>%
  dplyr::filter(!is.na(key_AB) & !is.na(key_BC) & !is.na(key_AC))

message("Removed ", nrow(trimers_annotated) - nrow(trimers_with_keys), " trimers with missing edges.")

# Join edge states from lysoip_edge_states
trimers_lyso_states <- trimers_with_keys %>%
  dplyr::left_join(
    lysoip_edge_states %>% dplyr::select(key, state),
    by = c("key_AB" = "key")
  ) %>%
  dplyr::rename(state_AB = state) %>%
  dplyr::left_join(
    lysoip_edge_states %>% dplyr::select(key, state),
    by = c("key_BC" = "key")
  ) %>%
  dplyr::rename(state_BC = state) %>%
  dplyr::left_join(
    lysoip_edge_states %>% dplyr::select(key, state),
    by = c("key_AC" = "key")
  ) %>%
  dplyr::rename(state_AC = state) %>%
  dplyr::mutate(
    state_AB = tidyr::replace_na(state_AB, "retained"),
    state_BC = tidyr::replace_na(state_BC, "retained"),
    state_AC = tidyr::replace_na(state_AC, "retained"),
    n_lost = (state_AB == "lost") + (state_BC == "lost") + (state_AC == "lost"),
    trimer_state = dplyr::case_when(
      n_lost >= 2 ~ "broken",
      n_lost == 1 ~ "partial",
      n_lost == 0 ~ "intact",
      TRUE ~ "unknown"
    )
  )

message("Edge states joined successfully!")

# Save detailed trimer states
write_csv(trimers_lyso_states, file.path(out_dir_lysoip_ppi, "tables/trimers", "trimers_lyso_states.csv"))

# --- Step 5: Summary: broken trimers for Control vs ASAH1 ---
trimer_lyso_summary <- trimers_lyso_states %>%
  dplyr::summarise(
    n_total = n(),
    n_intact = sum(trimer_state == "intact", na.rm = TRUE),
    n_broken = sum(trimer_state == "broken", na.rm = TRUE),
    n_partial = sum(trimer_state == "partial", na.rm = TRUE),
    n_unknown = sum(trimer_state == "unknown", na.rm = TRUE),
    frac_broken = n_broken / n_total,
    frac_partial = n_partial / n_total
  )

# save csv
write_csv(trimer_lyso_summary, file.path(out_dir_lysoip_ppi, "tables/trimers", "trimer_lyso_summary.csv"))

message("Trimer breakage summary:")
print(trimer_lyso_summary)

# --- Step 6: Affected trimers by organelle ---
trimers_affected <- trimers_lyso_states %>%
  dplyr::mutate(
    trimer_affected = ifelse(n_lost >= 1, "affected", "intact")
  )

# Count affected vs intact trimers per organelle
trimer_affected_org <- trimers_affected %>%
  tidyr::separate_rows(shared_organelle, sep = ";") %>%
  dplyr::group_by(shared_organelle, trimer_affected) %>%
  dplyr::summarise(n_trimers = dplyr::n(), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = trimer_affected,
    values_from = n_trimers,
    values_fill = 0
  ) %>%
  dplyr::mutate(
    total = affected + intact,
    frac_affected = affected / total
  ) %>%
  dplyr::arrange(desc(frac_affected))

# save csv
write_csv(trimer_affected_org, file.path(out_dir_lysoip_ppi, "tables/trimers", "trimer_affected_by_organelle.csv"))

message("Affected trimers by organelle:")
print(trimer_affected_org)

# --- Step 7: Visualizations ---
# Barplot of trimer states
trimer_state_counts <- trimers_lyso_states %>%
  dplyr::count(trimer_state)

p_trimer_states <- ggplot(trimer_state_counts, aes(x = trimer_state, y = n, fill = trimer_state)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.7) +
  geom_text(aes(label = n), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c(intact = "grey70", partial = "dodgerblue", broken = "dodgerblue4"), name = "Trimer State") +
  labs(title = "Trimer States: Control → ASAH1 KO", x = NULL, y = "Number of Trimers") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), legend.position = "none")

ggsave(file.path(out_dir_lysoip_ppi, "plots/trimers", "bars_trimer_states.pdf"), p_trimer_states, width = 4, height = 4, device = cairo_pdf)


# --- Step 8: Plot trimer status by annotation ---
# First create overall trimer state counts (like complex_state_counts)
trimer_state_counts <- trimers_lyso_states %>%
  dplyr::count(trimer_state)

p_trimer_states <- ggplot(trimer_state_counts, aes(x = trimer_state, y = n, fill = trimer_state)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.7) +
  geom_text(aes(label = n), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c(intact = "grey70", partial = "dodgerblue", broken = "dodgerblue4"), name = "Trimer State") +
  labs(title = "Trimer States: Control → ASAH1 KO", x = NULL, y = "Number of Trimers") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), legend.position = "none")

ggsave(file.path(out_dir_lysoip_ppi, "plots/trimers", "bars_trimer_states.pdf"), p_trimer_states, width = 4, height = 4, device = cairo_pdf)

# Count trimers by annotation and state
trimer_status_by_annotation <- trimers_lyso_states %>%
  tidyr::separate_rows(shared_organelle, sep = ";") %>%
  dplyr::filter(!is.na(shared_organelle), shared_organelle != "Other") %>%
  dplyr::filter(shared_organelle %in% c("Mito", "ER", "Lysosome", "Endo_iN_curated_Hundley", "EarlyEndosome", "RecyclingEndosome", "Autophagy", "vATPase")) %>%
  dplyr::count(shared_organelle, trimer_state, name = "n_trimers") %>%
  dplyr::mutate(
    shared_organelle = factor(shared_organelle, levels = c("Mito", "ER", "Lysosome", "Endo_iN_curated_Hundley", "EarlyEndosome", "RecyclingEndosome", "Autophagy", "vATPase")),
    trimer_state = factor(trimer_state, levels = c("intact", "partial", "broken"))
  )

# Plot trimer status by annotation
p_trimer_status_annotation <- ggplot(trimer_status_by_annotation, aes(x = shared_organelle, y = n_trimers, fill = trimer_state)) +
  geom_col(position = "stack", color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(intact = "grey70", partial = "dodgerblue", broken = "dodgerblue4"), name = "Trimer State") +
  labs(title = "Trimer Status by Annotation", x = "Annotation", y = "Number of Trimers") +
  theme_bw(base_size = 8) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# Save the plot (matching complex dimensions)
ggsave(file.path(out_dir_lysoip_ppi, "plots/trimers", "trimer_status_by_annotation.pdf"), p_trimer_status_annotation, width = 4, height = 4, device = cairo_pdf)

```


### Multimer Analysis of Lysosmal PPIs 
```{r}
# =========================================================================
#  Identify protein complexes via community detection 
# =========================================================================

# --- Step 0: Create output directories ---
dir.create(file.path(out_dir_lysoip_ppi, "plots/multimer"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(out_dir_lysoip_ppi, "tables/multimer"), recursive = TRUE, showWarnings = FALSE)

# --- Step 1: Build baseline network from high-confidence edges ---
message("Building baseline PPI network...")

# Use same high_conf_edges from trimer analysis
baseline_net <- igraph::graph_from_data_frame(
  d = high_conf_edges %>% dplyr::select(Gene1, Gene2),
  directed = FALSE,
  vertices = data.frame(name = unique(c(high_conf_edges$Gene1, high_conf_edges$Gene2)))
)

# Add edge weights (confidence scores)
igraph::E(baseline_net)$weight <- high_conf_edges$Predictome_conf

message("Baseline network: ", vcount(baseline_net), " proteins, ", ecount(baseline_net), " edges")

# --- Step 2: Community detection using Louvain algorithm ---
message("Running Louvain community detection...")

louvain_communities <- igraph::cluster_louvain(
  baseline_net,
  weights = igraph::E(baseline_net)$weight
)

# Extract community assignments
protein_communities <- data.frame(
  Gene = igraph::V(baseline_net)$name,
  community_id = as.integer(igraph::membership(louvain_communities)),
  stringsAsFactors = FALSE
)

message("Found ", max(protein_communities$community_id), " communities")


# --- Step 3: Filter to meaningful complexes (size 3-10) ---
community_sizes <- protein_communities %>%
  dplyr::count(community_id, name = "size")

valid_communities <- community_sizes %>%
  dplyr::filter(size >= 3, size <= 10) %>%
  dplyr::pull(community_id)

message("Communities with 3-10 proteins: ", length(valid_communities))

complexes_baseline <- protein_communities %>%
  dplyr::filter(community_id %in% valid_communities) %>%
  dplyr::arrange(community_id, Gene)

complexes_summary <- complexes_baseline %>%
  dplyr::group_by(community_id) %>%
  dplyr::summarise(
    complex_id = paste0("Complex_", sprintf("%04d", community_id[1])),
    n_proteins = n(),
    proteins = paste(Gene, collapse = ";"),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(n_proteins))

# save as csv
write_csv(complexes_summary, file.path(out_dir_lysoip_ppi, "tables/multimer", "baseline_complexes.csv"))

message("Baseline complexes summary:")
print(head(complexes_summary, 20))


# --- Step 4: Annotate complexes with organelles ---
complexes_annotated <- complexes_baseline %>%
  dplyr::left_join(gene_to_org, by = c("Gene" = "Gene.Symbol"))

# Determine dominant organelle per complex
complex_organelle <- complexes_annotated %>%
  tidyr::pivot_longer(
    cols = all_of(annotations),
    names_to = "organelle",
    values_to = "present"
  ) %>%
  dplyr::filter(present == TRUE | present == 1) %>%
  dplyr::group_by(community_id, organelle) %>%
  dplyr::summarise(n_proteins = n(), .groups = "drop_last") %>%
  dplyr::slice_max(n_proteins, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(organelle = ifelse(is.na(organelle), "Other", organelle))

complexes_summary <- complexes_summary %>%
  dplyr::left_join(complex_organelle %>% dplyr::select(community_id, organelle), by = "community_id") %>%
  dplyr::mutate(organelle = tidyr::replace_na(organelle, "Other"))

# Count by organelle
complex_org_summary <- complexes_summary %>%
  dplyr::count(organelle, name = "n_complexes") %>%
  dplyr::arrange(desc(n_complexes))

write_csv(complex_org_summary, file.path(out_dir_lysoip_ppi, "tables/multimer", "baseline_complexes_by_organelle.csv"))

message("Complexes by organelle:")
print(complex_org_summary)


# --- Step 5: Assess complex integrity for Control vs ASAH1 ---
# Get all edges within each complex
complex_edges_list <- list()

for (cid in valid_communities) {
  complex_genes <- complexes_baseline %>%
    dplyr::filter(community_id == cid) %>%
    dplyr::pull(Gene)
  
  # Get all edges between proteins in this complex
  complex_edges <- high_conf_edges %>%
    dplyr::filter(Gene1 %in% complex_genes, Gene2 %in% complex_genes) %>%
    dplyr::mutate(community_id = cid)
  
  complex_edges_list[[as.character(cid)]] <- complex_edges
}

complex_edges_df <- dplyr::bind_rows(complex_edges_list)

# Add complex_id
complex_edges_df <- complex_edges_df %>%
  dplyr::left_join(
    complexes_summary %>% dplyr::select(community_id, complex_id),
    by = "community_id"
  )

message("Total edges in complexes: ", nrow(complex_edges_df))


# --- Step 6: Match complex edges to LysoIP states ---
complex_edges_df <- complex_edges_df %>%
  dplyr::mutate(key_gene = .make_undirected_key(Gene1, Gene2))

# Join edge states from lysoip_edge_states
complexes_lyso_states <- complex_edges_df %>%
  dplyr::left_join(
    lysoip_edge_states %>% dplyr::select(key, state),
    by = c("key_gene" = "key")
  ) %>%
  dplyr::left_join(
    complexes_summary %>% dplyr::select(complex_id, organelle),
    by = "complex_id"
  ) %>%
  dplyr::mutate(
    state = tidyr::replace_na(state, "retained"),
    edge_lost = state == "lost"
  )

# Aggregate per complex
complex_lyso_summary <- complexes_lyso_states %>%
  dplyr::group_by(complex_id) %>%
  dplyr::summarise(
    n_edges_total = n(),
    n_edges_lost = sum(edge_lost, na.rm = TRUE),
    frac_edges_lost = n_edges_lost / n_edges_total,
    .groups = "drop"
  ) %>%
  dplyr::left_join(
    complexes_summary %>% dplyr::select(complex_id, n_proteins, organelle),
    by = "complex_id"
  ) %>%
  dplyr::mutate(
    complex_state = case_when(
      frac_edges_lost >= 0.75 ~ "disrupted",   # ≥75% edges lost
      frac_edges_lost >= 0.25 ~ "partial",     # 25-75% lost
      TRUE                    ~ "intact"
    )
  )

# save csv
write_csv(complex_lyso_summary, file.path(out_dir_lysoip_ppi, "tables/multimer", "complexes_lyso_states.csv"))


# --- Step 7: Summary statistics ---
complex_disruption_summary <- complex_lyso_summary %>%
  dplyr::summarise(
    n_complexes = n(),
    n_intact = sum(complex_state == "intact"),
    n_disrupted = sum(complex_state == "disrupted"),
    n_partial = sum(complex_state == "partial"),
    frac_disrupted = n_disrupted / n_complexes
  )

write_csv(complex_disruption_summary, file.path(out_dir_lysoip_ppi, "tables/multimer", "complex_disruption_summary.csv"))

message("Complex disruption summary:")
print(complex_disruption_summary)

# Per organelle
complex_org_disrupted <- complex_lyso_summary %>%
  dplyr::filter(complex_state == "disrupted", organelle != "Other") %>%
  dplyr::count(organelle, name = "n_disrupted") %>%
  dplyr::arrange(desc(n_disrupted))

write_csv(complex_org_disrupted, file.path(out_dir_lysoip_ppi, "tables/multimer", "complex_disrupted_by_organelle.csv"))


# --- Step 8: Visualizations ---
# Barplot: complex states
complex_state_counts <- complex_lyso_summary %>%
  dplyr::count(complex_state)

p_complex_states <- ggplot(complex_state_counts, aes(x = complex_state, y = n, fill = complex_state)) +
  geom_col(color = "black", linewidth = 0.3, width = 0.7) +
  geom_text(aes(label = n), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c(intact = "grey70", partial = "dodgerblue", disrupted = "dodgerblue4"), name = "Complex State") +
  labs(title = "Complex States: Control → ASAH1 KO", x = NULL, y = "Number of Complexes") +
  theme_bw(base_size = 8) +
  theme(panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), legend.position = "none")

ggsave(file.path(out_dir_lysoip_ppi, "plots/multimer", "bars_complex_states.pdf"), p_complex_states, width = 4, height = 4, device = cairo_pdf)

# --- Step 9: Plot complex status by annotation ---
# Count complexes by annotation and state from complex_lyso_summary
complex_status_by_annotation <- complex_lyso_summary %>%
  dplyr::filter(!is.na(organelle), organelle != "Other") %>%
  dplyr::filter(organelle %in% c("Mito", "ER", "Lysosome", "Endo_iN_curated_Hundley", "EarlyEndosome", "RecyclingEndosome", "Autophagy", "vATPase")) %>%
  dplyr::count(organelle, complex_state, name = "n_complexes") %>%
  dplyr::mutate(
    organelle = factor(organelle, levels = c("Mito", "ER", "Lysosome", "Endo_iN_curated_Hundley", "EarlyEndosome", "RecyclingEndosome", "Autophagy", "vATPase")),
    complex_state = factor(complex_state, levels = c("intact", "partial", "disrupted"))
  )

# Plot complex status by annotation
p_complex_status_annotation <- ggplot(complex_status_by_annotation, aes(x = organelle, y = n_complexes, fill = complex_state)) +
  geom_col(position = "stack", color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(intact = "grey70", partial = "dodgerblue", disrupted = "dodgerblue4"), name = "Complex State") +
  labs(title = "Complex Status by Annotation", x = "Annotation", y = "Number of Complexes") +
  theme_bw(base_size = 8) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# Save the plot
ggsave(file.path(out_dir_lysoip_ppi, "plots/multimer", "complex_status_by_annotation.pdf"), p_complex_status_annotation, width = 4, height = 4, device = cairo_pdf)

message("Complex analysis complete!")
```
